### 2019-07-01 submission/manuscript.Rmd
Developing outline

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
  print(x)
  if(is.list(x)){
    x <- unlist(x)
  }
  if(is.numeric(x)){
    if(abs(x - round(x)) < .Machine$double.eps^0.5){
      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
    } else {
      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
    }
  } else {
      paste(x)
  }
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace*{10mm}

# Reintroducing mothur: 10 years later

\vspace{35mm}

Patrick D. Schloss${^1}$${^\dagger}$

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

\vspace{35mm}

## Observation format

\newpage
\linenumbers

## Abstract
75 words

\newpage

## Importance
150 words

\newpage

3000 words total

Goal: describe how to eat an elephant
mothur's origin story...
* Curvy thingy
* dotur/sons
* Sogin email
* ARB but for microbial ecology

Big points: identify the big problems, rate limiting steps, things that were just done wrong
* We may look back and shutter at how things were done, but we need some historical context
* If we solved all of the problems we would never publish or move the field forward
* The most important problem is answering your biological question
* Connect to Lenski's retrospective article

Competition is good and healthy
* A lot of people have been working in this space - e.g. clustering algorithms
* Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))

What would we have done differently?
* Develop mothur as an R package - but tools weren't available, better graphics
* Testing suite from the beginning
* Better job of engaging community

What are we proud of?
* Open source packages
* Development of open instructional materials
* Data driven development of methods
* Database independent - allows people outside of our field to use mothur

Helped create standards
* When something is this popular, you force people to use a certain suite of tools and methods

The future
* 16S rRNA gene sequencing is not dead
* Always improving sequencing methods bring new problems
* Larger datasets, need a replacement for naive open and closed reference clustering
* Lingering controversies - singletons, ASVs

Longterm funding

\newpage

## Acknowledgements

\newpage

## References
25 references

<div id="refs"></div>

\newpage

<!--- \includegraphics{figure_1.png} -->

**Figure 1. Caption caption caption.** Footnotes footnotes footnotes.

### 2019-07-01 submission/manuscript.Rmd
Developing outline

@@
 * The most important problem is answering your biological question
-* Connect to Lenski's retrospective article
 

### 2019-07-02 submission/manuscript.Rmd
Rough in dotur and mothur origin stories

@@
 
-Goal: describe how to eat an elephant
-mothur's origin story...
-* Curvy thingy
-* dotur/sons
-* Sogin email
+Few scientists set out on a nearly two decade long journey with a specific goal in mind; even fewer probably remember the day that the journey started.
+Often we fail to start a scientific journey because it looks too hard.
+Perhaps we get bogged down in all of the things that could go wrong.
+Perhaps we go astray from the path because we find something else that appears more interesting.
+Every scientist picks their own path and takes their own fork in the road.
+It may appear to be a random walk to an outsider.
+Nevertheless, for most established scientists they can tell the circuitous journey and what they have learned along the way.
+
+By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on.
+Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology*, it is worth stepping back and using the development of mothur as a story that likely has parallels to many other research stories that have taken time to develop.
+
+I fondly recall preparing a poster for the 2002 meeting of research groups supported by the NSF-supported Microbial Observatories Program.
+I wanted to triumphantly show that I had sequenced more than 600 16S rRNA gene sequences from a single 0.5-g sample of Alaskan soil.
+This was greater sequencing depth than anyone else had achieved for a single sample.
+As I was preparing the poster, I walked into the office of Jo Handelsman, my postdoctoral research advisor, and laid out the outline for the poster.
+She asked if I could add one of those "curvy things", a rarefaction curve, to show where I was in sampling the community.
+Rarefaction curves and attempts to estimate the taxonomic richness of soil had become popular because of the simple, but impactful mini-review by Jennifer Hughes and her colleagues, which introduced the field to operational taxonomic units (OTUs), rarefaction curves, and richness estimates [DOI: 10.1128/AEM.67.10.4399-4406.2001].
+I do not recall whether that poster had a rarefaction curve on it, but her question has primed much of my career.
+
+***Introducing DOTUR and friends.*** When Jo asked me to generate a rarefaction curve for the poster, the request was not trivial.
+How would I bin the sequences into OTUs?
+Hughes and her colleagues did it manually for datasets that had fewer than 284 sequences.
+Although I could possibly do that for my 600 sequences, my goal was to generate 1,000 sequences from the sample and to repeat that sampling effort for other samples.
+I needed something that could be automated.
+Furthermore, the software that they used, EstimateS, required a series of tedious data formatting steps.
+I had found my first problem.
+How would I assign sequences to OTUs and use that data to estimate the richness and diversity of a sample?
+The second problem would be how could I compare the sequences found in one sample to another sample?
+The solution to the first problem, DOTUR (Distance-based OTUs and Richness), took us two years to develop [10.1128/AEM.71.3.1501-1506.2005].
+DOTUR did two things: given a matrix describing the genetic distance between pairs of sequences, it would cluster those sequences into OTUs for any distance threshold to define the OTUs and then it would use the frequency of each OTU to calculate a variety of alpha diversity metrics.
+The solutions to the second problem would come from our work to develop software including S-LIBSHUFF [2004], SONS (Shared OTUs and Similarity) [2006], and TreeClimber [2006].
+Around the same time, Catherine Lozupone and Rob Knight were developing their UniFrac tools to compare communities with a phylogenetic rather than OTU-based approach [PMID: 16332807; PMID: 17220268].
+With these tools, the field of microbial ecology had a quantitative toolbox for describing and comparing microbial communities.
+
+It is important to remember that we knew there were many problems with 16S rRNA gene sequencing.
+We knew there were biases from extractions and amplification [XXXX].
+We knew there were chimeras [XXXX].
+Getting to the distance matrix required trimming and correcting the sequences, aligning them, and finding the most appropriate way to calculate a pairwise distance.
+It was also rare to have experimental replication to perform statistical tests to compare treatment groups.
+We frequently used a dataset comparing Scottish soils from from Alison McCaig and colleagues.
+This dataset consisted of two experimental groups, each replicated three times with 45 sequences in each replicate.
+Nevertheless, we had excuses and work arounds for these problems that served our needs.
+At the time, we felt that the biggest problems were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
+Along the way we would demonstrate the utility of such tools to answer questions like where are we in the bacterial census?
+How many sequences would it take to see every OTU in that sample of Alaskan soil?
+How does the word usage of *Goodnight, Moon* compare to that of *Portrait of a Lady*?
+More importantly, 1,900 papers used DOTUR to facilitate their own research questions.
+Had we waited to solve all of the problems that plague 16S rRNA gene sequencing, we would still be waiting.
+
+As we developed these tools, I found a unique niche in microbiology.
+I believe that my undergraduate and graduate training as a biological engineer prepared me to think about research questions from a systems perspective, to think quantitatively, and to understand the value of using computer programs to help solve problems.
+As an undergraduate engineering student, I learned the Pascal programming language and promptly forgot much of it as an engineering graduate student.
+As a postdoc, I learned the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [DOI: 10.1128/AEM.70.9.5485-5492.2004].
+After writing my own version of LIBSHUFF and seeing the speed of my collaborator's version written in C++, I converted my Perl version of DOTUR into C++.
+At the time, the conversion from Perl to C++ seemed like an academic exercise.
+My Perl version took a couple of minutes to process the final collection of 1,000 sequences and the C++ version took seconds.
+In hindsight, the decision to convert to C++ was critical.
+The ability to pick up computer languages to solve problems was enabled by my prior training.
+It was also a skill that was virtually unheard of in microbiology.
+Today researchers without the ability to program in Python or R are at a significant disadvantage.
+
+***Introducing mothur.*** Shortly after DOTUR was published, I received an email from Mitch Sogin asking whether DOTUR could handle more than a million sequences.
+Without answering his question, I asked where he found a million sequences.
+Little did I know that his email would represent another pivot in the development of these tools.
+His group would be the first to use 454 sequencing technology to generate 16S rRNA gene sequences [PMID: 16880384].
+Although mothur could assign those sequences to OTUs, it was slow and required a significant amount of RAM.
+As I left my postdoc to start my independent career across the state from Sogin's lab at the University of Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS, S-LIBSHUFF, and TreeClimber for the new world of massively parallelized sequencing.
+The new tool was mothur.
+
 * ARB but for microbial ecology
-
-Big points: identify the big problems, rate limiting steps, things that were just done wrong
-* We may look back and shutter at how things were done, but we need some historical context
-* If we solved all of the problems we would never publish or move the field forward
-* The most important problem is answering your biological question
 
@@
 * Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))
+
+Big points: identify the big problems, rate limiting steps, things that were just done wrong
+* The most important problem is answering your biological question
+* Connect to Lenski's retrospective article
 
@@
 
+Failed experiments
+* Visualization
+* GUI
+* Community-supported tool development
+
 What are we proud of?
 * Open source packages
+* Platform independent
 * Development of open instructional materials
@@
 * Database independent - allows people outside of our field to use mothur
+* Reproducibility
 
@@
 * Larger datasets, need a replacement for naive open and closed reference clustering
-* Lingering controversies - singletons, ASVs
+* Lingering controversies - singletons, ASVs, contamination
 
 Longterm funding
+* The development of mothur was enabled by initial funding from a grant to Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
+* NSF
+* NIH
+* Bootstrapping from other projects
+
+Red Queen
 

### 2019-08-02 submission/manuscript.Rmd
Roughing in more text

@@
 Perhaps we go astray from the path because we find something else that appears more interesting.
-Every scientist picks their own path and takes their own fork in the road.
-It may appear to be a random walk to an outsider.
-Nevertheless, for most established scientists they can tell the circuitous journey and what they have learned along the way.
+Every scientist picks their own path and takes their own forks in the road.
+From the outside, it may appear to be a random walk.
+Nevertheless, these meandering journeys in are common in science.
 
+At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski and Chemistry guy].
 By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on.
@@
 Rarefaction curves and attempts to estimate the taxonomic richness of soil had become popular because of the simple, but impactful mini-review by Jennifer Hughes and her colleagues, which introduced the field to operational taxonomic units (OTUs), rarefaction curves, and richness estimates [DOI: 10.1128/AEM.67.10.4399-4406.2001].
-I do not recall whether that poster had a rarefaction curve on it, but her question has primed much of my career.
+I do not recall whether that poster had a rarefaction curve on it, but her question primed my career.
 
@@
 I needed something that could be automated.
-Furthermore, the software that they used, EstimateS, required a series of tedious data formatting steps.
+Furthermore, the software that Hughes used, EstimateS, required a series of tedious data formatting steps.
 I had found my first problem.
@@
 We knew there were chimeras [XXXX].
-Getting to the distance matrix required trimming and correcting the sequences, aligning them, and finding the most appropriate way to calculate a pairwise distance.
+Getting to the distance matrix required trimming and correcting sequence errors, aligning them, and finding the most appropriate way to calculate a pairwise distance.
 It was also rare to have experimental replication to perform statistical tests to compare treatment groups.
@@
 My Perl version took a couple of minutes to process the final collection of 1,000 sequences and the C++ version took seconds.
-In hindsight, the decision to convert to C++ was critical.
+In hindsight, as we process datasets with millions of sequences, the decision to learn to C++ was critical.
 The ability to pick up computer languages to solve problems was enabled by my prior training.
@@
 
-* ARB but for microbial ecology
+Milling about at a poster session at an ASM General Meeting in New Orleans, I again ran into Mitch Sogin who asked what my plans were for new tools.
+I told him that I wanted to make a tool like ARB (a powerful database tool and phylogenetics package), but for microbial ecology analysis.
+His retort was, "You and what army?" That "army" ended up being Sarah Westcott who has worked on the mothur project largely from its inception.
+Today, mothur is over 200,000 lines of code and Sarah has touched or written nearly every line of code.
+Beyond writing and testing mothur's code base, she has become a conduit for many learning the tools of microbial ecology by patiently answering questions via email and the package's discussion forum.
+The community and I are lucky that Sarah has stayed with the project for more than a decade.
+To be honest, such dependency on a single person makes the project brittle.
+In hindsight, it would have been better to have developed mothur with a team so that there is overlap in people's understanding of how mothur works.
+Although such an arrangement might work in a software engineering firm, it is not practical in an academic setting where funding is limited for developers of free, open source software packages.
+There are certainly projects that make this work, but they are rare.
 
-Competition is good and healthy
+***Challenges of making open source count.*** Anyone can post code to GitHub with a permissive license and claim to be an open source software developer.
+Far more challenging is engaging the target community to make contributions to that code.
+Frankly, we have struggled to expand the number of people that make contributions to the mothur code base.
+One challenge we face is that if we looked to third parties to contribute code to mothur, they would need to know C++.
+Given the paucity of microbiologists with any programming skills, expecting that community to provide contributors that can write code in a syntax that prizes execution efficiency over developer efficiency was not likely.
+In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prices developer efficiency and exists as a series of wrappers to execute other developers' code.
+These choices resulted in many tradeoffs that have impacted ease of installation, usability, execution speed, and flexibility.
+If we were offered a grant to rewrite mothur, we would likely rewrite it as an R package that leaned heavily on the R language's C++ packages.
+Of course, such choices are always best in hindsight and when we started developing mothur, the ability to interface between scripting languages like R and Python and C++ code was not as well developed as it is today, For example, the modern version of the Rcpp package was first released in 2009 and its popularity was not immediate.
+Again, the development of mothur has been a product of the environment that it was created in.
+Although these decisions have largely had positive outcomes, there have been tradeoffs that caused us to sacrifice other goals.
+
+Beyond contributing to the mothur code base, we sought out other ways to include the community as developers.
+The paper describing mothur included N co-authors, all but three (Schloss, Ryabin, and Westcott) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
+Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for other seeking to adapt mothur for their uses.
+Again, this vision was a product of the environment.
+IPython notebooks (2011) and R markdown (2012) would not be developed until later.
+Unfortunately, once the incentive of co-authorship was removed, researchers stopped contributing their workflows to the wiki.
+Part of the difficulty of recruiting wiki contributors was a perception by some that the wiki was not a community resource.
+For example, I would frequently receive emails from people telling me that there was a typo on a specific page when the intention was that they could correct the typos without my input.
+We have been more successful in soliciting input and contributions from the user community through the mothur discussion forum and GitHub issue tracker.
+As mothur has matured, we have been dependent on the user community to use these resources to tell us what features they would like to see included in mothur.
+The user community also tells us where our documentation is confusing.
+Often we can count on people not directly affiliated with mothur to provide instruction and their own experience to other users.
+We are constantly trying to recruit our "army" and are happy to take any contributions we can.
+Whether the contributions are to the code base, discussion forum, or suggestions for new tools, these contributions have been invaluable to the growth and popularity of mothur.
+
+***Failed experiments.*** If we never failed, we would not be trying hard enough.
+Over the past decade we have tried a number of experiments to improve the usability and utility of mothur.
+One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and venn diagrams depicting the overlap between microbial communities.
+I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote.
+Such visuals require far too much customization to be publication-quality.
+Although QIIME has incorporated visualization tools through the Emperor package, the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature.
+Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials.
+A second experiment we attempted was to create a graphical user interface (GUI) for running mothur.
+Forcing users to interact with mothur through the command line has been a significant hurdle for many.
+Unfortunately, the development effort required to create and maintain a GUI is significant and there is limited funding for such efforts.
+The newest version of QIIME (version 2) has emphasized interaction with the tools through a GUI and it remains to be seen how this experiment will go.
+Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks.
+This documentation is explicit in mothur as all commands and output is recorded in a logfile.
+Given the heightened recent focus on reproducibility we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project.
+Finally, we collaborated with programmers through Google Summer of Code to develop commands in mothur to implement the random forest and SVM machine learning algorithms.
+Similar to the challenges of developing attractive visuals, fitting the algorithms' hyperparameters, testing, and deploying the resulting models requires a significant amount of customization.
+Furthermore, this is an active area of research where methods are still being developed and improved.
+Thankfully, there are numerous R and Python packages that do a better job of developing these models.
+Again, we have put our efforts into developing materials that mothur users can use to fit such models to their data.
+In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.
+In hindsight, our solution to these failure has been to provide tutorials to as a conduit between mothur and their goals.
+
+***Competition is good and healthy.***
+
 * A lot of people have been working in this space - e.g. clustering algorithms
 * Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))
-
-Big points: identify the big problems, rate limiting steps, things that were just done wrong
-* The most important problem is answering your biological question
-* Connect to Lenski's retrospective article
-
-What would we have done differently?
-* Develop mothur as an R package - but tools weren't available, better graphics
-* Testing suite from the beginning
-* Better job of engaging community
-
-Failed experiments
-* Visualization
-* GUI
-* Community-supported tool development
 
@@
 * Reproducibility
-
-Helped create standards
-* When something is this popular, you force people to use a certain suite of tools and methods
+* Helped create standards - when something is this popular, you force people to use a certain suite of tools and methods
 

### 2019-09-16 submission/manuscript.Rmd
Editing and figuring out where we are

@@
 
-Few scientists set out on a nearly two decade long journey with a specific goal in mind; even fewer probably remember the day that the journey started.
+Few scientists set out on a nearly two decade long journey with a specific goal in mind.
 Often we fail to start a scientific journey because it looks too hard.
@@
 Nevertheless, we had excuses and work arounds for these problems that served our needs.
-At the time, we felt that the biggest problems were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
+At the time, I felt that the biggest problems were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
 Along the way we would demonstrate the utility of such tools to answer questions like where are we in the bacterial census?
@@
 As a postdoc, I learned the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [DOI: 10.1128/AEM.70.9.5485-5492.2004].
-After writing my own version of LIBSHUFF and seeing the speed of my collaborator's version written in C++, I converted my Perl version of DOTUR into C++.
-At the time, the conversion from Perl to C++ seemed like an academic exercise.
-My Perl version took a couple of minutes to process the final collection of 1,000 sequences and the C++ version took seconds.
-In hindsight, as we process datasets with millions of sequences, the decision to learn to C++ was critical.
+After writing my own version of LIBSHUFF and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
+At the time, the conversion from Perl to C++ seemed like an academic exercise to learn a new language.
+My Perl version only took a minute or so to process the final collection of 1,000 sequences and the C++ version took seconds.
+Was that really such a big difference?
+In hindsight, as we now process datasets with millions of sequences, the decision to learn to C++ was critical.
 The ability to pick up computer languages to solve problems was enabled by my prior training.
@@
 
-Milling about at a poster session at an ASM General Meeting in New Orleans, I again ran into Mitch Sogin who asked what my plans were for new tools.
+Milling about at a poster session at an ASM General Meeting in New Orleans, I again ran into Mitch who asked what my plans were for new tools.
 I told him that I wanted to make a tool like ARB (a powerful database tool and phylogenetics package), but for microbial ecology analysis.
-His retort was, "You and what army?" That "army" ended up being Sarah Westcott who has worked on the mothur project largely from its inception.
+His retort was, "You and what army?" To that point, I had written every line of code and been answering many emails from people asking for help.
+It would be difficulty, but I needed to learn to let go and share the development process with someone else.
+He was right, I would need an army.
+That "army" ended up being Sarah Westcott who has worked on the mothur project largely from its inception.
 Today, mothur is over 200,000 lines of code and Sarah has touched or written nearly every line of code.
@@
 To be honest, such dependency on a single person makes the project brittle.
-In hindsight, it would have been better to have developed mothur with a team so that there is overlap in people's understanding of how mothur works.
+In hindsight, it would have been better to have developed mothur with more of an "army" or team so that there is overlap in people's understanding of how mothur works.
 Although such an arrangement might work in a software engineering firm, it is not practical in an academic setting where funding is limited for developers of free, open source software packages.
@@
 Given the paucity of microbiologists with any programming skills, expecting that community to provide contributors that can write code in a syntax that prizes execution efficiency over developer efficiency was not likely.
-In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prices developer efficiency and exists as a series of wrappers to execute other developers' code.
+In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prizes developer efficiency over execution efficiency and exists as a series of wrappers to execute other developers' code.
 These choices resulted in many tradeoffs that have impacted ease of installation, usability, execution speed, and flexibility.
@@
 Beyond contributing to the mothur code base, we sought out other ways to include the community as developers.
-The paper describing mothur included N co-authors, all but three (Schloss, Ryabin, and Westcott) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
+The paper describing mothur included **N** co-authors, all but three (Schloss, Ryabin, and Westcott) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
 Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for other seeking to adapt mothur for their uses.
@@
 Thankfully, there are numerous R and Python packages that do a better job of developing these models.
-Again, we have put our efforts into developing materials that mothur users can use to fit such models to their data.
+Again, we have put our efforts into developing instructional materials that mothur users can use to fit such models to their data.
 In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.

### 2019-09-16 submission/manuscript.tex
Initial rendering

--- 2019-09-16 submission/manuscript.Rmd
@@
----
-bibliography: references.bib
-output:
-  pdf_document:
-    keep_tex: true
-    includes:
-      in_header: header.tex
-csl: mbio.csl
-fontsize: 11pt
-geometry: margin=1.0in
----
+\documentclass[11pt,]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1.0in]{geometry}
+\usepackage{hyperref}
+\hypersetup{unicode=true,
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
-```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
-options(tidyverse.quiet = TRUE)
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
 
-library(tidyverse)
-library(knitr)
+%%% Change title format to be more compact
+\usepackage{titling}
 
-opts_chunk$set("tidy" = TRUE)
-opts_chunk$set("echo" = FALSE)
-opts_chunk$set("eval" = TRUE)
-opts_chunk$set("warning" = FALSE)
-opts_chunk$set("message" = FALSE)
-opts_chunk$set("cache" = FALSE)
-
-inline_hook <- function(x){
-  print(x)
-  if(is.list(x)){
-    x <- unlist(x)
-  }
-  if(is.numeric(x)){
-    if(abs(x - round(x)) < .Machine$double.eps^0.5){
-      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-    } else {
-      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
-    }
-  } else {
-      paste(x)
+% Create subtitle command for use in maketitle
+\providecommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
     }
 }
-knitr::knit_hooks$set(inline=inline_hook)
-```
+
+\setlength{\droptitle}{-2em}
+
+  \title{}
+    \pretitle{\vspace{\droptitle}}
+  \posttitle{}
+    \author{}
+    \preauthor{}\postauthor{}
+    \date{}
+    \predate{}\postdate{}
+
+\usepackage{helvet} % Helvetica font
+\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+\usepackage[T1]{fontenc}
+
+\usepackage[none]{hyphenat}
+
+\usepackage{setspace}
+\doublespacing
+\setlength{\parskip}{1em}
+
+\usepackage{lineno}
+
+\usepackage{pdfpages}
+
+\begin{document}
 
@@
 
-# Reintroducing mothur: 10 years later
+\hypertarget{reintroducing-mothur-10-years-later}{%
+\section{Reintroducing mothur: 10 years
+later}\label{reintroducing-mothur-10-years-later}}
 
@@
 
-Patrick D. Schloss${^1}$${^\dagger}$
+Patrick D. Schloss\({^1}\)\({^\dagger}\)
 
@@
 
-$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
+\(\dagger\) To whom correspondence should be addressed:
+\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
 
-$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
+\(1\) Department of Microbiology and Immunology, University of Michigan,
+Ann Arbor, MI 48109
 
@@
 
-## Observation format
+\hypertarget{observation-format}{%
+\subsection{Observation format}\label{observation-format}}
 
@@
 
-## Abstract
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
+
 75 words
@@
 
-## Importance
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
+
 150 words
@@
 
-Few scientists set out on a nearly two decade long journey with a specific goal in mind.
-Often we fail to start a scientific journey because it looks too hard.
-Perhaps we get bogged down in all of the things that could go wrong.
-Perhaps we go astray from the path because we find something else that appears more interesting.
-Every scientist picks their own path and takes their own forks in the road.
-From the outside, it may appear to be a random walk.
-Nevertheless, these meandering journeys in are common in science.
+Few scientists set out on a nearly two decade long journey with a
+specific goal in mind. Often we fail to start a scientific journey
+because it looks too hard. Perhaps we get bogged down in all of the
+things that could go wrong. Perhaps we go astray from the path because
+we find something else that appears more interesting. Every scientist
+picks their own path and takes their own forks in the road. From the
+outside, it may appear to be a random walk. Nevertheless, these
+meandering journeys in are common in science.
 
-At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski and Chemistry guy].
-By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on.
-Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology*, it is worth stepping back and using the development of mothur as a story that likely has parallels to many other research stories that have taken time to develop.
+At the risk of navel gazing, looking back on our scientific journeys can
+be instructive to other scientists who are overwhelmed at the prospect
+of looking forward at their careers {[}Lenski and Chemistry guy{]}. By
+no means is my personal journey over, but since 2002 I have been on a
+journey that I did not realize I was on. Now that the paper introducing
+the mothur software package is ten years old and has become the most
+cited paper published by \emph{Applied and Environmental Microbiology},
+it is worth stepping back and using the development of mothur as a story
+that likely has parallels to many other research stories that have taken
+time to develop.
 
-I fondly recall preparing a poster for the 2002 meeting of research groups supported by the NSF-supported Microbial Observatories Program.
-I wanted to triumphantly show that I had sequenced more than 600 16S rRNA gene sequences from a single 0.5-g sample of Alaskan soil.
-This was greater sequencing depth than anyone else had achieved for a single sample.
-As I was preparing the poster, I walked into the office of Jo Handelsman, my postdoctoral research advisor, and laid out the outline for the poster.
-She asked if I could add one of those "curvy things", a rarefaction curve, to show where I was in sampling the community.
-Rarefaction curves and attempts to estimate the taxonomic richness of soil had become popular because of the simple, but impactful mini-review by Jennifer Hughes and her colleagues, which introduced the field to operational taxonomic units (OTUs), rarefaction curves, and richness estimates [DOI: 10.1128/AEM.67.10.4399-4406.2001].
-I do not recall whether that poster had a rarefaction curve on it, but her question primed my career.
+I fondly recall preparing a poster for the 2002 meeting of research
+groups supported by the NSF-supported Microbial Observatories Program. I
+wanted to triumphantly show that I had sequenced more than 600 16S rRNA
+gene sequences from a single 0.5-g sample of Alaskan soil. This was
+greater sequencing depth than anyone else had achieved for a single
+sample. As I was preparing the poster, I walked into the office of Jo
+Handelsman, my postdoctoral research advisor, and laid out the outline
+for the poster. She asked if I could add one of those ``curvy things'',
+a rarefaction curve, to show where I was in sampling the community.
+Rarefaction curves and attempts to estimate the taxonomic richness of
+soil had become popular because of the simple, but impactful mini-review
+by Jennifer Hughes and her colleagues, which introduced the field to
+operational taxonomic units (OTUs), rarefaction curves, and richness
+estimates {[}DOI: 10.1128/AEM.67.10.4399-4406.2001{]}. I do not recall
+whether that poster had a rarefaction curve on it, but her question
+primed my career.
 
-***Introducing DOTUR and friends.*** When Jo asked me to generate a rarefaction curve for the poster, the request was not trivial.
-How would I bin the sequences into OTUs?
-Hughes and her colleagues did it manually for datasets that had fewer than 284 sequences.
-Although I could possibly do that for my 600 sequences, my goal was to generate 1,000 sequences from the sample and to repeat that sampling effort for other samples.
-I needed something that could be automated.
-Furthermore, the software that Hughes used, EstimateS, required a series of tedious data formatting steps.
-I had found my first problem.
-How would I assign sequences to OTUs and use that data to estimate the richness and diversity of a sample?
-The second problem would be how could I compare the sequences found in one sample to another sample?
-The solution to the first problem, DOTUR (Distance-based OTUs and Richness), took us two years to develop [10.1128/AEM.71.3.1501-1506.2005].
-DOTUR did two things: given a matrix describing the genetic distance between pairs of sequences, it would cluster those sequences into OTUs for any distance threshold to define the OTUs and then it would use the frequency of each OTU to calculate a variety of alpha diversity metrics.
-The solutions to the second problem would come from our work to develop software including S-LIBSHUFF [2004], SONS (Shared OTUs and Similarity) [2006], and TreeClimber [2006].
-Around the same time, Catherine Lozupone and Rob Knight were developing their UniFrac tools to compare communities with a phylogenetic rather than OTU-based approach [PMID: 16332807; PMID: 17220268].
-With these tools, the field of microbial ecology had a quantitative toolbox for describing and comparing microbial communities.
+\textbf{\emph{Introducing DOTUR and friends.}} When Jo asked me to
+generate a rarefaction curve for the poster, the request was not
+trivial. How would I bin the sequences into OTUs? Hughes and her
+colleagues did it manually for datasets that had fewer than 284
+sequences. Although I could possibly do that for my 600 sequences, my
+goal was to generate 1,000 sequences from the sample and to repeat that
+sampling effort for other samples. I needed something that could be
+automated. Furthermore, the software that Hughes used, EstimateS,
+required a series of tedious data formatting steps. I had found my first
+problem. How would I assign sequences to OTUs and use that data to
+estimate the richness and diversity of a sample? The second problem
+would be how could I compare the sequences found in one sample to
+another sample? The solution to the first problem, DOTUR (Distance-based
+OTUs and Richness), took us two years to develop
+{[}10.1128/AEM.71.3.1501-1506.2005{]}. DOTUR did two things: given a
+matrix describing the genetic distance between pairs of sequences, it
+would cluster those sequences into OTUs for any distance threshold to
+define the OTUs and then it would use the frequency of each OTU to
+calculate a variety of alpha diversity metrics. The solutions to the
+second problem would come from our work to develop software including
+S-LIBSHUFF {[}2004{]}, SONS (Shared OTUs and Similarity) {[}2006{]}, and
+TreeClimber {[}2006{]}. Around the same time, Catherine Lozupone and Rob
+Knight were developing their UniFrac tools to compare communities with a
+phylogenetic rather than OTU-based approach {[}PMID: 16332807; PMID:
+17220268{]}. With these tools, the field of microbial ecology had a
+quantitative toolbox for describing and comparing microbial communities.
 
-It is important to remember that we knew there were many problems with 16S rRNA gene sequencing.
-We knew there were biases from extractions and amplification [XXXX].
-We knew there were chimeras [XXXX].
-Getting to the distance matrix required trimming and correcting sequence errors, aligning them, and finding the most appropriate way to calculate a pairwise distance.
-It was also rare to have experimental replication to perform statistical tests to compare treatment groups.
-We frequently used a dataset comparing Scottish soils from from Alison McCaig and colleagues.
-This dataset consisted of two experimental groups, each replicated three times with 45 sequences in each replicate.
-Nevertheless, we had excuses and work arounds for these problems that served our needs.
-At the time, I felt that the biggest problems were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
-Along the way we would demonstrate the utility of such tools to answer questions like where are we in the bacterial census?
-How many sequences would it take to see every OTU in that sample of Alaskan soil?
-How does the word usage of *Goodnight, Moon* compare to that of *Portrait of a Lady*?
-More importantly, 1,900 papers used DOTUR to facilitate their own research questions.
-Had we waited to solve all of the problems that plague 16S rRNA gene sequencing, we would still be waiting.
+It is important to remember that we knew there were many problems with
+16S rRNA gene sequencing. We knew there were biases from extractions and
+amplification {[}XXXX{]}. We knew there were chimeras {[}XXXX{]}.
+Getting to the distance matrix required trimming and correcting sequence
+errors, aligning them, and finding the most appropriate way to calculate
+a pairwise distance. It was also rare to have experimental replication
+to perform statistical tests to compare treatment groups. We frequently
+used a dataset comparing Scottish soils from from Alison McCaig and
+colleagues. This dataset consisted of two experimental groups, each
+replicated three times with 45 sequences in each replicate.
+Nevertheless, we had excuses and work arounds for these problems that
+served our needs. At the time, I felt that the biggest problems were how
+to cluster the sequences into OTUs and how to use those clusterings to
+test our hypotheses. Along the way we would demonstrate the utility of
+such tools to answer questions like where are we in the bacterial
+census? How many sequences would it take to see every OTU in that sample
+of Alaskan soil? How does the word usage of \emph{Goodnight, Moon}
+compare to that of \emph{Portrait of a Lady}? More importantly, 1,900
+papers used DOTUR to facilitate their own research questions. Had we
+waited to solve all of the problems that plague 16S rRNA gene
+sequencing, we would still be waiting.
 
-As we developed these tools, I found a unique niche in microbiology.
-I believe that my undergraduate and graduate training as a biological engineer prepared me to think about research questions from a systems perspective, to think quantitatively, and to understand the value of using computer programs to help solve problems.
-As an undergraduate engineering student, I learned the Pascal programming language and promptly forgot much of it as an engineering graduate student.
-As a postdoc, I learned the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [DOI: 10.1128/AEM.70.9.5485-5492.2004].
-After writing my own version of LIBSHUFF and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
-At the time, the conversion from Perl to C++ seemed like an academic exercise to learn a new language.
-My Perl version only took a minute or so to process the final collection of 1,000 sequences and the C++ version took seconds.
-Was that really such a big difference?
-In hindsight, as we now process datasets with millions of sequences, the decision to learn to C++ was critical.
-The ability to pick up computer languages to solve problems was enabled by my prior training.
-It was also a skill that was virtually unheard of in microbiology.
-Today researchers without the ability to program in Python or R are at a significant disadvantage.
+As we developed these tools, I found a unique niche in microbiology. I
+believe that my undergraduate and graduate training as a biological
+engineer prepared me to think about research questions from a systems
+perspective, to think quantitatively, and to understand the value of
+using computer programs to help solve problems. As an undergraduate
+engineering student, I learned the Pascal programming language and
+promptly forgot much of it as an engineering graduate student. As a
+postdoc, I learned the Perl programming language to better understand
+how LIBSHUFF, a tool for comparing the structure of two communities,
+worked since it was written in Perl {[}DOI:
+10.1128/AEM.70.9.5485-5492.2004{]}. After writing my own version of
+LIBSHUFF and seeing the speed of the version written in C++ by my
+collaborator, Bret Larget, I converted my Perl version of DOTUR into
+C++. At the time, the conversion from Perl to C++ seemed like an
+academic exercise to learn a new language. My Perl version only took a
+minute or so to process the final collection of 1,000 sequences and the
+C++ version took seconds. Was that really such a big difference? In
+hindsight, as we now process datasets with millions of sequences, the
+decision to learn to C++ was critical. The ability to pick up computer
+languages to solve problems was enabled by my prior training. It was
+also a skill that was virtually unheard of in microbiology. Today
+researchers without the ability to program in Python or R are at a
+significant disadvantage.
 
-***Introducing mothur.*** Shortly after DOTUR was published, I received an email from Mitch Sogin asking whether DOTUR could handle more than a million sequences.
-Without answering his question, I asked where he found a million sequences.
-Little did I know that his email would represent another pivot in the development of these tools.
-His group would be the first to use 454 sequencing technology to generate 16S rRNA gene sequences [PMID: 16880384].
-Although mothur could assign those sequences to OTUs, it was slow and required a significant amount of RAM.
-As I left my postdoc to start my independent career across the state from Sogin's lab at the University of Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS, S-LIBSHUFF, and TreeClimber for the new world of massively parallelized sequencing.
-The new tool was mothur.
+\textbf{\emph{Introducing mothur.}} Shortly after DOTUR was published, I
+received an email from Mitch Sogin asking whether DOTUR could handle
+more than a million sequences. Without answering his question, I asked
+where he found a million sequences. Little did I know that his email
+would represent another pivot in the development of these tools. His
+group would be the first to use 454 sequencing technology to generate
+16S rRNA gene sequences {[}PMID: 16880384{]}. Although mothur could
+assign those sequences to OTUs, it was slow and required a significant
+amount of RAM. As I left my postdoc to start my independent career
+across the state from Sogin's lab at the University of Massachusetts in
+Amherst, my plan was to rewrite DOTUR, SONS, S-LIBSHUFF, and TreeClimber
+for the new world of massively parallelized sequencing. The new tool was
+mothur.
 
-Milling about at a poster session at an ASM General Meeting in New Orleans, I again ran into Mitch who asked what my plans were for new tools.
-I told him that I wanted to make a tool like ARB (a powerful database tool and phylogenetics package), but for microbial ecology analysis.
-His retort was, "You and what army?" To that point, I had written every line of code and been answering many emails from people asking for help.
-It would be difficulty, but I needed to learn to let go and share the development process with someone else.
-He was right, I would need an army.
-That "army" ended up being Sarah Westcott who has worked on the mothur project largely from its inception.
-Today, mothur is over 200,000 lines of code and Sarah has touched or written nearly every line of code.
-Beyond writing and testing mothur's code base, she has become a conduit for many learning the tools of microbial ecology by patiently answering questions via email and the package's discussion forum.
-The community and I are lucky that Sarah has stayed with the project for more than a decade.
-To be honest, such dependency on a single person makes the project brittle.
-In hindsight, it would have been better to have developed mothur with more of an "army" or team so that there is overlap in people's understanding of how mothur works.
-Although such an arrangement might work in a software engineering firm, it is not practical in an academic setting where funding is limited for developers of free, open source software packages.
-There are certainly projects that make this work, but they are rare.
+Milling about at a poster session at an ASM General Meeting in New
+Orleans, I again ran into Mitch who asked what my plans were for new
+tools. I told him that I wanted to make a tool like ARB (a powerful
+database tool and phylogenetics package), but for microbial ecology
+analysis. His retort was, ``You and what army?'' To that point, I had
+written every line of code and been answering many emails from people
+asking for help. It would be difficulty, but I needed to learn to let go
+and share the development process with someone else. He was right, I
+would need an army. That ``army'' ended up being Sarah Westcott who has
+worked on the mothur project largely from its inception. Today, mothur
+is over 200,000 lines of code and Sarah has touched or written nearly
+every line of code. Beyond writing and testing mothur's code base, she
+has become a conduit for many learning the tools of microbial ecology by
+patiently answering questions via email and the package's discussion
+forum. The community and I are lucky that Sarah has stayed with the
+project for more than a decade. To be honest, such dependency on a
+single person makes the project brittle. In hindsight, it would have
+been better to have developed mothur with more of an ``army'' or team so
+that there is overlap in people's understanding of how mothur works.
+Although such an arrangement might work in a software engineering firm,
+it is not practical in an academic setting where funding is limited for
+developers of free, open source software packages. There are certainly
+projects that make this work, but they are rare.
 
-***Challenges of making open source count.*** Anyone can post code to GitHub with a permissive license and claim to be an open source software developer.
-Far more challenging is engaging the target community to make contributions to that code.
-Frankly, we have struggled to expand the number of people that make contributions to the mothur code base.
-One challenge we face is that if we looked to third parties to contribute code to mothur, they would need to know C++.
-Given the paucity of microbiologists with any programming skills, expecting that community to provide contributors that can write code in a syntax that prizes execution efficiency over developer efficiency was not likely.
-In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prizes developer efficiency over execution efficiency and exists as a series of wrappers to execute other developers' code.
-These choices resulted in many tradeoffs that have impacted ease of installation, usability, execution speed, and flexibility.
-If we were offered a grant to rewrite mothur, we would likely rewrite it as an R package that leaned heavily on the R language's C++ packages.
-Of course, such choices are always best in hindsight and when we started developing mothur, the ability to interface between scripting languages like R and Python and C++ code was not as well developed as it is today, For example, the modern version of the Rcpp package was first released in 2009 and its popularity was not immediate.
-Again, the development of mothur has been a product of the environment that it was created in.
-Although these decisions have largely had positive outcomes, there have been tradeoffs that caused us to sacrifice other goals.
+\textbf{\emph{Challenges of making open source count.}} Anyone can post
+code to GitHub with a permissive license and claim to be an open source
+software developer. Far more challenging is engaging the target
+community to make contributions to that code. Frankly, we have struggled
+to expand the number of people that make contributions to the mothur
+code base. One challenge we face is that if we looked to third parties
+to contribute code to mothur, they would need to know C++. Given the
+paucity of microbiologists with any programming skills, expecting that
+community to provide contributors that can write code in a syntax that
+prizes execution efficiency over developer efficiency was not likely. In
+contrast, the QIIME development team could be more distributed because
+their code base was primarily written in Python, which prizes developer
+efficiency over execution efficiency and exists as a series of wrappers
+to execute other developers' code. These choices resulted in many
+tradeoffs that have impacted ease of installation, usability, execution
+speed, and flexibility. If we were offered a grant to rewrite mothur, we
+would likely rewrite it as an R package that leaned heavily on the R
+language's C++ packages. Of course, such choices are always best in
+hindsight and when we started developing mothur, the ability to
+interface between scripting languages like R and Python and C++ code was
+not as well developed as it is today, For example, the modern version of
+the Rcpp package was first released in 2009 and its popularity was not
+immediate. Again, the development of mothur has been a product of the
+environment that it was created in. Although these decisions have
+largely had positive outcomes, there have been tradeoffs that caused us
+to sacrifice other goals.
 
-Beyond contributing to the mothur code base, we sought out other ways to include the community as developers.
-The paper describing mothur included **N** co-authors, all but three (Schloss, Ryabin, and Westcott) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
-Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for other seeking to adapt mothur for their uses.
-Again, this vision was a product of the environment.
-IPython notebooks (2011) and R markdown (2012) would not be developed until later.
-Unfortunately, once the incentive of co-authorship was removed, researchers stopped contributing their workflows to the wiki.
-Part of the difficulty of recruiting wiki contributors was a perception by some that the wiki was not a community resource.
-For example, I would frequently receive emails from people telling me that there was a typo on a specific page when the intention was that they could correct the typos without my input.
-We have been more successful in soliciting input and contributions from the user community through the mothur discussion forum and GitHub issue tracker.
-As mothur has matured, we have been dependent on the user community to use these resources to tell us what features they would like to see included in mothur.
-The user community also tells us where our documentation is confusing.
-Often we can count on people not directly affiliated with mothur to provide instruction and their own experience to other users.
-We are constantly trying to recruit our "army" and are happy to take any contributions we can.
-Whether the contributions are to the code base, discussion forum, or suggestions for new tools, these contributions have been invaluable to the growth and popularity of mothur.
+Beyond contributing to the mothur code base, we sought out other ways to
+include the community as developers. The paper describing mothur
+included \textbf{N} co-authors, all but three (Schloss, Ryabin, and
+Westcott) responded to a call to provide a wiki page that described how
+they used an early version of mothur to analyze a data set. Our vision
+was that authors might use the mothur wiki to document reproducible
+workflows for papers using mothur but to also provide instructional
+materials for other seeking to adapt mothur for their uses. Again, this
+vision was a product of the environment. IPython notebooks (2011) and R
+markdown (2012) would not be developed until later. Unfortunately, once
+the incentive of co-authorship was removed, researchers stopped
+contributing their workflows to the wiki. Part of the difficulty of
+recruiting wiki contributors was a perception by some that the wiki was
+not a community resource. For example, I would frequently receive emails
+from people telling me that there was a typo on a specific page when the
+intention was that they could correct the typos without my input. We
+have been more successful in soliciting input and contributions from the
+user community through the mothur discussion forum and GitHub issue
+tracker. As mothur has matured, we have been dependent on the user
+community to use these resources to tell us what features they would
+like to see included in mothur. The user community also tells us where
+our documentation is confusing. Often we can count on people not
+directly affiliated with mothur to provide instruction and their own
+experience to other users. We are constantly trying to recruit our
+``army'' and are happy to take any contributions we can. Whether the
+contributions are to the code base, discussion forum, or suggestions for
+new tools, these contributions have been invaluable to the growth and
+popularity of mothur.
 
-***Failed experiments.*** If we never failed, we would not be trying hard enough.
-Over the past decade we have tried a number of experiments to improve the usability and utility of mothur.
-One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and venn diagrams depicting the overlap between microbial communities.
-I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote.
-Such visuals require far too much customization to be publication-quality.
-Although QIIME has incorporated visualization tools through the Emperor package, the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature.
-Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials.
-A second experiment we attempted was to create a graphical user interface (GUI) for running mothur.
-Forcing users to interact with mothur through the command line has been a significant hurdle for many.
-Unfortunately, the development effort required to create and maintain a GUI is significant and there is limited funding for such efforts.
-The newest version of QIIME (version 2) has emphasized interaction with the tools through a GUI and it remains to be seen how this experiment will go.
-Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks.
-This documentation is explicit in mothur as all commands and output is recorded in a logfile.
-Given the heightened recent focus on reproducibility we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project.
-Finally, we collaborated with programmers through Google Summer of Code to develop commands in mothur to implement the random forest and SVM machine learning algorithms.
-Similar to the challenges of developing attractive visuals, fitting the algorithms' hyperparameters, testing, and deploying the resulting models requires a significant amount of customization.
-Furthermore, this is an active area of research where methods are still being developed and improved.
-Thankfully, there are numerous R and Python packages that do a better job of developing these models.
-Again, we have put our efforts into developing instructional materials that mothur users can use to fit such models to their data.
-In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.
-In hindsight, our solution to these failure has been to provide tutorials to as a conduit between mothur and their goals.
+\textbf{\emph{Failed experiments.}} If we never failed, we would not be
+trying hard enough. Over the past decade we have tried a number of
+experiments to improve the usability and utility of mothur. One of our
+first experiments was to use mothur to generate standard vector graphic
+(SVG)-formatted files of heatmaps and venn diagrams depicting the
+overlap between microbial communities. I quickly realized that I would
+never put a mothur-generated figure into a manuscript I wrote. Such
+visuals require far too much customization to be publication-quality.
+Although QIIME has incorporated visualization tools through the Emperor
+package, the challenge of users taking default values has downsides as
+ordinations with black background or publishing 3-D ordinations in a 2-D
+medium litter the literature. Instead, we have encouraged users to use R
+packages to visualize mothur-generated results using the minimalR
+instructional materials. A second experiment we attempted was to create
+a graphical user interface (GUI) for running mothur. Forcing users to
+interact with mothur through the command line has been a significant
+hurdle for many. Unfortunately, the development effort required to
+create and maintain a GUI is significant and there is limited funding
+for such efforts. The newest version of QIIME (version 2) has emphasized
+interaction with the tools through a GUI and it remains to be seen how
+this experiment will go. Another downside of using a GUI is that there
+is a risk that reproducibility will suffer if users do not have a
+mechanism to document their mouse clicks. This documentation is explicit
+in mothur as all commands and output is recorded in a logfile. Given the
+heightened recent focus on reproducibility we have extended significant
+effort in developing instructional materials teaching users how to
+organize, document, and execute reproducible pipelines that allow a user
+to go from raw sequence data to a compiled manuscript with figures
+through the Riffomonas project. Finally, we collaborated with
+programmers through Google Summer of Code to develop commands in mothur
+to implement the random forest and SVM machine learning algorithms.
+Similar to the challenges of developing attractive visuals, fitting the
+algorithms' hyperparameters, testing, and deploying the resulting models
+requires a significant amount of customization. Furthermore, this is an
+active area of research where methods are still being developed and
+improved. Thankfully, there are numerous R and Python packages that do a
+better job of developing these models. Again, we have put our efforts
+into developing instructional materials that mothur users can use to fit
+such models to their data. In each of our ``failed'' experiments, the
+real problems were straying from what mothur does well and failing to
+grasp what we really wanted the innovation to do. In hindsight, our
+solution to these failure has been to provide tutorials to as a conduit
+between mothur and their goals.
 
-***Competition is good and healthy.***
+\textbf{\emph{Competition is good and healthy.}}
 
-* A lot of people have been working in this space - e.g. clustering algorithms
-* Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))
+\begin{itemize}
+\tightlist
+\item
+  A lot of people have been working in this space - e.g.~clustering
+  algorithms
+\item
+  Approaches to developing software have differed (e.g.~web-based
+  (greengenes/RDP), boutique (Quince), wrappers (QIIME))
+\end{itemize}
 
-What are we proud of?
-* Open source packages
-* Platform independent
-* Development of open instructional materials
-* Data driven development of methods
-* Database independent - allows people outside of our field to use mothur
-* Reproducibility
-* Helped create standards - when something is this popular, you force people to use a certain suite of tools and methods
+What are we proud of? * Open source packages * Platform independent *
+Development of open instructional materials * Data driven development of
+methods * Database independent - allows people outside of our field to
+use mothur * Reproducibility * Helped create standards - when something
+is this popular, you force people to use a certain suite of tools and
+methods
 
-The future
-* 16S rRNA gene sequencing is not dead
-* Always improving sequencing methods bring new problems
-* Larger datasets, need a replacement for naive open and closed reference clustering
-* Lingering controversies - singletons, ASVs, contamination
+The future * 16S rRNA gene sequencing is not dead * Always improving
+sequencing methods bring new problems * Larger datasets, need a
+replacement for naive open and closed reference clustering * Lingering
+controversies - singletons, ASVs, contamination
 
-Longterm funding
-* The development of mothur was enabled by initial funding from a grant to Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
-* NSF
-* NIH
-* Bootstrapping from other projects
+Longterm funding * The development of mothur was enabled by initial
+funding from a grant to Sogin from the Sloan Foundation to support his
+VAMPS (Visualization and Analysis of Microbial Population Structures)
+initiative. * NSF * NIH * Bootstrapping from other projects
 
@@
 
-## Acknowledgements
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 
-## References
+\hypertarget{references}{%
+\subsection{References}\label{references}}
+
 25 references
 
-<div id="refs"></div>
+\hypertarget{refs}{}
 
@@
 
-<!--- \includegraphics{figure_1.png} -->
+\textbf{Figure 1. Caption caption caption.} Footnotes footnotes
+footnotes.
 
-**Figure 1. Caption caption caption.** Footnotes footnotes footnotes.
+\end{document}

### 2019-09-19 submission/manuscript.Rmd
Add more text

@@
 
-At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski and Chemistry guy].
+At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski, Chemistry guy, Lyme disease mbio, (A)historical paper].
 By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on.
@@
 
-***Competition is good and healthy.***
+***Competition is good and healthy.*** From the beginning there have been online tools the Ribosomal Database Project (RDP), greengenes, and SILVA.
+These allowed users a straightforward method of comparing their data to those collected in a database.
+There are two primary downsides to these tools.
+First, researchers running the online tool must pay the computational expenses leading to slow process times as numerous users simultaneously attempt an analysis and when hardware becomes outdated.
+Eventually this limitation would result in the termination of the greengenes website.
+Second, these platforms provide a one-size-fits-all analysis.
+These tools only allow a user to analyze 16S and in some cases 18S rRNA gene sequences.
+If a users analyzes a different gene then the tool will not serve them.
+These observations resulted in two design goals we have had with mothur: bringing the analysis to a user's computer and separating a tool from the database.
+For example, we commonly use a sequence alignment method that was originally developed for greengenes, but use the SILVA databases.
+In addition, we offer the naive Bayesian classifier developed by the RDP and allow users to train it to any database they want, including customized databases.
+In both examples, users can align or classify non rRNA gene sequence data.
+As the bioinformatics tools have matured, both the RDP and SILVA offer pipelines for analyzing large datasets, albeit in one-size-fits-all black box implementations.
 
-* A lot of people have been working in this space - e.g. clustering algorithms
-* Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))
+With the growth in popularity of 16S rRNA gene sequencing there has naturally been an expansion in the number of people developing tools to analyze these data.
+Months after the paper describing mothur was released, QIIME was released with similar goals.
+Over the past 10 years, many have attempted to compare these two programs: Pepsi vs Coke, Apple vs Windows, etc.
+It is never clear which software is which brand and whether it is a complement or an insult.
+Regardless, both programs are very popular.
+From my perspective, most of the differences are cosmetic.
+QIIME is effectively a bundle of scripts to run other developers' software.
+For example, with QIIME v1, it was possible to run mothur as part of QIIME.
+One can also run the naive Bayesian classifier through QIIME using the original code developed by the RDP.
+From my perspective, this caused QIIME many problems because there were numerous software dependencies that had to be installed causing frustration for users.
+Although the QIIME developers would create virtual machines and packaging tools to simplify installation, these fixes required sophistication by the user.
+In contrast, when a user runs mothur, they are running mothur.
+The naive Bayesian classifier code that is in mothur is a rewritten version of the original code.
+When we rewrite someone's software we do it with an eye to improving performance, access, and utility for non-16S rRNA gene sequence data.
+For example, while 454 data was popular, PyroNoise was an effective tool for denoising flowgram data.
+Running the original code required a linux computer and knowledge of bash and Perl scripting.
+When we rewrote the code for mothur, we made it accessible to people using any operating system with a simple command interface.
+I have found that our approach requires significant developer effort, but saves considerable user effort.
+
+Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone boutique software tools for sequence curation (e.g.
+PyroNoise and DADA), chimera checking (e.g.
+UCHIME, ChimeraSlayer, and Perseus), and clustering (e.g.
+USEARCH and Swarm).
+Where possible, we have implemented these algorithms directly into mothur.
+We have also used this diversity of methods to perform head-to-head comparisons.
+Most notable is the area of clustering algorithms where there have been a large number of algorithms developed without an obvious method to objectively compare them.
+Through a series of studies, we applied an objective metric, the Matthew's Correlation Coefficient (MCC), to numerous algorithms for clustering sequences into OTUs.
+By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing algorithms.
+We have embraced the competition and diversity of all methods being used to analyze amplicon data as an opportunity to identify the strengths and weaknesses of the methods in an attempt to make recommendations to other researchers.
 
@@
 
-The future
-* 16S rRNA gene sequencing is not dead
-* Always improving sequencing methods bring new problems
-* Larger datasets, need a replacement for naive open and closed reference clustering
-* Lingering controversies - singletons, ASVs, contamination
+**The future.** I plan to continue developing mothur for as long as other researchers find it useful.
+One challenge of such a plan is maintaining the funding to support its development.
+The development of mothur was initially enabled by a grant to Mitch Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
+We used that seed funding to secure an NSF grant and then a grant from NIH for tool development as part of their Human Microbiome Project.
+Since that project expired in 201X, we have not had funding to specifically support mothur's development.
+I have been fortunate to have discretionary funds generated from other projects to help support mothur.
+Although there is funding for new tools, there appears to be little appetite by funders to support existing tools.
+Emblematic of this was the NIH program, Big Data To Knowledge (BD2K), which solicited proposals through the program announcement "Extended Development, Hardening and Dissemination of Technologies in Biomedical Computing, Informatics and Big Data Science (PA-14-156)".
+This opportunity appeared perfect, except that the National Institute of Allergy and Infectious Diseases (NIAID), the primary supporter of microbiome research at NIH, did not participate in the announcement.
+Tools like mothur have been successful and need mechanisms to continue to mature and support the needs of the research community.
 
-Longterm funding
-* The development of mothur was enabled by initial funding from a grant to Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
-* NSF
-* NIH
-* Bootstrapping from other projects
+As with anything in science, methods become passe.
+When we first developed mothur, tools like T-RFLP and DGGE were still commonly used.
+Today it would be hard to argue that data from those methods meaningfully advance a study relative to what one could get using 16S rRNA gene sequence data.
+Looking forward, many want to claim that amplicon sequencing is today's DGGE.
+They claim that instead researchers should move on to shotgun metagenomic sequence data.
+It is important to note that the two methods answer fundamentally different questions.
+Metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community and 16S rRNA gene sequence data describes the taxonomic composition.
+Both tools provide important information but they cannot easily replace each other.
+Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is much higher than that of amplicon data.
+For example, we analyzed 10,000 16S rRNA sequences from each of about 500 subjects.
+We can think of this as representing about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome).
+Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4 Tbp.
+Although such a sequencing effort is possible, albeit expensive, analyzing such a large dataset with an approach that captures the full genetic diversity of the community would be financially and technically prohibitive.
+Going forward, there is still a place for 16S rRNA gene sequencing.
+Although sequencing technologies will continue to evolve to capture longer and more high quality data, there will likely always be a need for characterizing the taxonomic diversity of microbial communities.
+With this in mind, there will always be a place for tools like mothur.
 
-Red Queen
+Of course this does not mean that such tools will remain static.
+We see three key areas that we will continue to help to develop.
+First, just as we adapted through the transitions from Sanger to 454 to MiSeq sequencing platforms, we must learn how to adapt data from PacBio, Oxford Nanopore, and other tools to our analysis pipelines.
+As with these earlier platforms, we must understand the error profiles so that they can be corrected.
+We have learned that moving forward requires that we maintain or improve sequence quality.
+No doubt, datasets and read lengths will improve.
+Second, with these improvements, we will need to continue to improve our algorithms.
+We have already seen that attempts to use low quality MiSeq and HiSeq data causes computational problems leading to the creation of open and closed reference clustering methods.
+Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods.
+More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality.
+Finally, there are ongoing controversies that need further exploration.
+These include the validity and utility of amplicon sequence variants, the validity and impact of removing low frequency sequences, and methods of identifying and removing 16S rRNA gene sequences from contaminants.
+In the original mothur paper, we commented that 16S rRNA gene sequencing and analysis is very much like the Red Queen in Alice in Wonderland.
+Although some disagreed with this analogy, we still feel it fits.
+The sequencing technology and rapacious appetite of researchers continues to race on.
+At the same time, bioinformatics tools must adapt to facilitate our research.
+I am confident that mothur will be up to this exciting challenge.
 

### 2019-09-19 submission/manuscript.tex
Add more text

@@
 be instructive to other scientists who are overwhelmed at the prospect
-of looking forward at their careers {[}Lenski and Chemistry guy{]}. By
-no means is my personal journey over, but since 2002 I have been on a
-journey that I did not realize I was on. Now that the paper introducing
-the mothur software package is ten years old and has become the most
-cited paper published by \emph{Applied and Environmental Microbiology},
-it is worth stepping back and using the development of mothur as a story
-that likely has parallels to many other research stories that have taken
-time to develop.
+of looking forward at their careers {[}Lenski, Chemistry guy, Lyme
+disease mbio, (A)historical paper{]}. By no means is my personal journey
+over, but since 2002 I have been on a journey that I did not realize I
+was on. Now that the paper introducing the mothur software package is
+ten years old and has become the most cited paper published by
+\emph{Applied and Environmental Microbiology}, it is worth stepping back
+and using the development of mothur as a story that likely has parallels
+to many other research stories that have taken time to develop.
 
@@
 
-\textbf{\emph{Competition is good and healthy.}}
+\textbf{\emph{Competition is good and healthy.}} From the beginning
+there have been online tools the Ribosomal Database Project (RDP),
+greengenes, and SILVA. These allowed users a straightforward method of
+comparing their data to those collected in a database. There are two
+primary downsides to these tools. First, researchers running the online
+tool must pay the computational expenses leading to slow process times
+as numerous users simultaneously attempt an analysis and when hardware
+becomes outdated. Eventually this limitation would result in the
+termination of the greengenes website. Second, these platforms provide a
+one-size-fits-all analysis. These tools only allow a user to analyze 16S
+and in some cases 18S rRNA gene sequences. If a users analyzes a
+different gene then the tool will not serve them. These observations
+resulted in two design goals we have had with mothur: bringing the
+analysis to a user's computer and separating a tool from the database.
+For example, we commonly use a sequence alignment method that was
+originally developed for greengenes, but use the SILVA databases. In
+addition, we offer the naive Bayesian classifier developed by the RDP
+and allow users to train it to any database they want, including
+customized databases. In both examples, users can align or classify non
+rRNA gene sequence data. As the bioinformatics tools have matured, both
+the RDP and SILVA offer pipelines for analyzing large datasets, albeit
+in one-size-fits-all black box implementations.
 
-\begin{itemize}
-\tightlist
-\item
-  A lot of people have been working in this space - e.g.~clustering
-  algorithms
-\item
-  Approaches to developing software have differed (e.g.~web-based
-  (greengenes/RDP), boutique (Quince), wrappers (QIIME))
-\end{itemize}
+With the growth in popularity of 16S rRNA gene sequencing there has
+naturally been an expansion in the number of people developing tools to
+analyze these data. Months after the paper describing mothur was
+released, QIIME was released with similar goals. Over the past 10 years,
+many have attempted to compare these two programs: Pepsi vs Coke, Apple
+vs Windows, etc. It is never clear which software is which brand and
+whether it is a complement or an insult. Regardless, both programs are
+very popular. From my perspective, most of the differences are cosmetic.
+QIIME is effectively a bundle of scripts to run other developers'
+software. For example, with QIIME v1, it was possible to run mothur as
+part of QIIME. One can also run the naive Bayesian classifier through
+QIIME using the original code developed by the RDP. From my perspective,
+this caused QIIME many problems because there were numerous software
+dependencies that had to be installed causing frustration for users.
+Although the QIIME developers would create virtual machines and
+packaging tools to simplify installation, these fixes required
+sophistication by the user. In contrast, when a user runs mothur, they
+are running mothur. The naive Bayesian classifier code that is in mothur
+is a rewritten version of the original code. When we rewrite someone's
+software we do it with an eye to improving performance, access, and
+utility for non-16S rRNA gene sequence data. For example, while 454 data
+was popular, PyroNoise was an effective tool for denoising flowgram
+data. Running the original code required a linux computer and knowledge
+of bash and Perl scripting. When we rewrote the code for mothur, we made
+it accessible to people using any operating system with a simple command
+interface. I have found that our approach requires significant developer
+effort, but saves considerable user effort.
+
+Beyond the large packages like mothur and QIIME, there has been
+significant growth in stand alone boutique software tools for sequence
+curation (e.g.~PyroNoise and DADA), chimera checking (e.g.~UCHIME,
+ChimeraSlayer, and Perseus), and clustering (e.g.~USEARCH and Swarm).
+Where possible, we have implemented these algorithms directly into
+mothur. We have also used this diversity of methods to perform
+head-to-head comparisons. Most notable is the area of clustering
+algorithms where there have been a large number of algorithms developed
+without an obvious method to objectively compare them. Through a series
+of studies, we applied an objective metric, the Matthew's Correlation
+Coefficient (MCC), to numerous algorithms for clustering sequences into
+OTUs. By performing this type of analysis, we were able to objectively
+compare the algorithms, make recommendations to the field, and develop
+new algorithms that outperformed the existing algorithms. We have
+embraced the competition and diversity of all methods being used to
+analyze amplicon data as an opportunity to identify the strengths and
+weaknesses of the methods in an attempt to make recommendations to other
+researchers.
 
@@
 
-The future * 16S rRNA gene sequencing is not dead * Always improving
-sequencing methods bring new problems * Larger datasets, need a
-replacement for naive open and closed reference clustering * Lingering
-controversies - singletons, ASVs, contamination
+\textbf{The future.} I plan to continue developing mothur for as long as
+other researchers find it useful. One challenge of such a plan is
+maintaining the funding to support its development. The development of
+mothur was initially enabled by a grant to Mitch Sogin from the Sloan
+Foundation to support his VAMPS (Visualization and Analysis of Microbial
+Population Structures) initiative. We used that seed funding to secure
+an NSF grant and then a grant from NIH for tool development as part of
+their Human Microbiome Project. Since that project expired in 201X, we
+have not had funding to specifically support mothur's development. I
+have been fortunate to have discretionary funds generated from other
+projects to help support mothur. Although there is funding for new
+tools, there appears to be little appetite by funders to support
+existing tools. Emblematic of this was the NIH program, Big Data To
+Knowledge (BD2K), which solicited proposals through the program
+announcement ``Extended Development, Hardening and Dissemination of
+Technologies in Biomedical Computing, Informatics and Big Data Science
+(PA-14-156)''. This opportunity appeared perfect, except that the
+National Institute of Allergy and Infectious Diseases (NIAID), the
+primary supporter of microbiome research at NIH, did not participate in
+the announcement. Tools like mothur have been successful and need
+mechanisms to continue to mature and support the needs of the research
+community.
 
-Longterm funding * The development of mothur was enabled by initial
-funding from a grant to Sogin from the Sloan Foundation to support his
-VAMPS (Visualization and Analysis of Microbial Population Structures)
-initiative. * NSF * NIH * Bootstrapping from other projects
+As with anything in science, methods become passe. When we first
+developed mothur, tools like T-RFLP and DGGE were still commonly used.
+Today it would be hard to argue that data from those methods
+meaningfully advance a study relative to what one could get using 16S
+rRNA gene sequence data. Looking forward, many want to claim that
+amplicon sequencing is today's DGGE. They claim that instead researchers
+should move on to shotgun metagenomic sequence data. It is important to
+note that the two methods answer fundamentally different questions.
+Metagenomic sequence data tells a researcher about the functional
+potential and genetic diversity of a community and 16S rRNA gene
+sequence data describes the taxonomic composition. Both tools provide
+important information but they cannot easily replace each other.
+Although metagenomic data does provide highly resolved taxonomic
+information, the limit of detection is much higher than that of amplicon
+data. For example, we analyzed 10,000 16S rRNA sequences from each of
+about 500 subjects. We can think of this as representing about 1,000,000
+genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S
+rRNA gene sequences/genome). Assuming a genome is 4 Mbp, this would
+represent a sequencing depth of 4 Tbp. Although such a sequencing effort
+is possible, albeit expensive, analyzing such a large dataset with an
+approach that captures the full genetic diversity of the community would
+be financially and technically prohibitive. Going forward, there is
+still a place for 16S rRNA gene sequencing. Although sequencing
+technologies will continue to evolve to capture longer and more high
+quality data, there will likely always be a need for characterizing the
+taxonomic diversity of microbial communities. With this in mind, there
+will always be a place for tools like mothur.
 
-Red Queen
+Of course this does not mean that such tools will remain static. We see
+three key areas that we will continue to help to develop. First, just as
+we adapted through the transitions from Sanger to 454 to MiSeq
+sequencing platforms, we must learn how to adapt data from PacBio,
+Oxford Nanopore, and other tools to our analysis pipelines. As with
+these earlier platforms, we must understand the error profiles so that
+they can be corrected. We have learned that moving forward requires that
+we maintain or improve sequence quality. No doubt, datasets and read
+lengths will improve. Second, with these improvements, we will need to
+continue to improve our algorithms. We have already seen that attempts
+to use low quality MiSeq and HiSeq data causes computational problems
+leading to the creation of open and closed reference clustering methods.
+Unfortunately, comparative analyses showed that these methods fail
+relative to \emph{de novo} clustering methods. More work is needed to
+improve reference-based clustering methods so that larger datasets can
+be analyzed without sacrificing quality. Finally, there are ongoing
+controversies that need further exploration. These include the validity
+and utility of amplicon sequence variants, the validity and impact of
+removing low frequency sequences, and methods of identifying and
+removing 16S rRNA gene sequences from contaminants. In the original
+mothur paper, we commented that 16S rRNA gene sequencing and analysis is
+very much like the Red Queen in Alice in Wonderland. Although some
+disagreed with this analogy, we still feel it fits. The sequencing
+technology and rapacious appetite of researchers continues to race on.
+At the same time, bioinformatics tools must adapt to facilitate our
+research. I am confident that mothur will be up to this exciting
+challenge.
 

### 2019-09-20 submission/manuscript.Rmd
First full draft

@@
 ## Abstract
-75 words
+250 words
 
@@
 \newpage
-
-3000 words total
 
@@
 In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.
-In hindsight, our solution to these failure has been to provide tutorials to as a conduit between mothur and their goals.
 
@@
 Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone boutique software tools for sequence curation (e.g.
-PyroNoise and DADA), chimera checking (e.g.
-UCHIME, ChimeraSlayer, and Perseus), and clustering (e.g.
-USEARCH and Swarm).
+PyroNoise, DADA), chimera checking (e.g.
+UCHIME, ChimeraSlayer, Perseus), and clustering (e.g.
+USEARCH, Swarm).
 Where possible, we have implemented these algorithms directly into mothur.
@@
 
-What are we proud of?
-* Open source packages
-* Platform independent
-* Development of open instructional materials
-* Data driven development of methods
-* Database independent - allows people outside of our field to use mothur
-* Reproducibility
-* Helped create standards - when something is this popular, you force people to use a certain suite of tools and methods
+***mothur's core principles.*** As mothur has evolved with the needs of the community and matured, several core principles have emerged that direct its development.
+First, mothur is a free, open source software package.
+This has been critical in shaping the direction of mothur.
+We were content for mothur to be an improved combination of DOTUR and SONS and leverage existing tools for other steps.
+Yet, when we learned that the code for NAST, the algorithm behind the greengenes's aligner,  was not open source or publicly available, we realized that such an important tool needed to be opened to the community to make the tool adaptable to other databases and potential improvements.
+More recently, the rejection of closed source tools such as USEARCH can be seen by the broader adoption of its competitor, VSEARCH, within the microbial ecology community.
+Related to insuring that mothur's code is open source, our second core principle is that we maintain transparency for our users.
+Perhaps a user does not need to interrogate every line of code, but we do feel that they need to understand what is happening.
+Many programs including QIIME and DADA2 encapsulate large elements of a pipeline in a single command.
+In contrast, mothur forces the user to specify each step of the pipeline.
+Although the former approach makes an analysis easier for a beginner, it potentially stifles users that need greater control or understanding of the assumptions for each element of the pipeline.
+This control over the pipeline has made it easier for researchers to customize databases or adapt the pipeline to analyze non 16S rRNA gene sequence data.
+Third, as I mentioned above, there has been a plethora of methods proposed for generating amplicon sequence data, and curating, aligning, checking for chimeras, classifying and clustering the data.
+I am proud of the data-driven approach we have taken to comparing these methods.
+A description of a new method is of limited value if it is not benchmarked against other methods or control datasets.
+Through this core principle and mothur's large reach into the community, we have helped to develop standards in the analysis of 16S rRNA gene sequence data.
+Fourth, a focus on enabling reproducibility has always been central to the functionality of mothur.
+From the beginning, mothur's logfiles have represented a transcript of the user's command and outputs.
+When researchers were reluctant to submit sequence data to the Sequence Read Archive (SRA), we worked with the SRA developers to create a mothur command (make.sra) to create the package to submit sequence data through a special mothur portal.
+A more ambitious project had its seed on April 1, 2013 when we announced a new "function" in mothur: write.paper.
+The new command required that the user provide a 454 sff file and a journal title or impact factor.
+With this information, mothur would generate a manuscript.
+This April Fools joke was poking fun at software that provided an analysis black box but also at the sentiment that data analysis should be so cut and dry.
+A few years later, we revisited this concept in the scope of reproducibility.
+Why not explicitly script an analysis from downloading data from the Sequence Read Archive (SRA) through the rendering of a manuscript ready for submission?
+This idea gave rise to the development of the Riffomonas reproducible research tutorial series that enables researchers to write their own write.paper program.
+Perhaps the most important core principle is that my research group uses mothur to analyze the data we generate.
+We "eat our own dogfood".
+This has proven critical as it again represents transparency to our users and hopefully provides confidence to users that we are not making recommendations that we do not follow ourselves.
 
-**The future.** I plan to continue developing mothur for as long as other researchers find it useful.
+***The future.*** Buoyed by these principles, I plan to continue developing mothur for as long as other researchers find it useful.
 One challenge of such a plan is maintaining the funding to support its development.
@@
 Looking forward, many want to claim that amplicon sequencing is today's DGGE.
-They claim that instead researchers should move on to shotgun metagenomic sequence data.
+They claim that researchers should instead move on to shotgun metagenomic sequencing.
 It is important to note that the two methods answer fundamentally different questions.
@@
 We see three key areas that we will continue to help to develop.
-First, just as we adapted through the transitions from Sanger to 454 to MiSeq sequencing platforms, we must learn how to adapt data from PacBio, Oxford Nanopore, and other tools to our analysis pipelines.
+First, just as we adapted through the transitions from Sanger to 454 to MiSeq sequencing platforms, we must learn whether data from PacBio and Oxford Nanopore are acceptable alternatives and then our analysis pipelines to facilitate their analysis.
 As with these earlier platforms, we must understand the error profiles so that they can be corrected.
@@
 These include the validity and utility of amplicon sequence variants, the validity and impact of removing low frequency sequences, and methods of identifying and removing 16S rRNA gene sequences from contaminants.
-In the original mothur paper, we commented that 16S rRNA gene sequencing and analysis is very much like the Red Queen in Alice in Wonderland.
+In the paper announcing mothur we commented that the relationship between 16S rRNA gene sequencing and analysis is very much like the Red Queen in Alice in Wonderland.
 Although some disagreed with this analogy, we still feel it fits.
@@
 I am confident that mothur will be up to this exciting challenge.
+
+***Conclusion.*** Beyond its utility for analyzing amplicon sequence data, I believe that mothur's history provides lessons that are helpful for other projects with a long historical arc.
+First, mothur is a product of its time.
+It has always sought to solve a current need to the best of our ability with the tools we had at the time.
+There are certainly caveats to any analysis of 16S rRNA gene sequence data, but if we had waited until those caveats were resolved, the field never would have progressed.
+Similarly, we made design choices that we probably would not have made had we started the project today.
+Second, as we have developed mothur, we have attempted to do so in a data driven approach where we compare multiple competing methods to each other.
+It has not merely been enough to propose a new method, we must show that it meaningfully advances the field.
+Third, through our failures and successes we have learned to focus on what mothur is good at and create alternative products to help users.
+For example, mothur will not have a graphical interface.
+Instead, we will provide instructional materials to teach users how to use the command line interface and other programs like R to generate a reproducible workflow.
+Finally, mothur was born out of a need for automating the analysis of large 16S rRNA gene sequence datasets.
+I realized I had a set of skills to fill that need.
+It has been refreshing to see the computational skills of the microbial ecology field grow over the past decade.
+Looking ahead, we must all take stock of the challenges we face in microbial ecology and how our unique skills and interests can address these challenges to turn them into opportunities.
 

### 2019-09-30 submission/manuscript.Rmd
Editing
Edited manuscript for wording, structure, and to add references.

--- 2019-09-16 submission/manuscript.Rmd
@@
   pdf_document:
+    latex_engine: xelatex
     keep_tex: true
@@
 
-## Observation format
+### Mini-review
 
@@
 
-3000 words total
-
-Few scientists set out on a nearly two decade long journey with a specific goal in mind.
+Few scientists set out on a nearly two decade-long journey with a specific goal in mind.
 Often we fail to start a scientific journey because it looks too hard.
-Perhaps we get bogged down in all of the things that could go wrong.
-Perhaps we go astray from the path because we find something else that appears more interesting.
+Perhaps we get discouraged by all of the things that could go wrong.
+Maybe we stray from the path because we find something that is more interesting.
 Every scientist picks their own path and takes their own forks in the road.
 From the outside, it may appear to be a random walk.
-Nevertheless, these meandering journeys in are common in science.
+Nevertheless, these meanderings are common in science.
 
-At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski and Chemistry guy].
+Looking back on scientific journeys can be instructive to others who are overwhelmed at the prospect of looking forward at their careers [@Lenski2017; @Smith2018; @Barbour2019, @Casadevall2015; @Colwell2014; @Glckner2017].
 By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on.
-Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology*, it is worth stepping back and using the development of mothur as a story that likely has parallels to many other research stories that have taken time to develop.
+Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology* [@Schloss2009a], it is worth stepping back and using the continued development of mothur as a story that has parallels to many other research stories.
 
@@
 She asked if I could add one of those "curvy things", a rarefaction curve, to show where I was in sampling the community.
-Rarefaction curves and attempts to estimate the taxonomic richness of soil had become popular because of the simple, but impactful mini-review by Jennifer Hughes and her colleagues, which introduced the field to operational taxonomic units (OTUs), rarefaction curves, and richness estimates [DOI: 10.1128/AEM.67.10.4399-4406.2001].
-I do not recall whether that poster had a rarefaction curve on it, but her question primed my career.
+Rarefaction curves and attempts to estimate the taxonomic richness of soil had become popular because of the impactful review by Jennifer Hughes and her colleagues [@Hughes2001].
+Their seminal paper introduced the field to operational taxonomic units (OTUs), rarefaction curves, and richness estimates.
+I do not recall whether my poster had a rarefaction curve on it, but Jo's question primed my career.
 
@@
 How would I bin the sequences into OTUs?
-Hughes and her colleagues did it manually for datasets that had fewer than 284 sequences.
+Hughes and her colleagues did it manually and with fewer than 300 sequences.
 Although I could possibly do that for my 600 sequences, my goal was to generate 1,000 sequences from the sample and to repeat that sampling effort for other samples.
 I needed something that could be automated.
-Furthermore, the software that Hughes used, EstimateS, required a series of tedious data formatting steps.
+Furthermore, the software that Hughes used, EstimateS [@Colwell2014], required a series of tedious data formatting steps to perform the analyses we were interested in performing.
 I had found my first problem.
 How would I assign sequences to OTUs and use that data to estimate the richness and diversity of a sample?
-The second problem would be how could I compare the sequences found in one sample to another sample?
-The solution to the first problem, DOTUR (Distance-based OTUs and Richness), took us two years to develop [10.1128/AEM.71.3.1501-1506.2005].
-DOTUR did two things: given a matrix describing the genetic distance between pairs of sequences, it would cluster those sequences into OTUs for any distance threshold to define the OTUs and then it would use the frequency of each OTU to calculate a variety of alpha diversity metrics.
-The solutions to the second problem would come from our work to develop software including S-LIBSHUFF [2004], SONS (Shared OTUs and Similarity) [2006], and TreeClimber [2006].
-Around the same time, Catherine Lozupone and Rob Knight were developing their UniFrac tools to compare communities with a phylogenetic rather than OTU-based approach [PMID: 16332807; PMID: 17220268].
+The second problem would involve comparing the abundance of OTUs found in one sample to another sample.
+The solution to the first problem, DOTUR (Distance-based OTUs and Richness), took us two years to develop [@Schloss2005].
+DOTUR did two things: given a matrix quantifying the genetic distance between pairs of sequences, it would cluster those sequences into OTUs for any distance threshold to define the OTUs and then it would use the frequency of each OTU to calculate a variety of alpha diversity metrics.
+The solutions to the second problem would come from our work to develop software including $\int$-LIBSHUFF [@Schloss2004a], SONS (Shared OTUs and Similarity) [@Schloss2006b], and TreeClimber [@Schloss2006a].
+Around the same time, Catherine Lozupone and Rob Knight were developing their UniFrac tools to compare communities with a phylogenetic rather than OTU-based approach [@Lozupone2005; @Lozupone2007].
 With these tools, the field of microbial ecology had a quantitative toolbox for describing and comparing microbial communities.
+Along the way Jo and I would demonstrate the utility of such tools to answer questions like how many OTUs were there in that sample of Alaskan soil and how many sequences were needed to sample each of those OTUs [@Schloss2006c]?
+Where were we in the global bacterial census [@Schloss2004b]?
+How does the word usage of *Goodnight, Moon* compare to that of *Portrait of a Lady* and more importantly how is this relevant to microbial ecology [@Schloss2007]?
+Most edifying were the more than 2,400 papers that used DOTUR, SONS, TreeClimber, or $\int$-LIBSHUFF to facilitate their own research questions (Web of Science, 10/1/2019).
+Had we waited to solve all of the problems that plagued 16S rRNA gene sequencing, we would still be waiting.
 
 It is important to remember that we knew there were many problems with 16S rRNA gene sequencing.
-We knew there were biases from extractions and amplification [XXXX].
-We knew there were chimeras [XXXX].
-Getting to the distance matrix required trimming and correcting sequence errors, aligning them, and finding the most appropriate way to calculate a pairwise distance.
-It was also rare to have experimental replication to perform statistical tests to compare treatment groups.
-We frequently used a dataset comparing Scottish soils from from Alison McCaig and colleagues.
-This dataset consisted of two experimental groups, each replicated three times with 45 sequences in each replicate.
-Nevertheless, we had excuses and work arounds for these problems that served our needs.
-At the time, I felt that the biggest problems were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
-Along the way we would demonstrate the utility of such tools to answer questions like where are we in the bacterial census?
-How many sequences would it take to see every OTU in that sample of Alaskan soil?
-How does the word usage of *Goodnight, Moon* compare to that of *Portrait of a Lady*?
-More importantly, 1,900 papers used DOTUR to facilitate their own research questions.
-Had we waited to solve all of the problems that plague 16S rRNA gene sequencing, we would still be waiting.
+We knew there were biases from extractions and amplification [@Zhou1996; @Suzuki1996; @Chandler1997; @Polz1998; @Wagner1994; @Hansen1998].
+We knew there were chimeras [@Qiu2001; @Komatsoulis1997; @Wang1997; @Hugenholtz2003].
+We knew that bacteria varied in their *rrn* copy number.
+Generating a distance matrix was a prerequisite to using my tools.
+This wasn't trivial, but by cobbling together other tools it was possible.
+We would assemble, trim and and correct Sanger sequence reads using Chromas or STADEN [@Bonfield1995], align the sequences using ClustalW [@Thompson1994] or ARB [@Ludwig2004], check for chimeras using partial treeing or Bellerephon [@Hugenholtz2003], and calculate a pairwise distance matrix using DNADIST from the PHYLIP package [@Felsenstein1989].
+At the time, we knew that we only had a loose concept of a species based on these distances [@Stackebrandt1994].
+We hoped that an OTU defined as a group of sequences more than 97% similar to each other would be a biologically meaningful unit regardless of whether it fit our notion of a bacterial species.
+At the time, I felt that the biggest problems that I could solve were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
+The only tool available at the time that automated the clustering step was FastGroup, which implemented a greedy version of the single linkage algorithm [@Seguritan2001].
+The high cost of sequencing was also an impediment to experimentation and analysis in microbial ecology.
+It was rare for a study design to have experimental replicates so that one could perform a statistical test to compare treatment groups.
+For example, in our testing we frequently used a dataset comparing Scottish soils from from Alison McCaig and colleagues [@McCaig1999].
+This dataset consisted of two experimental groups, each replicated three times with 45 sequences per replicate.
+Although great focus has been placed on the depth of sampling afforded by 454 and Illumina sequencing, the true benefit of the modern sequencing platforms is the ability to affordably sequence a large number of technical and biological replicates.
+In spite of the many technical challenges, we had excuses and heuristics to solve problems that served our needs.
+It is telling that a recent review of "best practices" in generating and analyzing 16S rRNA gene sequences shows that we still have not solved many of these issues and that we have even identified additional problems [@Pollock2018].
 
 As we developed these tools, I found a unique niche in microbiology.
-I believe that my undergraduate and graduate training as a biological engineer prepared me to think about research questions from a systems perspective, to think quantitatively, and to understand the value of using computer programs to help solve problems.
-As an undergraduate engineering student, I learned the Pascal programming language and promptly forgot much of it as an engineering graduate student.
-As a postdoc, I learned the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [DOI: 10.1128/AEM.70.9.5485-5492.2004].
-After writing my own version of LIBSHUFF and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
+My undergraduate and graduate training as a biological engineer prepared me to think about research questions from a systems perspective, to think quantitatively, and to understand the value of using computer programs to help solve problems.
+As an undergraduate student, I learned the Pascal programming language and promptly forgot much of it as a graduate student and learned MatLab in its place.
+As a postdoc, I learned the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [@Singleton2001].
+After writing my own version of LIBSHUFF, $\int$-LIBSHUFF and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
 At the time, the conversion from Perl to C++ seemed like an academic exercise to learn a new language.
@@
 It was also a skill that was virtually unheard of in microbiology.
-Today researchers without the ability to program in Python or R are at a significant disadvantage.
+Today, researchers without the ability to program are at a significant disadvantage [@Carey2018].
 
-***Introducing mothur.*** Shortly after DOTUR was published, I received an email from Mitch Sogin asking whether DOTUR could handle more than a million sequences.
+***Introducing mothur.*** Shortly after DOTUR was published, I received an email from Mitch Sogin, a scientist at the Marine Biology Laboratory (Woods Hole, MA), who asked whether DOTUR could handle more than a million sequences.
 Without answering his question, I asked where he found a million sequences.
 Little did I know that his email would represent another pivot in the development of these tools.
-His group would be the first to use 454 sequencing technology to generate 16S rRNA gene sequences [PMID: 16880384].
-Although mothur could assign those sequences to OTUs, it was slow and required a significant amount of RAM.
-As I left my postdoc to start my independent career across the state from Sogin's lab at the University of Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS, S-LIBSHUFF, and TreeClimber for the new world of massively parallelized sequencing.
-The new tool was mothur.
+His group would be the first to use 454 sequencing technology to generate 16S rRNA gene sequences [@Sogin2006].
+Although DOTUR could assign those sequences to OTUs, at the scale of millions of sequences, it was slow and required a significant amount of RAM.
+As I left my postdoc to start my independent career across the state from Sogin's lab at the University of Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS, $\int$-LIBSHUFF, and TreeClimber for the new world of massively parallelized sequencing.
+The new tool would become mothur.
 
-Milling about at a poster session at an ASM General Meeting in New Orleans, I again ran into Mitch who asked what my plans were for new tools.
-I told him that I wanted to make a tool like ARB (a powerful database tool and phylogenetics package), but for microbial ecology analysis.
-His retort was, "You and what army?" To that point, I had written every line of code and been answering many emails from people asking for help.
-It would be difficulty, but I needed to learn to let go and share the development process with someone else.
+Milling about at a poster session at an ASM General Meeting in New Orleans, I ran into Mitch who asked what my plans were for my new lab.
+I told him that I wanted to make a tool like ARB, a powerful database tool and phylogenetics package [@Ludwig2004], but for microbial ecology analysis.
+His retort was, "You and what army?" Up to that point, I had written every line of code and been answering many emails from people asking for help.
 He was right, I would need an army.
-That "army" ended up being Sarah Westcott who has worked on the mothur project largely from its inception.
-Today, mothur is over 200,000 lines of code and Sarah has touched or written nearly every line of code.
-Beyond writing and testing mothur's code base, she has become a conduit for many learning the tools of microbial ecology by patiently answering questions via email and the package's discussion forum.
+It would be difficult, but I needed to learn to let go and share the development process with someone else.
+My "army" ended up being Sarah Westcott who has worked on the mothur project from its inception.
+Today, mothur is over 200,000 lines of code and Sarah has touched or written nearly every line of it.
+Beyond writing and testing mothur's code base, she has become a conduit for many who are trying to learn the tools of microbial ecology.
+She patiently answers questions via email and on the package's discussion forum (https://forum.mothur.org).
 The community and I are lucky that Sarah has stayed with the project for more than a decade.
@@
 In hindsight, it would have been better to have developed mothur with more of an "army" or team so that there is overlap in people's understanding of how mothur works.
-Although such an arrangement might work in a software engineering firm, it is not practical in an academic setting where funding is limited for developers of free, open source software packages.
+Although a distributed team approach might work in a software engineering firm, it is not practical in most academic environments where there is limited funding.
 There are certainly projects that make this work, but they are rare.
+
+***Competition has been good and healthy.*** mothur has not been developed in a vacuum and it does not have a monopoly within the field.
+As indicated above, each of our decisions were made in the historical context of the field and with constant pressure from others developing their own tools for analyzing 16S rRNA gene sequence data.
+Competition has been good for mothur and for the field.
+
+From the beginning there have been online tools available at the Ribosomal Database Project (RDP) [@Cole2013], greengenes [@DeSantis2006a], and SILVA [@Yilmaz2013].
+These allowed users a straightforward method of comparing their data to those collected in a database.
+There are two primary downsides to these tools.
+First, researchers running the online tool must pay the computational expenses leading to slow process times when hardware becomes outdated and when numerous users simultaneously attempt their analysis.
+Eventually this limitation would result in the termination of the greengenes website.
+Second, these platforms provide a one-size-fits-all analysis.
+These tools only allow a user to analyze 16S and in some cases 18S rRNA gene sequences.
+If a user sequences a different gene, then the tool will not serve them.
+These observations resulted in two design goals we have had with mothur: bringing the analysis to a user's computer and separating a tool from a specific database.
+For example, we commonly use a sequence alignment method that was originally developed for greengenes [@DeSantis2006b], but use a SILVA-based reference alignment because its superior quality [@Schloss2009b; @Schloss2010a].
+In addition, we offer the \text{na\"ive} Bayesian classifier developed by the RDP [@Wang2007] and allow users to train it to any database they want, including customized databases.
+In both examples, users can align or classify non-rRNA gene sequence data.
+As the bioinformatics tools have matured, both the RDP and SILVA offer integrated pipelines for analyzing large datasets, albeit in one-size-fits-all black box implementations.
+
+With the growth in popularity of 16S rRNA gene sequencing there has naturally been an expansion in the number of people developing tools to analyze these data.
+Months after the paper describing mothur was published, the paper describing QIIME was published [@Caporaso2010a].
+Over the past 10 years, many have attempted to create analogies comparing the two programs: Pepsi vs Coke, Apple vs Windows, etc.
+It is never clear which software is which brand and whether the comparisons are meant as a complement or an insult.
+Regardless, both programs are very popular.
+From my perspective, most of the differences are cosmetic.
+QIIME is effectively a bundle of scripts to run other developers' software.
+For example, with QIIME (through version 1.9.1), it was even possible to run mothur through QIIME.
+One can also run the \text{na\"ive} Bayesian classifier through QIIME using the original code developed by the RDP.
+This caused great frustration for many users because there were numerous software dependencies that had to be installed.
+Although the QIIME developers would go on to create virtual machines and use packaging tools to simplify installation, these fixes required sophistication by users who we knew struggled with the basics of navigating a command line.
+In contrast, when a user runs mothur, they are running mothur.
+The \text{na\"ive} Bayesian classifier code that is in mothur is a rewritten version of the original code.
+When we rewrite someone's software we do it with an eye to improving performance, access, and utility for non-16S rRNA gene sequence data.
+For example, while 454 data was popular, PyroNoise was an effective tool for denoising flowgram data [@Quince2009].
+Running the original code required a large Linux computer cluster and knowledge of bash and Perl scripting.
+When we rewrote the code for mothur, we made it accessible to people using any operating system with a simple command interface (i.e.
+trim.flows and shhh.flows).
+Our approach requires significant developer effort, but saves considerable user effort.
+As this benefit is multiplied across thousands of projects, the savings to users has been considerable.
+
+Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone software tools for sequence curation (e.g.
+PyroNoise [@Quince2009], PANDAseq [@Masella2012], DADA2 [@Callahan2016]), chimera checking (e.g.
+UCHIME [@Edgar2011], ChimeraSlayer [@Haas2011], Perseus [@Quince2011]), and clustering (e.g.
+USEARCH [@Edgar2010], VSEARCH [@Rognes2016], Swarm [@Mah2015]).
+Where possible and when warrented, we have implemented many of these algorithms directly into mothur.
+We have also used this diversity of methods to perform head-to-head comparisons.
+Most notable is the area of clustering algorithms where there have been a large number of algorithms developed without an obvious method to objectively compare them [@Schloss2011a; @Westcott2015; @Schloss2016a; @Westcott2017].
+We applied an objective metric, the Matthew's Correlation Coefficient (MCC), to evaluate numerous algorithms for clustering sequences into OTUs.
+By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing algorithms.
+Beyond evaluating clustering algorithms, we have also evaluated methods of denoising sequence data [@Kozich2013; @Schloss2011b;  @Schloss2016b], assessed reference alignments [@Schloss2009b;  @Schloss2010a], considered the importance of incorporating secondary structure information in alignments [@Schloss2012a], quantified the variation along the 16S rRNA gene [@Schloss2010a], and compared the statistical hypotheses tested by commonly used tools [@Schloss2008].
+We have embraced the competition and diversity of all methods being used to analyze amplicon data as an opportunity to identify the strengths and weaknesses of the methods in an attempt to make recommendations to other researchers.
+
+***mothur's core principles.*** As mothur has evolved with the needs of the community, several core principles have emerged that direct its development.
+First, mothur is a free, open source software package.
+This has been critical in shaping the direction of mothur.
+We were content for mothur to be an improved combination of DOTUR and SONS and leverage existing tools for other steps.
+Yet, when we learned that the code for NAST, the algorithm behind the greengenes's aligner [@DeSantis2006b],  was not open source or publicly available.
+Similarly, although the SINA aligner was available through ARB and the SILVA website performed well it was closed source.
+Because the ARB implementation did not scale to large datasets, researchers were left to pay a processing fee to the SILVA website to align sequences.
+Thus, we realized that such an important tool needed to be opened to the community [@Schloss2009b].
+More recently, the rejection of closed source, commercial tools such as USEARCH can be seen by the broader adoption of its open source, free competitor, VSEARCH, within the microbial ecology community [@Edgar2010; @Rognes2016].
+Related to insuring that mothur's code is open source, our second core principle is that we maintain transparency to our users.
+Perhaps a user does not need to interrogate every line of code, but they need to understand what is happening.
+Many programs including online workflows encapsulate large elements of a pipeline in a single command.
+In contrast, mothur forces the user to specify each step of the pipeline.
+Although the former approach makes an analysis easier for a beginner, it potentially stifles users that need greater control or understanding of the assumptions at each step.
+This control over the pipeline has made it easier for researchers to customize databases or adapt the pipeline to analyze non-16S rRNA gene sequence data.
+Third, as I mentioned above, there has been a plethora of methods proposed for generating amplicon sequence data, and curating, aligning, checking for chimeras, classifying, and clustering the data.
+I am proud of the data-driven approach we have taken to comparing these methods.
+A description of a new method is of limited value if it is not benchmarked against other methods or control datasets.
+Through this core principle and mothur's large reach into the community, we have helped to develop standards in the analysis of 16S rRNA gene sequence data.
+Fourth, a focus on enabling reproducibility has always been central to the functionality of mothur.
+From the beginning, mothur's logfiles have represented a transcript of the user's command and outputs.
+When researchers were reluctant to submit sequence data to the Sequence Read Archive (SRA), we worked with the SRA developers to create a mothur command (make.sra) to create the package to submit sequence data through a special mothur portal.
+A more ambitious project had its seed on April 1, 2013 when we announced a new "function" in mothur: write.paper.
+The new command required that the user provide a 454 sff file and a journal title or impact factor.
+With this information, mothur would generate a manuscript.
+This April Fools' Day joke was poking fun at software that provided an analysis black box but also at many users' sentiments that data analysis should be so cut and dry.
+A few years later, we revisited this concept in the scope of reproducibility.
+Why not explicitly script an analysis from downloading data from the SRA through the rendering of a manuscript ready for submission?
+This idea gave rise to the development of the Riffomonas reproducible research tutorial series that enables researchers to write their own version of write.paper [@Schloss2018a].
+Perhaps the most important core principle is that my research group uses mothur to analyze the data we generate.
+We "eat our own dogfood".
+This has proven critical as it again represents transparency and hopefully provides confidence to mothur's users that we are not making recommendations that we do not follow ourselves.
 
@@
 One challenge we face is that if we looked to third parties to contribute code to mothur, they would need to know C++.
-Given the paucity of microbiologists with any programming skills, expecting that community to provide contributors that can write code in a syntax that prizes execution efficiency over developer efficiency was not likely.
-In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prizes developer efficiency over execution efficiency and exists as a series of wrappers to execute other developers' code.
-These choices resulted in many tradeoffs that have impacted ease of installation, usability, execution speed, and flexibility.
-If we were offered a grant to rewrite mothur, we would likely rewrite it as an R package that leaned heavily on the R language's C++ packages.
-Of course, such choices are always best in hindsight and when we started developing mothur, the ability to interface between scripting languages like R and Python and C++ code was not as well developed as it is today, For example, the modern version of the Rcpp package was first released in 2009 and its popularity was not immediate.
+Given the paucity of microbiologists with skills programing in a compiled language like C++, expecting that community to provide contributors that can write code in a syntax that prizes execution efficiency over developer efficiency was not likely.
+In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prizes developer efficiency over execution efficiency.
+QIIME is a series of wrappers that allow users to execute other developers' code making the use of a scripting language like Python attractive.
+Their choices resulted in many tradeoffs that have impacted ease of installation, usability, execution speed, and flexibility.
+If we were offered funding to rewrite mothur, we would likely rewrite it as an R package that leaned heavily on the R language's C++ interface packages.
+Of course, such choices are always best in hindsight.
+Yet, when we started developing mothur, the ability to interface between scripting languages like R and Python and C++ code was not as well developed as it is today.
+For example, the modern version of the Rcpp package was first released in 2009 and its popularity was not immediate [@Eddelbuettel2011].
 Again, the development of mothur has been a product of the environment that it was created in.
@@
 Beyond contributing to the mothur code base, we sought out other ways to include the community as developers.
-The paper describing mothur included **N** co-authors, all but three (Schloss, Ryabin, and Westcott) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
-Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for other seeking to adapt mothur for their uses.
-Again, this vision was a product of the environment.
-IPython notebooks (2011) and R markdown (2012) would not be developed until later.
+The paper describing mothur included 15 co-authors, all but three (Schloss, Westcott, and Ryabin) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
+Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for other seeking to adapt mothur for their uses (https://www.mothur.org/wiki).
 Unfortunately, once the incentive of co-authorship was removed, researchers stopped contributing their workflows to the wiki.
-Part of the difficulty of recruiting wiki contributors was a perception by some that the wiki was not a community resource.
-For example, I would frequently receive emails from people telling me that there was a typo on a specific page when the intention was that they could correct the typos without my input.
-We have been more successful in soliciting input and contributions from the user community through the mothur discussion forum and GitHub issue tracker.
-As mothur has matured, we have been dependent on the user community to use these resources to tell us what features they would like to see included in mothur.
-The user community also tells us where our documentation is confusing.
-Often we can count on people not directly affiliated with mothur to provide instruction and their own experience to other users.
+Again, this vision and the lack of the community's adoption of wikis as a mechanism for reporting workflows was a product of the environment.
+Although wikis were popular in the late 2000's, they lacked the ability to directly execute the commands that researchers reported.
+Such technology would not be possible until the creation of IPython notebooks (2011) and R markdown (2012).
+Another problem with the wiki approach was that potential contributors did not see the wiki as a community resource.
+I frequently received emails from scientists telling me that there was a typo on a specific page when the intention was that they could correct the typos without my input.
+We have been more successful in soliciting input and contributions from the user community through the mothur discussion forum and GitHub-based issue tracker.
+As mothur has matured, we have been dependent on the user community to use these resources to tell us what features they would like to see included in mothur and where the documentation is confusing or incomplete (https://forum.mothur.org).
+Often we can count on people not directly affiliated with mothur to provide instruction and their own experience to other users on the forum.
 We are constantly trying to recruit our "army" and are happy to take any contributions we can.
@@
 Over the past decade we have tried a number of experiments to improve the usability and utility of mothur.
-One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and venn diagrams depicting the overlap between microbial communities.
-I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote.
+One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and Venn diagrams depicting the overlap between microbial communities.
+Such visuals were helpful for exploring or data; however, I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote.
 Such visuals require far too much customization to be publication-quality.
-Although QIIME has incorporated visualization tools through the Emperor package, the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature.
-Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials.
-A second experiment we attempted was to create a graphical user interface (GUI) for running mothur.
+Although QIIME has incorporated visualization tools through the Emperor package [@VzquezBaeza2013], the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature.
+Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials that I have developed (http://www.riffomonas.org/minimalR/).
+A second experiment was the creation of a graphical user interface (GUI) for running mothur.
 Forcing users to interact with mothur through the command line has been a significant hurdle for many.
 Unfortunately, the development effort required to create and maintain a GUI is significant and there is limited funding for such efforts.
-The newest version of QIIME (version 2) has emphasized interaction with the tools through a GUI and it remains to be seen how this experiment will go.
+The newest version of QIIME (starting with version 2.0.0) has emphasized interaction with the tools through a GUI [@Bolyen2019] and the related QIITA project offers a web-based GUI [@Gonzalez2018].
+It remains to be seen how this experiment will go.
 Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks.
-This documentation is explicit in mothur as all commands and output is recorded in a logfile.
-Given the heightened recent focus on reproducibility we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project.
-Finally, we collaborated with programmers through Google Summer of Code to develop commands in mothur to implement the random forest and SVM machine learning algorithms.
-Similar to the challenges of developing attractive visuals, fitting the algorithms' hyperparameters, testing, and deploying the resulting models requires a significant amount of customization.
-Furthermore, this is an active area of research where methods are still being developed and improved.
-Thankfully, there are numerous R and Python packages that do a better job of developing these models.
+Documentation of commands and parameter values is explicit in mothur as users can provide the software a file with a list of commands and all commands and output are recorded in a logfile.
+Given the heightened focus on reproducibility in recent years [@Schloss2018b], we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project [@Schloss2018a].
+A final example of a failed experiment was a collaboration with programmers through Google Summer of Code to develop commands in mothur that ran the random forest and SVM machine learning algorithms.
+Similar to the challenges of developing attractive visuals, fitting the algorithms' hyperparameters, testing, and deploying the resulting models require a significant amount of customization.
+Furthermore, machine learning is an active area of research where methods are still being developed and improved.
+Thankfully, there are numerous R and Python packages that do a better job of developing these models [@Paszke2017; @Kuhn2008].
 Again, we have put our efforts into developing instructional materials that mothur users can use to fit such models to their data.
 In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.
-In hindsight, our solution to these failure has been to provide tutorials to as a conduit between mothur and their goals.
 
-***Competition is good and healthy.***
+***The future.*** I will continue to develop mothur for as long as other researchers find it useful.
+One challenge of such a plan is maintaining the funding to support its development.
+The development of mothur was initially enabled by a subcontract from a Sloan Foundation grant to Mitch Sogin to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
+We used that seed funding to secure an NSF grant and then a grant from NIH for tool development as part of their Human Microbiome Project.
+Since that project expired in 2013, we have not had funding to specifically support mothur's development.
+I have been fortunate to have start-up and discretionary funds generated from other projects to help support mothur.
+Although there is funding for new tools, there appears to be little appetite by funders to support existing tools.
+Emblematic of this was the NIH program, Big Data To Knowledge (BD2K), which solicited proposals through the program announcement "Extended Development, Hardening and Dissemination of Technologies in Biomedical Computing, Informatics and Big Data Science (PA-14-156)".
+This opportunity appeared perfect, except that the National Institute of Allergy and Infectious Diseases (NIAID), the primary supporter of microbiome research at NIH, did not participate in the announcement.
+Tools like mothur are clearly successful, but need funding mechanisms to continue to mature and support the needs of the research community.
 
-* A lot of people have been working in this space - e.g. clustering algorithms
-* Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))
+As with anything in science, methods become \text{pass\'e}.
+When we first developed mothur, T-RFLP and DGGE were still commonly used.
+Today it would be hard to argue that data from those methods meaningfully advance a study relative to what one could get using 16S rRNA gene sequence data.
+Looking forward, many want to claim that amplicon sequencing is today's DGGE.
+They claim that researchers should instead move on to shotgun metagenomic sequencing.
+It is important to note that the two methods answer fundamentally different questions.
+16S rRNA gene sequence data describes the taxonomic composition and metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community.
+Both tools provide important information but they cannot easily replace each other.
+Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is at least an order of magnitude higher than that of amplicon data.
+For example, we analyzed 10,000 16S rRNA sequences from each of about 500 subjects [@Baxter2016].
+We can think of this as representing about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome).
+Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4 Tbp.
+Although such a sequencing effort is technically possible, the cost of such an endeavor would be considerable and unlikely to be pursued by most researchers.
+We estimate that generating and sequencing the libraries at the University of Michigan sequencing core would cost approximately $150 per library.
+The parallel 16S rRNA gene sequences data would cost approximately $8 per library.
+Furthermore, analyzing such a large dataset with an approach that captures the full genetic diversity of the community would be financially and technically prohibitive.
+Going forward, there is still a place for 16S rRNA gene sequencing.
+Although sequencing technologies will evolve to capture longer and more high quality data, there will likely always be a need for characterizing the taxonomic diversity of microbial communities.
+With this in mind, there will always be a place for tools like mothur that can analyze amplicon sequence data.
 
-What are we proud of?
-* Open source packages
-* Platform independent
-* Development of open instructional materials
-* Data driven development of methods
-* Database independent - allows people outside of our field to use mothur
-* Reproducibility
-* Helped create standards - when something is this popular, you force people to use a certain suite of tools and methods
+Of course this does not mean that such tools will remain static.
+We see three key areas that we will continue to help the field to move forward.
+First, just as we adapted through the transitions from Sanger to 454 to MiSeq and PacBio sequencing platforms [@Kozich2013; @Schloss2011b; @Schloss2016b], we must learn whether data from Oxford Nanopore can be an alternative sequencing approach that generates sequence data that is the same quality as existing approaches; thus far, the approach has significant shortcomings for sequencing 16S rRNA gene sequences [@Calus2018].
+As with the earlier platforms, we must better understand its error profile so that sequencing errors can be corrected.
+We have learned that moving forward requires that we maintain or improve sequence quality.
+No doubt, datasets and read lengths will improve, but these advances should not be made at the cost of data quality.
+Second, with these improvements, we will need to continue to improve our algorithms.
+We have already seen that attempts to use low quality MiSeq and HiSeq data causes computational problems leading to the creation of open and closed reference clustering methods [@NavasMolina2013; @Rideout2014].
+Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods [@Westcott2015].
+More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality.
+Finally, there are ongoing controversies that need further exploration.
+These include the validity and utility of amplicon sequence variants [@Callahan2017], the wisdom of removing low frequency sequences [@Bokulich2012], and methods of identifying and removing contaminant 16S rRNA gene sequences [@Salter2014; @Davis2018].
+With each of these areas of development, the broader community can count on our same data-driven approach to meeting evaluating these questions.
+It is common for researchers to comment that they pick a specific method or deviate from a suggestion because they "like how the data look".
+When pressed for an objective definition of how they know the data look "right", they go quiet.
+Through the use of mock communities and simulations where we actually know what looks right and objective metrics of quality like the MCC or sequencing error rates, we will continue to base recommendations on data rather than a gut feeling.
 
-The future
-* 16S rRNA gene sequencing is not dead
-* Always improving sequencing methods bring new problems
-* Larger datasets, need a replacement for naive open and closed reference clustering
-* Lingering controversies - singletons, ASVs, contamination
-
-Longterm funding
-* The development of mothur was enabled by initial funding from a grant to Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
-* NSF
-* NIH
-* Bootstrapping from other projects
-
-Red Queen
-
-\newpage
+***Conclusion.*** In the paper announcing mothur, we commented that the relationship between 16S rRNA gene sequencing and analysis is very much like the Red Queen in Lewis Carroll's book, *Through the Looking-Glass*.
+Although some disagreed with this analogy [@Caporaso2010b], I still feel it is apt.
+The sequencing technology and rapacious appetite of researchers continues to race on.
+At the same time, bioinformatics tools must adapt to facilitate our research.
+I am confident that mothur will be up to this exciting challenge.
+Beyond its utility for analyzing amplicon sequence data, mothur's history provides lessons that are helpful for other projects that hope to develop a long historical arc.
+First, mothur is a product of its time.
+We have always sought to solve a current need to the best of our ability with the tools we had at the time.
+There are certainly caveats to any analysis of 16S rRNA gene sequence data, but if we had waited until those caveats were resolved, the field never would have progressed.
+Similarly, we made design choices that we probably would not have made had we started the project today.
+Second, as we have developed mothur, we have attempted to do so in a data-driven approach where we compare multiple methods.
+It has not merely been enough to propose a new method, we must show that it meaningfully advances the field.
+Third, through our failures and successes we have learned to focus on what mothur is good at and create products separate from mothur when distinct needs arise.
+For example, we have learned that mothur should not have a graphical interface or data visualization tools.
+Instead, we will provide instructional materials to teach users how to use the command line interface and other programs like R for data visualization.
+Finally, mothur was born out of a need for automating the analysis of large 16S rRNA gene sequence datasets.
+I realized I had a set of skills to fill that need.
+It has been refreshing to see the computational skills of the microbial ecology field grow over the past two decades.
+Looking ahead, we must all take stock of the challenges we face in microbial ecology and how our individual skills and interests can address these challenges to turn them into opportunities.
 
 ## Acknowledgements
+The development of mothur would not be possible without the contributions of its many supporters, developers, and users.
+Although not a complete list, I would be remiss if I did not express my gratitude to Sarah Westcott and the other members of my research group, Jo Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra Mass, and Marcy Balunas for their unique contributions to the continued development of mothur.
 
@@
 ## References
-25 references
 
@@
 **Figure 1. Caption caption caption.** Footnotes footnotes footnotes.
+
+* Timeline of mothur/QIIME/etc citations
+* Timeline of SRA deposits / 16S/microbiome in PubMed
+
+**Figure 2. Caption caption caption.** Footnotes footnotes footnotes.
+
+* Screenshot of mothur welcoming page
+
+**Figure 3. Caption caption caption.** Footnotes footnotes footnotes.
+
+* Screenshot of mothur homepage

### 2019-09-30 submission/manuscript.tex
Editing
Edited manuscript for wording, structure, and to add references.

--- 2019-09-16 submission/manuscript.tex
@@
 \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+
+%\usepackage{mathpazo} % Palatino font
+%\renewcommand*\familydefault{\rmdefault} % Use the roman version of the font
+
 \usepackage[T1]{fontenc}
@@
 
-\hypertarget{observation-format}{%
-\subsection{Observation format}\label{observation-format}}
+\hypertarget{mini-review}{%
+\subsubsection{Mini-review}\label{mini-review}}
 
@@
 
-75 words
+250 words
 
@@
 
-3000 words total
+Few scientists set out on a nearly two decade-long journey with a
+specific goal in mind. Often we fail to start a scientific journey
+because it looks too hard. Perhaps we get discouraged by all of the
+things that could go wrong. Maybe we stray from the path because we find
+something that is more interesting. Every scientist picks their own path
+and takes their own forks in the road. From the outside, it may appear
+to be a random walk. Nevertheless, these meanderings are common in
+science.
 
-Few scientists set out on a nearly two decade long journey with a
-specific goal in mind. Often we fail to start a scientific journey
-because it looks too hard. Perhaps we get bogged down in all of the
-things that could go wrong. Perhaps we go astray from the path because
-we find something else that appears more interesting. Every scientist
-picks their own path and takes their own forks in the road. From the
-outside, it may appear to be a random walk. Nevertheless, these
-meandering journeys in are common in science.
-
-At the risk of navel gazing, looking back on our scientific journeys can
-be instructive to other scientists who are overwhelmed at the prospect
-of looking forward at their careers {[}Lenski and Chemistry guy{]}. By
-no means is my personal journey over, but since 2002 I have been on a
+Looking back on scientific journeys can be instructive to others who are
+overwhelmed at the prospect of looking forward at their careers (1--5).
+By no means is my personal journey over, but since 2002 I have been on a
 journey that I did not realize I was on. Now that the paper introducing
 the mothur software package is ten years old and has become the most
-cited paper published by \emph{Applied and Environmental Microbiology},
-it is worth stepping back and using the development of mothur as a story
-that likely has parallels to many other research stories that have taken
-time to develop.
+cited paper published by \emph{Applied and Environmental Microbiology}
+(7), it is worth stepping back and using the continued development of
+mothur as a story that has parallels to many other research stories.
 
@@
 Rarefaction curves and attempts to estimate the taxonomic richness of
-soil had become popular because of the simple, but impactful mini-review
-by Jennifer Hughes and her colleagues, which introduced the field to
-operational taxonomic units (OTUs), rarefaction curves, and richness
-estimates {[}DOI: 10.1128/AEM.67.10.4399-4406.2001{]}. I do not recall
-whether that poster had a rarefaction curve on it, but her question
-primed my career.
+soil had become popular because of the impactful review by Jennifer
+Hughes and her colleagues (8). Their seminal paper introduced the field
+to operational taxonomic units (OTUs), rarefaction curves, and richness
+estimates. I do not recall whether my poster had a rarefaction curve on
+it, but Jo's question primed my career.
 
@@
 trivial. How would I bin the sequences into OTUs? Hughes and her
-colleagues did it manually for datasets that had fewer than 284
-sequences. Although I could possibly do that for my 600 sequences, my
-goal was to generate 1,000 sequences from the sample and to repeat that
-sampling effort for other samples. I needed something that could be
-automated. Furthermore, the software that Hughes used, EstimateS,
-required a series of tedious data formatting steps. I had found my first
-problem. How would I assign sequences to OTUs and use that data to
-estimate the richness and diversity of a sample? The second problem
-would be how could I compare the sequences found in one sample to
-another sample? The solution to the first problem, DOTUR (Distance-based
-OTUs and Richness), took us two years to develop
-{[}10.1128/AEM.71.3.1501-1506.2005{]}. DOTUR did two things: given a
-matrix describing the genetic distance between pairs of sequences, it
-would cluster those sequences into OTUs for any distance threshold to
-define the OTUs and then it would use the frequency of each OTU to
-calculate a variety of alpha diversity metrics. The solutions to the
-second problem would come from our work to develop software including
-S-LIBSHUFF {[}2004{]}, SONS (Shared OTUs and Similarity) {[}2006{]}, and
-TreeClimber {[}2006{]}. Around the same time, Catherine Lozupone and Rob
-Knight were developing their UniFrac tools to compare communities with a
-phylogenetic rather than OTU-based approach {[}PMID: 16332807; PMID:
-17220268{]}. With these tools, the field of microbial ecology had a
-quantitative toolbox for describing and comparing microbial communities.
+colleagues did it manually and with fewer than 300 sequences. Although I
+could possibly do that for my 600 sequences, my goal was to generate
+1,000 sequences from the sample and to repeat that sampling effort for
+other samples. I needed something that could be automated. Furthermore,
+the software that Hughes used, EstimateS (4), required a series of
+tedious data formatting steps to perform the analyses we were interested
+in performing. I had found my first problem. How would I assign
+sequences to OTUs and use that data to estimate the richness and
+diversity of a sample? The second problem would involve comparing the
+abundance of OTUs found in one sample to another sample. The solution to
+the first problem, DOTUR (Distance-based OTUs and Richness), took us two
+years to develop (9). DOTUR did two things: given a matrix quantifying
+the genetic distance between pairs of sequences, it would cluster those
+sequences into OTUs for any distance threshold to define the OTUs and
+then it would use the frequency of each OTU to calculate a variety of
+alpha diversity metrics. The solutions to the second problem would come
+from our work to develop software including \(\int\)-LIBSHUFF (10), SONS
+(Shared OTUs and Similarity) (11), and TreeClimber (12). Around the same
+time, Catherine Lozupone and Rob Knight were developing their UniFrac
+tools to compare communities with a phylogenetic rather than OTU-based
+approach (13, 14). With these tools, the field of microbial ecology had
+a quantitative toolbox for describing and comparing microbial
+communities. Along the way Jo and I would demonstrate the utility of
+such tools to answer questions like how many OTUs were there in that
+sample of Alaskan soil and how many sequences were needed to sample each
+of those OTUs (15)? Where were we in the global bacterial census (16)?
+How does the word usage of \emph{Goodnight, Moon} compare to that of
+\emph{Portrait of a Lady} and more importantly how is this relevant to
+microbial ecology (17)? Most edifying were the more than 2,400 papers
+that used DOTUR, SONS, TreeClimber, or \(\int\)-LIBSHUFF to facilitate
+their own research questions (Web of Science, 10/1/2019). Had we waited
+to solve all of the problems that plagued 16S rRNA gene sequencing, we
+would still be waiting.
 
@@
 16S rRNA gene sequencing. We knew there were biases from extractions and
-amplification {[}XXXX{]}. We knew there were chimeras {[}XXXX{]}.
-Getting to the distance matrix required trimming and correcting sequence
-errors, aligning them, and finding the most appropriate way to calculate
-a pairwise distance. It was also rare to have experimental replication
-to perform statistical tests to compare treatment groups. We frequently
-used a dataset comparing Scottish soils from from Alison McCaig and
-colleagues. This dataset consisted of two experimental groups, each
-replicated three times with 45 sequences in each replicate.
-Nevertheless, we had excuses and work arounds for these problems that
-served our needs. At the time, I felt that the biggest problems were how
-to cluster the sequences into OTUs and how to use those clusterings to
-test our hypotheses. Along the way we would demonstrate the utility of
-such tools to answer questions like where are we in the bacterial
-census? How many sequences would it take to see every OTU in that sample
-of Alaskan soil? How does the word usage of \emph{Goodnight, Moon}
-compare to that of \emph{Portrait of a Lady}? More importantly, 1,900
-papers used DOTUR to facilitate their own research questions. Had we
-waited to solve all of the problems that plague 16S rRNA gene
-sequencing, we would still be waiting.
+amplification (18--23). We knew there were chimeras (24--27). We knew
+that bacteria varied in their \emph{rrn} copy number. Generating a
+distance matrix was a prerequisite to using my tools. This wasn't
+trivial, but by cobbling together other tools it was possible. We would
+assemble, trim and and correct Sanger sequence reads using Chromas or
+STADEN (28), align the sequences using ClustalW (29) or ARB (30), check
+for chimeras using partial treeing or Bellerephon (27), and calculate a
+pairwise distance matrix using DNADIST from the PHYLIP package (31). At
+the time, we knew that we only had a loose concept of a species based on
+these distances (32). We hoped that an OTU defined as a group of
+sequences more than 97\% similar to each other would be a biologically
+meaningful unit regardless of whether it fit our notion of a bacterial
+species. At the time, I felt that the biggest problems that I could
+solve were how to cluster the sequences into OTUs and how to use those
+clusterings to test our hypotheses. The only tool available at the time
+that automated the clustering step was FastGroup, which implemented a
+greedy version of the single linkage algorithm (33). The high cost of
+sequencing was also an impediment to experimentation and analysis in
+microbial ecology. It was rare for a study design to have experimental
+replicates so that one could perform a statistical test to compare
+treatment groups. For example, in our testing we frequently used a
+dataset comparing Scottish soils from from Alison McCaig and colleagues
+(34). This dataset consisted of two experimental groups, each replicated
+three times with 45 sequences per replicate. Although great focus has
+been placed on the depth of sampling afforded by 454 and Illumina
+sequencing, the true benefit of the modern sequencing platforms is the
+ability to affordably sequence a large number of technical and
+biological replicates. In spite of the many technical challenges, we had
+excuses and heuristics to solve problems that served our needs. It is
+telling that a recent review of ``best practices'' in generating and
+analyzing 16S rRNA gene sequences shows that we still have not solved
+many of these issues and that we have even identified additional
+problems (35).
 
-As we developed these tools, I found a unique niche in microbiology. I
-believe that my undergraduate and graduate training as a biological
-engineer prepared me to think about research questions from a systems
-perspective, to think quantitatively, and to understand the value of
-using computer programs to help solve problems. As an undergraduate
-engineering student, I learned the Pascal programming language and
-promptly forgot much of it as an engineering graduate student. As a
-postdoc, I learned the Perl programming language to better understand
-how LIBSHUFF, a tool for comparing the structure of two communities,
-worked since it was written in Perl {[}DOI:
-10.1128/AEM.70.9.5485-5492.2004{]}. After writing my own version of
-LIBSHUFF and seeing the speed of the version written in C++ by my
-collaborator, Bret Larget, I converted my Perl version of DOTUR into
+As we developed these tools, I found a unique niche in microbiology. My
+undergraduate and graduate training as a biological engineer prepared me
+to think about research questions from a systems perspective, to think
+quantitatively, and to understand the value of using computer programs
+to help solve problems. As an undergraduate student, I learned the
+Pascal programming language and promptly forgot much of it as a graduate
+student and learned MatLab in its place. As a postdoc, I learned the
+Perl programming language to better understand how LIBSHUFF, a tool for
+comparing the structure of two communities, worked since it was written
+in Perl (36). After writing my own version of LIBSHUFF,
+\(\int\)-LIBSHUFF and seeing the speed of the version written in C++ by
+my collaborator, Bret Larget, I converted my Perl version of DOTUR into
 C++. At the time, the conversion from Perl to C++ seemed like an
@@
 languages to solve problems was enabled by my prior training. It was
-also a skill that was virtually unheard of in microbiology. Today
-researchers without the ability to program in Python or R are at a
-significant disadvantage.
+also a skill that was virtually unheard of in microbiology. Today,
+researchers without the ability to program are at a significant
+disadvantage (37).
 
 \textbf{\emph{Introducing mothur.}} Shortly after DOTUR was published, I
-received an email from Mitch Sogin asking whether DOTUR could handle
-more than a million sequences. Without answering his question, I asked
-where he found a million sequences. Little did I know that his email
-would represent another pivot in the development of these tools. His
-group would be the first to use 454 sequencing technology to generate
-16S rRNA gene sequences {[}PMID: 16880384{]}. Although mothur could
-assign those sequences to OTUs, it was slow and required a significant
-amount of RAM. As I left my postdoc to start my independent career
-across the state from Sogin's lab at the University of Massachusetts in
-Amherst, my plan was to rewrite DOTUR, SONS, S-LIBSHUFF, and TreeClimber
-for the new world of massively parallelized sequencing. The new tool was
-mothur.
+received an email from Mitch Sogin, a scientist at the Marine Biology
+Laboratory (Woods Hole, MA), who asked whether DOTUR could handle more
+than a million sequences. Without answering his question, I asked where
+he found a million sequences. Little did I know that his email would
+represent another pivot in the development of these tools. His group
+would be the first to use 454 sequencing technology to generate 16S rRNA
+gene sequences (38). Although DOTUR could assign those sequences to
+OTUs, at the scale of millions of sequences, it was slow and required a
+significant amount of RAM. As I left my postdoc to start my independent
+career across the state from Sogin's lab at the University of
+Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS,
+\(\int\)-LIBSHUFF, and TreeClimber for the new world of massively
+parallelized sequencing. The new tool would become mothur.
 
 Milling about at a poster session at an ASM General Meeting in New
-Orleans, I again ran into Mitch who asked what my plans were for new
-tools. I told him that I wanted to make a tool like ARB (a powerful
-database tool and phylogenetics package), but for microbial ecology
-analysis. His retort was, ``You and what army?'' To that point, I had
-written every line of code and been answering many emails from people
-asking for help. It would be difficulty, but I needed to learn to let go
-and share the development process with someone else. He was right, I
-would need an army. That ``army'' ended up being Sarah Westcott who has
-worked on the mothur project largely from its inception. Today, mothur
-is over 200,000 lines of code and Sarah has touched or written nearly
-every line of code. Beyond writing and testing mothur's code base, she
-has become a conduit for many learning the tools of microbial ecology by
-patiently answering questions via email and the package's discussion
-forum. The community and I are lucky that Sarah has stayed with the
-project for more than a decade. To be honest, such dependency on a
-single person makes the project brittle. In hindsight, it would have
-been better to have developed mothur with more of an ``army'' or team so
-that there is overlap in people's understanding of how mothur works.
-Although such an arrangement might work in a software engineering firm,
-it is not practical in an academic setting where funding is limited for
-developers of free, open source software packages. There are certainly
-projects that make this work, but they are rare.
+Orleans, I ran into Mitch who asked what my plans were for my new lab. I
+told him that I wanted to make a tool like ARB, a powerful database tool
+and phylogenetics package (30), but for microbial ecology analysis. His
+retort was, ``You and what army?'' Up to that point, I had written every
+line of code and been answering many emails from people asking for help.
+He was right, I would need an army. It would be difficult, but I needed
+to learn to let go and share the development process with someone else.
+My ``army'' ended up being Sarah Westcott who has worked on the mothur
+project from its inception. Today, mothur is over 200,000 lines of code
+and Sarah has touched or written nearly every line of it. Beyond writing
+and testing mothur's code base, she has become a conduit for many who
+are trying to learn the tools of microbial ecology. She patiently
+answers questions via email and on the package's discussion forum
+(\url{https://forum.mothur.org}). The community and I are lucky that
+Sarah has stayed with the project for more than a decade. To be honest,
+such dependency on a single person makes the project brittle. In
+hindsight, it would have been better to have developed mothur with more
+of an ``army'' or team so that there is overlap in people's
+understanding of how mothur works. Although a distributed team approach
+might work in a software engineering firm, it is not practical in most
+academic environments where there is limited funding. There are
+certainly projects that make this work, but they are rare.
+
+\textbf{\emph{Competition has been good and healthy.}} mothur has not
+been developed in a vacuum and it does not have a monopoly within the
+field. As indicated above, each of our decisions were made in the
+historical context of the field and with constant pressure from others
+developing their own tools for analyzing 16S rRNA gene sequence data.
+Competition has been good for mothur and for the field.
+
+From the beginning there have been online tools available at the
+Ribosomal Database Project (RDP) (39), greengenes (40), and SILVA (41).
+These allowed users a straightforward method of comparing their data to
+those collected in a database. There are two primary downsides to these
+tools. First, researchers running the online tool must pay the
+computational expenses leading to slow process times when hardware
+becomes outdated and when numerous users simultaneously attempt their
+analysis. Eventually this limitation would result in the termination of
+the greengenes website. Second, these platforms provide a
+one-size-fits-all analysis. These tools only allow a user to analyze 16S
+and in some cases 18S rRNA gene sequences. If a user sequences a
+different gene, then the tool will not serve them. These observations
+resulted in two design goals we have had with mothur: bringing the
+analysis to a user's computer and separating a tool from a specific
+database. For example, we commonly use a sequence alignment method that
+was originally developed for greengenes (42), but use a SILVA-based
+reference alignment because its superior quality (43, 44). In addition,
+we offer the \text{na\"ive} Bayesian classifier developed by the RDP
+(45) and allow users to train it to any database they want, including
+customized databases. In both examples, users can align or classify
+non-rRNA gene sequence data. As the bioinformatics tools have matured,
+both the RDP and SILVA offer integrated pipelines for analyzing large
+datasets, albeit in one-size-fits-all black box implementations.
+
+With the growth in popularity of 16S rRNA gene sequencing there has
+naturally been an expansion in the number of people developing tools to
+analyze these data. Months after the paper describing mothur was
+published, the paper describing QIIME was published (46). Over the past
+10 years, many have attempted to create analogies comparing the two
+programs: Pepsi vs Coke, Apple vs Windows, etc. It is never clear which
+software is which brand and whether the comparisons are meant as a
+complement or an insult. Regardless, both programs are very popular.
+From my perspective, most of the differences are cosmetic. QIIME is
+effectively a bundle of scripts to run other developers' software. For
+example, with QIIME (through version 1.9.1), it was even possible to run
+mothur through QIIME. One can also run the \text{na\"ive} Bayesian
+classifier through QIIME using the original code developed by the RDP.
+This caused great frustration for many users because there were numerous
+software dependencies that had to be installed. Although the QIIME
+developers would go on to create virtual machines and use packaging
+tools to simplify installation, these fixes required sophistication by
+users who we knew struggled with the basics of navigating a command
+line. In contrast, when a user runs mothur, they are running mothur. The
+\text{na\"ive} Bayesian classifier code that is in mothur is a rewritten
+version of the original code. When we rewrite someone's software we do
+it with an eye to improving performance, access, and utility for non-16S
+rRNA gene sequence data. For example, while 454 data was popular,
+PyroNoise was an effective tool for denoising flowgram data (47).
+Running the original code required a large Linux computer cluster and
+knowledge of bash and Perl scripting. When we rewrote the code for
+mothur, we made it accessible to people using any operating system with
+a simple command interface (i.e.~trim.flows and shhh.flows). Our
+approach requires significant developer effort, but saves considerable
+user effort. As this benefit is multiplied across thousands of projects,
+the savings to users has been considerable.
+
+Beyond the large packages like mothur and QIIME, there has been
+significant growth in stand alone software tools for sequence curation
+(e.g.~PyroNoise (47), PANDAseq (48), DADA2 (49)), chimera checking
+(e.g.~UCHIME (50), ChimeraSlayer (51), Perseus (52)), and clustering
+(e.g.~USEARCH (53), VSEARCH (54), Swarm (55)). Where possible and when
+warrented, we have implemented many of these algorithms directly into
+mothur. We have also used this diversity of methods to perform
+head-to-head comparisons. Most notable is the area of clustering
+algorithms where there have been a large number of algorithms developed
+without an obvious method to objectively compare them (56--59). We
+applied an objective metric, the Matthew's Correlation Coefficient
+(MCC), to evaluate numerous algorithms for clustering sequences into
+OTUs. By performing this type of analysis, we were able to objectively
+compare the algorithms, make recommendations to the field, and develop
+new algorithms that outperformed the existing algorithms. Beyond
+evaluating clustering algorithms, we have also evaluated methods of
+denoising sequence data (60--62), assessed reference alignments (43,
+44), considered the importance of incorporating secondary structure
+information in alignments (63), quantified the variation along the 16S
+rRNA gene (44), and compared the statistical hypotheses tested by
+commonly used tools (64). We have embraced the competition and diversity
+of all methods being used to analyze amplicon data as an opportunity to
+identify the strengths and weaknesses of the methods in an attempt to
+make recommendations to other researchers.
+
+\textbf{\emph{mothur's core principles.}} As mothur has evolved with the
+needs of the community, several core principles have emerged that direct
+its development. First, mothur is a free, open source software package.
+This has been critical in shaping the direction of mothur. We were
+content for mothur to be an improved combination of DOTUR and SONS and
+leverage existing tools for other steps. Yet, when we learned that the
+code for NAST, the algorithm behind the greengenes's aligner (42), was
+not open source or publicly available. Similarly, although the SINA
+aligner was available through ARB and the SILVA website performed well
+it was closed source. Because the ARB implementation did not scale to
+large datasets, researchers were left to pay a processing fee to the
+SILVA website to align sequences. Thus, we realized that such an
+important tool needed to be opened to the community (43). More recently,
+the rejection of closed source, commercial tools such as USEARCH can be
+seen by the broader adoption of its open source, free competitor,
+VSEARCH, within the microbial ecology community (53, 54). Related to
+insuring that mothur's code is open source, our second core principle is
+that we maintain transparency to our users. Perhaps a user does not need
+to interrogate every line of code, but they need to understand what is
+happening. Many programs including online workflows encapsulate large
+elements of a pipeline in a single command. In contrast, mothur forces
+the user to specify each step of the pipeline. Although the former
+approach makes an analysis easier for a beginner, it potentially stifles
+users that need greater control or understanding of the assumptions at
+each step. This control over the pipeline has made it easier for
+researchers to customize databases or adapt the pipeline to analyze
+non-16S rRNA gene sequence data. Third, as I mentioned above, there has
+been a plethora of methods proposed for generating amplicon sequence
+data, and curating, aligning, checking for chimeras, classifying, and
+clustering the data. I am proud of the data-driven approach we have
+taken to comparing these methods. A description of a new method is of
+limited value if it is not benchmarked against other methods or control
+datasets. Through this core principle and mothur's large reach into the
+community, we have helped to develop standards in the analysis of 16S
+rRNA gene sequence data. Fourth, a focus on enabling reproducibility has
+always been central to the functionality of mothur. From the beginning,
+mothur's logfiles have represented a transcript of the user's command
+and outputs. When researchers were reluctant to submit sequence data to
+the Sequence Read Archive (SRA), we worked with the SRA developers to
+create a mothur command (make.sra) to create the package to submit
+sequence data through a special mothur portal. A more ambitious project
+had its seed on April 1, 2013 when we announced a new ``function'' in
+mothur: write.paper. The new command required that the user provide a
+454 sff file and a journal title or impact factor. With this
+information, mothur would generate a manuscript. This April Fools' Day
+joke was poking fun at software that provided an analysis black box but
+also at many users' sentiments that data analysis should be so cut and
+dry. A few years later, we revisited this concept in the scope of
+reproducibility. Why not explicitly script an analysis from downloading
+data from the SRA through the rendering of a manuscript ready for
+submission? This idea gave rise to the development of the Riffomonas
+reproducible research tutorial series that enables researchers to write
+their own version of write.paper (65). Perhaps the most important core
+principle is that my research group uses mothur to analyze the data we
+generate. We ``eat our own dogfood''. This has proven critical as it
+again represents transparency and hopefully provides confidence to
+mothur's users that we are not making recommendations that we do not
+follow ourselves.
 
@@
 to contribute code to mothur, they would need to know C++. Given the
-paucity of microbiologists with any programming skills, expecting that
-community to provide contributors that can write code in a syntax that
-prizes execution efficiency over developer efficiency was not likely. In
-contrast, the QIIME development team could be more distributed because
-their code base was primarily written in Python, which prizes developer
-efficiency over execution efficiency and exists as a series of wrappers
-to execute other developers' code. These choices resulted in many
-tradeoffs that have impacted ease of installation, usability, execution
-speed, and flexibility. If we were offered a grant to rewrite mothur, we
-would likely rewrite it as an R package that leaned heavily on the R
-language's C++ packages. Of course, such choices are always best in
-hindsight and when we started developing mothur, the ability to
-interface between scripting languages like R and Python and C++ code was
-not as well developed as it is today, For example, the modern version of
-the Rcpp package was first released in 2009 and its popularity was not
-immediate. Again, the development of mothur has been a product of the
+paucity of microbiologists with skills programing in a compiled language
+like C++, expecting that community to provide contributors that can
+write code in a syntax that prizes execution efficiency over developer
+efficiency was not likely. In contrast, the QIIME development team could
+be more distributed because their code base was primarily written in
+Python, which prizes developer efficiency over execution efficiency.
+QIIME is a series of wrappers that allow users to execute other
+developers' code making the use of a scripting language like Python
+attractive. Their choices resulted in many tradeoffs that have impacted
+ease of installation, usability, execution speed, and flexibility. If we
+were offered funding to rewrite mothur, we would likely rewrite it as an
+R package that leaned heavily on the R language's C++ interface
+packages. Of course, such choices are always best in hindsight. Yet,
+when we started developing mothur, the ability to interface between
+scripting languages like R and Python and C++ code was not as well
+developed as it is today. For example, the modern version of the Rcpp
+package was first released in 2009 and its popularity was not immediate
+(66). Again, the development of mothur has been a product of the
 environment that it was created in. Although these decisions have
@@
 include the community as developers. The paper describing mothur
-included \textbf{N} co-authors, all but three (Schloss, Ryabin, and
-Westcott) responded to a call to provide a wiki page that described how
-they used an early version of mothur to analyze a data set. Our vision
-was that authors might use the mothur wiki to document reproducible
-workflows for papers using mothur but to also provide instructional
-materials for other seeking to adapt mothur for their uses. Again, this
-vision was a product of the environment. IPython notebooks (2011) and R
-markdown (2012) would not be developed until later. Unfortunately, once
-the incentive of co-authorship was removed, researchers stopped
-contributing their workflows to the wiki. Part of the difficulty of
-recruiting wiki contributors was a perception by some that the wiki was
-not a community resource. For example, I would frequently receive emails
-from people telling me that there was a typo on a specific page when the
+included 15 co-authors, all but three (Schloss, Westcott, and Ryabin)
+responded to a call to provide a wiki page that described how they used
+an early version of mothur to analyze a data set. Our vision was that
+authors might use the mothur wiki to document reproducible workflows for
+papers using mothur but to also provide instructional materials for
+other seeking to adapt mothur for their uses
+(\url{https://www.mothur.org/wiki}). Unfortunately, once the incentive
+of co-authorship was removed, researchers stopped contributing their
+workflows to the wiki. Again, this vision and the lack of the
+community's adoption of wikis as a mechanism for reporting workflows was
+a product of the environment. Although wikis were popular in the late
+2000's, they lacked the ability to directly execute the commands that
+researchers reported. Such technology would not be possible until the
+creation of IPython notebooks (2011) and R markdown (2012). Another
+problem with the wiki approach was that potential contributors did not
+see the wiki as a community resource. I frequently received emails from
+scientists telling me that there was a typo on a specific page when the
 intention was that they could correct the typos without my input. We
 have been more successful in soliciting input and contributions from the
-user community through the mothur discussion forum and GitHub issue
-tracker. As mothur has matured, we have been dependent on the user
+user community through the mothur discussion forum and GitHub-based
+issue tracker. As mothur has matured, we have been dependent on the user
 community to use these resources to tell us what features they would
-like to see included in mothur. The user community also tells us where
-our documentation is confusing. Often we can count on people not
-directly affiliated with mothur to provide instruction and their own
-experience to other users. We are constantly trying to recruit our
-``army'' and are happy to take any contributions we can. Whether the
-contributions are to the code base, discussion forum, or suggestions for
-new tools, these contributions have been invaluable to the growth and
-popularity of mothur.
+like to see included in mothur and where the documentation is confusing
+or incomplete (\url{https://forum.mothur.org}). Often we can count on
+people not directly affiliated with mothur to provide instruction and
+their own experience to other users on the forum. We are constantly
+trying to recruit our ``army'' and are happy to take any contributions
+we can. Whether the contributions are to the code base, discussion
+forum, or suggestions for new tools, these contributions have been
+invaluable to the growth and popularity of mothur.
 
@@
 first experiments was to use mothur to generate standard vector graphic
-(SVG)-formatted files of heatmaps and venn diagrams depicting the
-overlap between microbial communities. I quickly realized that I would
-never put a mothur-generated figure into a manuscript I wrote. Such
-visuals require far too much customization to be publication-quality.
-Although QIIME has incorporated visualization tools through the Emperor
-package, the challenge of users taking default values has downsides as
-ordinations with black background or publishing 3-D ordinations in a 2-D
-medium litter the literature. Instead, we have encouraged users to use R
+(SVG)-formatted files of heatmaps and Venn diagrams depicting the
+overlap between microbial communities. Such visuals were helpful for
+exploring or data; however, I quickly realized that I would never put a
+mothur-generated figure into a manuscript I wrote. Such visuals require
+far too much customization to be publication-quality. Although QIIME has
+incorporated visualization tools through the Emperor package (67), the
+challenge of users taking default values has downsides as ordinations
+with black background or publishing 3-D ordinations in a 2-D medium
+litter the literature. Instead, we have encouraged users to use R
 packages to visualize mothur-generated results using the minimalR
-instructional materials. A second experiment we attempted was to create
-a graphical user interface (GUI) for running mothur. Forcing users to
-interact with mothur through the command line has been a significant
-hurdle for many. Unfortunately, the development effort required to
-create and maintain a GUI is significant and there is limited funding
-for such efforts. The newest version of QIIME (version 2) has emphasized
-interaction with the tools through a GUI and it remains to be seen how
-this experiment will go. Another downside of using a GUI is that there
-is a risk that reproducibility will suffer if users do not have a
-mechanism to document their mouse clicks. This documentation is explicit
-in mothur as all commands and output is recorded in a logfile. Given the
-heightened recent focus on reproducibility we have extended significant
-effort in developing instructional materials teaching users how to
-organize, document, and execute reproducible pipelines that allow a user
-to go from raw sequence data to a compiled manuscript with figures
-through the Riffomonas project. Finally, we collaborated with
-programmers through Google Summer of Code to develop commands in mothur
-to implement the random forest and SVM machine learning algorithms.
-Similar to the challenges of developing attractive visuals, fitting the
-algorithms' hyperparameters, testing, and deploying the resulting models
-requires a significant amount of customization. Furthermore, this is an
-active area of research where methods are still being developed and
-improved. Thankfully, there are numerous R and Python packages that do a
-better job of developing these models. Again, we have put our efforts
+instructional materials that I have developed
+(\url{http://www.riffomonas.org/minimalR/}). A second experiment was the
+creation of a graphical user interface (GUI) for running mothur. Forcing
+users to interact with mothur through the command line has been a
+significant hurdle for many. Unfortunately, the development effort
+required to create and maintain a GUI is significant and there is
+limited funding for such efforts. The newest version of QIIME (starting
+with version 2.0.0) has emphasized interaction with the tools through a
+GUI (68) and the related QIITA project offers a web-based GUI (69). It
+remains to be seen how this experiment will go. Another downside of
+using a GUI is that there is a risk that reproducibility will suffer if
+users do not have a mechanism to document their mouse clicks.
+Documentation of commands and parameter values is explicit in mothur as
+users can provide the software a file with a list of commands and all
+commands and output are recorded in a logfile. Given the heightened
+focus on reproducibility in recent years (70), we have extended
+significant effort in developing instructional materials teaching users
+how to organize, document, and execute reproducible pipelines that allow
+a user to go from raw sequence data to a compiled manuscript with
+figures through the Riffomonas project (65). A final example of a failed
+experiment was a collaboration with programmers through Google Summer of
+Code to develop commands in mothur that ran the random forest and SVM
+machine learning algorithms. Similar to the challenges of developing
+attractive visuals, fitting the algorithms' hyperparameters, testing,
+and deploying the resulting models require a significant amount of
+customization. Furthermore, machine learning is an active area of
+research where methods are still being developed and improved.
+Thankfully, there are numerous R and Python packages that do a better
+job of developing these models (71, 72). Again, we have put our efforts
 into developing instructional materials that mothur users can use to fit
@@
 real problems were straying from what mothur does well and failing to
-grasp what we really wanted the innovation to do. In hindsight, our
-solution to these failure has been to provide tutorials to as a conduit
-between mothur and their goals.
+grasp what we really wanted the innovation to do.
 
-\textbf{\emph{Competition is good and healthy.}}
+\textbf{\emph{The future.}} I will continue to develop mothur for as
+long as other researchers find it useful. One challenge of such a plan
+is maintaining the funding to support its development. The development
+of mothur was initially enabled by a subcontract from a Sloan Foundation
+grant to Mitch Sogin to support his VAMPS (Visualization and Analysis of
+Microbial Population Structures) initiative. We used that seed funding
+to secure an NSF grant and then a grant from NIH for tool development as
+part of their Human Microbiome Project. Since that project expired in
+2013, we have not had funding to specifically support mothur's
+development. I have been fortunate to have start-up and discretionary
+funds generated from other projects to help support mothur. Although
+there is funding for new tools, there appears to be little appetite by
+funders to support existing tools. Emblematic of this was the NIH
+program, Big Data To Knowledge (BD2K), which solicited proposals through
+the program announcement ``Extended Development, Hardening and
+Dissemination of Technologies in Biomedical Computing, Informatics and
+Big Data Science (PA-14-156)''. This opportunity appeared perfect,
+except that the National Institute of Allergy and Infectious Diseases
+(NIAID), the primary supporter of microbiome research at NIH, did not
+participate in the announcement. Tools like mothur are clearly
+successful, but need funding mechanisms to continue to mature and
+support the needs of the research community.
 
-\begin{itemize}
-\tightlist
-\item
-  A lot of people have been working in this space - e.g.~clustering
-  algorithms
-\item
-  Approaches to developing software have differed (e.g.~web-based
-  (greengenes/RDP), boutique (Quince), wrappers (QIIME))
-\end{itemize}
+As with anything in science, methods become \text{pass\'e}. When we
+first developed mothur, T-RFLP and DGGE were still commonly used. Today
+it would be hard to argue that data from those methods meaningfully
+advance a study relative to what one could get using 16S rRNA gene
+sequence data. Looking forward, many want to claim that amplicon
+sequencing is today's DGGE. They claim that researchers should instead
+move on to shotgun metagenomic sequencing. It is important to note that
+the two methods answer fundamentally different questions. 16S rRNA gene
+sequence data describes the taxonomic composition and metagenomic
+sequence data tells a researcher about the functional potential and
+genetic diversity of a community. Both tools provide important
+information but they cannot easily replace each other. Although
+metagenomic data does provide highly resolved taxonomic information, the
+limit of detection is at least an order of magnitude higher than that of
+amplicon data. For example, we analyzed 10,000 16S rRNA sequences from
+each of about 500 subjects (73). We can think of this as representing
+about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500
+subjects / 5 16S rRNA gene sequences/genome). Assuming a genome is 4
+Mbp, this would represent a sequencing depth of 4 Tbp. Although such a
+sequencing effort is technically possible, the cost of such an endeavor
+would be considerable and unlikely to be pursued by most researchers. We
+estimate that generating and sequencing the libraries at the University
+of Michigan sequencing core would cost approximately \$150 per library.
+The parallel 16S rRNA gene sequences data would cost approximately \$8
+per library. Furthermore, analyzing such a large dataset with an
+approach that captures the full genetic diversity of the community would
+be financially and technically prohibitive. Going forward, there is
+still a place for 16S rRNA gene sequencing. Although sequencing
+technologies will evolve to capture longer and more high quality data,
+there will likely always be a need for characterizing the taxonomic
+diversity of microbial communities. With this in mind, there will always
+be a place for tools like mothur that can analyze amplicon sequence
+data.
 
-What are we proud of? * Open source packages * Platform independent *
-Development of open instructional materials * Data driven development of
-methods * Database independent - allows people outside of our field to
-use mothur * Reproducibility * Helped create standards - when something
-is this popular, you force people to use a certain suite of tools and
-methods
+Of course this does not mean that such tools will remain static. We see
+three key areas that we will continue to help the field to move forward.
+First, just as we adapted through the transitions from Sanger to 454 to
+MiSeq and PacBio sequencing platforms (60--62), we must learn whether
+data from Oxford Nanopore can be an alternative sequencing approach that
+generates sequence data that is the same quality as existing approaches;
+thus far, the approach has significant shortcomings for sequencing 16S
+rRNA gene sequences (74). As with the earlier platforms, we must better
+understand its error profile so that sequencing errors can be corrected.
+We have learned that moving forward requires that we maintain or improve
+sequence quality. No doubt, datasets and read lengths will improve, but
+these advances should not be made at the cost of data quality. Second,
+with these improvements, we will need to continue to improve our
+algorithms. We have already seen that attempts to use low quality MiSeq
+and HiSeq data causes computational problems leading to the creation of
+open and closed reference clustering methods (75, 76). Unfortunately,
+comparative analyses showed that these methods fail relative to \emph{de
+novo} clustering methods (57). More work is needed to improve
+reference-based clustering methods so that larger datasets can be
+analyzed without sacrificing quality. Finally, there are ongoing
+controversies that need further exploration. These include the validity
+and utility of amplicon sequence variants (77), the wisdom of removing
+low frequency sequences (78), and methods of identifying and removing
+contaminant 16S rRNA gene sequences (79, 80). With each of these areas
+of development, the broader community can count on our same data-driven
+approach to meeting evaluating these questions. It is common for
+researchers to comment that they pick a specific method or deviate from
+a suggestion because they ``like how the data look''. When pressed for
+an objective definition of how they know the data look ``right'', they
+go quiet. Through the use of mock communities and simulations where we
+actually know what looks right and objective metrics of quality like the
+MCC or sequencing error rates, we will continue to base recommendations
+on data rather than a gut feeling.
 
-The future * 16S rRNA gene sequencing is not dead * Always improving
-sequencing methods bring new problems * Larger datasets, need a
-replacement for naive open and closed reference clustering * Lingering
-controversies - singletons, ASVs, contamination
-
-Longterm funding * The development of mothur was enabled by initial
-funding from a grant to Sogin from the Sloan Foundation to support his
-VAMPS (Visualization and Analysis of Microbial Population Structures)
-initiative. * NSF * NIH * Bootstrapping from other projects
-
-Red Queen
-
-\newpage
+\textbf{\emph{Conclusion.}} In the paper announcing mothur, we commented
+that the relationship between 16S rRNA gene sequencing and analysis is
+very much like the Red Queen in Lewis Carroll's book, \emph{Through the
+Looking-Glass}. Although some disagreed with this analogy (81), I still
+feel it is apt. The sequencing technology and rapacious appetite of
+researchers continues to race on. At the same time, bioinformatics tools
+must adapt to facilitate our research. I am confident that mothur will
+be up to this exciting challenge. Beyond its utility for analyzing
+amplicon sequence data, mothur's history provides lessons that are
+helpful for other projects that hope to develop a long historical arc.
+First, mothur is a product of its time. We have always sought to solve a
+current need to the best of our ability with the tools we had at the
+time. There are certainly caveats to any analysis of 16S rRNA gene
+sequence data, but if we had waited until those caveats were resolved,
+the field never would have progressed. Similarly, we made design choices
+that we probably would not have made had we started the project today.
+Second, as we have developed mothur, we have attempted to do so in a
+data-driven approach where we compare multiple methods. It has not
+merely been enough to propose a new method, we must show that it
+meaningfully advances the field. Third, through our failures and
+successes we have learned to focus on what mothur is good at and create
+products separate from mothur when distinct needs arise. For example, we
+have learned that mothur should not have a graphical interface or data
+visualization tools. Instead, we will provide instructional materials to
+teach users how to use the command line interface and other programs
+like R for data visualization. Finally, mothur was born out of a need
+for automating the analysis of large 16S rRNA gene sequence datasets. I
+realized I had a set of skills to fill that need. It has been refreshing
+to see the computational skills of the microbial ecology field grow over
+the past two decades. Looking ahead, we must all take stock of the
+challenges we face in microbial ecology and how our individual skills
+and interests can address these challenges to turn them into
+opportunities.
 
@@
 \subsection{Acknowledgements}\label{acknowledgements}}
+
+The development of mothur would not be possible without the
+contributions of its many supporters, developers, and users. Although
+not a complete list, I would be remiss if I did not express my gratitude
+to Sarah Westcott and the other members of my research group, Jo
+Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra
+Mass, and Marcy Balunas for their unique contributions to the continued
+development of mothur.
 
@@
 
-25 references
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Lenski2017}{}%
+1. \textbf{Lenski RE}. 2017. Experimental evolution and the dynamics of
+adaptation and genome evolution in microbial populations. The ISME
+Journal \textbf{11}:2181--2194.
+doi:\href{https://doi.org/10.1038/ismej.2017.69}{10.1038/ismej.2017.69}.
 
-\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Smith2018}{}%
+2. \textbf{Smith DK}. 2018. From fundamental supramolecular chemistry to
+self-assembled nanomaterials and medicines and back again how sam
+inspired SAMul. Chemical Communications \textbf{54}:4743--4760.
+doi:\href{https://doi.org/10.1039/c8cc01753k}{10.1039/c8cc01753k}.
+
+\leavevmode\hypertarget{ref-Barbour2019}{}%
+3. \textbf{Barbour AG}, \textbf{Benach JL}. 2019. Discovery of the lyme
+disease agent. mBio \textbf{10}:e02166--19.
+doi:\href{https://doi.org/10.1128/mbio.02166-19}{10.1128/mbio.02166-19}.
+
+\leavevmode\hypertarget{ref-Colwell2014}{}%
+4. \textbf{Colwell RK}, \textbf{Elsensohn JE}. 2014. EstimateS turns 20:
+Statistical estimation of species richness and shared species from
+samples, with non-parametric extrapolation. Ecography
+\textbf{37}:609--613.
+doi:\href{https://doi.org/10.1111/ecog.00814}{10.1111/ecog.00814}.
+
+\leavevmode\hypertarget{ref-Glckner2017}{}%
+5. \textbf{Glöckner FO}, \textbf{Yilmaz P}, \textbf{Quast C},
+\textbf{Gerken J}, \textbf{Beccati A}, \textbf{Ciuprina A},
+\textbf{Bruns G}, \textbf{Yarza P}, \textbf{Peplies J}, \textbf{Westram
+R}, \textbf{Ludwig W}. 2017. 25 years of serving the community with
+ribosomal RNA gene reference databases and tools. Journal of
+Biotechnology \textbf{261}:169--176.
+doi:\href{https://doi.org/10.1016/j.jbiotec.2017.06.1198}{10.1016/j.jbiotec.2017.06.1198}.
+
+\leavevmode\hypertarget{ref-Casadevall2015}{}%
+6. \textbf{Casadevall A}, \textbf{Fang FC}. 2015. (A)Historical science.
+Infection and Immunity \textbf{83}:4460--4464.
+doi:\href{https://doi.org/10.1128/iai.00921-15}{10.1128/iai.00921-15}.
+
+\leavevmode\hypertarget{ref-Schloss2009a}{}%
+7. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\leavevmode\hypertarget{ref-Hughes2001}{}%
+8. \textbf{Hughes JB}, \textbf{Hellmann JJ}, \textbf{Ricketts TH},
+\textbf{Bohannan BJM}. 2001. Counting the uncountable: Statistical
+approaches to estimating microbial diversity. Applied and Environmental
+Microbiology \textbf{67}:4399--4406.
+doi:\href{https://doi.org/10.1128/aem.67.10.4399-4406.2001}{10.1128/aem.67.10.4399-4406.2001}.
+
+\leavevmode\hypertarget{ref-Schloss2005}{}%
+9. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
+a computer program for defining operational taxonomic units and
+estimating species richness. Applied and Environmental Microbiology
+\textbf{71}:1501--1506.
+doi:\href{https://doi.org/10.1128/aem.71.3.1501-1506.2005}{10.1128/aem.71.3.1501-1506.2005}.
+
+\leavevmode\hypertarget{ref-Schloss2004a}{}%
+10. \textbf{Schloss PD}, \textbf{Larget BR}, \textbf{Handelsman J}.
+2004. Integration of microbial ecology and statistics: A test to compare
+gene libraries. Applied and Environmental Microbiology
+\textbf{70}:5485--5492.
+doi:\href{https://doi.org/10.1128/aem.70.9.5485-5492.2004}{10.1128/aem.70.9.5485-5492.2004}.
+
+\leavevmode\hypertarget{ref-Schloss2006b}{}%
+11. \textbf{Schloss PD}, \textbf{Handelsman J}. 2006. Introducing SONS,
+a tool for operational taxonomic unit-based comparisons of microbial
+community memberships and structures. Applied and Environmental
+Microbiology \textbf{72}:6773--6779.
+doi:\href{https://doi.org/10.1128/aem.00474-06}{10.1128/aem.00474-06}.
+
+\leavevmode\hypertarget{ref-Schloss2006a}{}%
+12. \textbf{Schloss PD}, \textbf{Handelsman J}. 2006. Introducing
+TreeClimber, a test to compare microbial community structures. Applied
+and Environmental Microbiology \textbf{72}:2379--2384.
+doi:\href{https://doi.org/10.1128/aem.72.4.2379-2384.2006}{10.1128/aem.72.4.2379-2384.2006}.
+
+\leavevmode\hypertarget{ref-Lozupone2005}{}%
+13. \textbf{Lozupone C}, \textbf{Knight R}. 2005. UniFrac: A new
+phylogenetic method for comparing microbial communities. Applied and
+Environmental Microbiology \textbf{71}:8228--8235.
+doi:\href{https://doi.org/10.1128/aem.71.12.8228-8235.2005}{10.1128/aem.71.12.8228-8235.2005}.
+
+\leavevmode\hypertarget{ref-Lozupone2007}{}%
+14. \textbf{Lozupone CA}, \textbf{Hamady M}, \textbf{Kelley ST},
+\textbf{Knight R}. 2007. Quantitative and qualitative ~ diversity
+measures lead to different insights into factors that structure
+microbial communities. Applied and Environmental Microbiology
+\textbf{73}:1576--1585.
+doi:\href{https://doi.org/10.1128/aem.01996-06}{10.1128/aem.01996-06}.
+
+\leavevmode\hypertarget{ref-Schloss2006c}{}%
+15. \textbf{Schloss PD}, \textbf{Handelsman J}. 2006. Toward a census of
+bacteria in soil. PLoS Computational Biology \textbf{2}:e92.
+doi:\href{https://doi.org/10.1371/journal.pcbi.0020092}{10.1371/journal.pcbi.0020092}.
+
+\leavevmode\hypertarget{ref-Schloss2004b}{}%
+16. \textbf{Schloss PD}, \textbf{Handelsman J}. 2004. Status of the
+microbial census. Microbiology and Molecular Biology Reviews
+\textbf{68}:686--691.
+doi:\href{https://doi.org/10.1128/mmbr.68.4.686-691.2004}{10.1128/mmbr.68.4.686-691.2004}.
+
+\leavevmode\hypertarget{ref-Schloss2007}{}%
+17. \textbf{Schloss PD}, \textbf{Handelsman J}. 2007. The last word:
+Books as a statistical metaphor for microbial communities. Annual Review
+of Microbiology \textbf{61}:23--34.
+
+\leavevmode\hypertarget{ref-Zhou1996}{}%
+18. \textbf{Zhou J}, \textbf{Bruns MA}, \textbf{Tiedje JM}. 1996. DNA
+recovery from soils of diverse composition. Applied and Environmental
+Microbiology \textbf{62}:316--322.
+
+\leavevmode\hypertarget{ref-Suzuki1996}{}%
+19. \textbf{Suzuki MT}, \textbf{Giovannoni SJ}. 1996. Bias caused by
+template annealing in the amplification of mixtures of 16S rRNA genes by
+pcr. Applied and Environmental Microbiology \textbf{62}:625--630.
+
+\leavevmode\hypertarget{ref-Chandler1997}{}%
+20. \textbf{Chandler DP}, \textbf{Fredrickson JK}, \textbf{Brockman FJ}.
+1997. Effect of pcr template concentration on the composition and
+distribution of total community 16S rDNA clone libraries. Molecular
+Ecology \textbf{6}:475--482.
+
+\leavevmode\hypertarget{ref-Polz1998}{}%
+21. \textbf{Polz MF}, \textbf{Cavanaugh CM}. 1998. Bias in
+template-to-product ratios in multitemplate pcr. Applied and
+Environmental Microbiology \textbf{64}:3724--3730.
+
+\leavevmode\hypertarget{ref-Wagner1994}{}%
+22. \textbf{Wagner A}, \textbf{Blackstone N}, \textbf{Cartwright P},
+\textbf{Dick M}, \textbf{Misof B}, \textbf{Snow P}, \textbf{Wagner GP},
+\textbf{Bartels J}, \textbf{Murtha M}, \textbf{Pendleton J}. 1994.
+Surveys of gene families using polymerase chain reaction: PCR selection
+and pcr drift. Systematic Biology \textbf{43}:250--261.
+
+\leavevmode\hypertarget{ref-Hansen1998}{}%
+23. \textbf{Hansen MC}, \textbf{Tolker-Nielsen T}, \textbf{Givskov M},
+\textbf{Molin S}. 1998. Biased 16S rDNA pcr amplification caused by
+interference from dna flanking the template region. FEMS Microbiology
+Ecology \textbf{26}:141--149.
+
+\leavevmode\hypertarget{ref-Qiu2001}{}%
+24. \textbf{Qiu X}, \textbf{Wu L}, \textbf{Huang H}, \textbf{McDonel
+PE}, \textbf{Palumbo AV}, \textbf{Tiedje JM}, \textbf{Zhou J}. 2001.
+Evaluation of pcr-generated chimeras, mutations, and heteroduplexes with
+16S rRNA gene-based cloning. Applied and Environmental Microbiology
+\textbf{67}:880--887.
+
+\leavevmode\hypertarget{ref-Komatsoulis1997}{}%
+25. \textbf{Komatsoulis GA}, \textbf{Waterman MS}. 1997. A new
+computational method for detection of chimeric 16S rRNA artifacts
+generated by pcr amplification from mixed bacterial populations. Applied
+and Environmental Microbiology \textbf{63}:2338--2346.
+
+\leavevmode\hypertarget{ref-Wang1997}{}%
+26. \textbf{Wang G}, \textbf{Wang Y}. 1997. Frequency of formation of
+chimeric molecules as a consequence of pcr coamplification of 16S rRNA
+genes from mixed bacterial genomes. Applied and Environmental
+Microbiology \textbf{63}:4645--4650.
+
+\leavevmode\hypertarget{ref-Hugenholtz2003}{}%
+27. \textbf{Hugenholtz P}, \textbf{Huber T}. 2003. Chimeric 16S rDNA
+sequences of diverse origin are accumulating in the public databases.
+International Journal of Systematic and Evolutionary Microbiology
+\textbf{53}:289--293.
+
+\leavevmode\hypertarget{ref-Bonfield1995}{}%
+28. \textbf{Bonfield JK}, \textbf{Smith KF}, \textbf{Staden R}. 1995. A
+new dna sequence assembly program. Nucleic Acids Research
+\textbf{23}:4992--4999.
+
+\leavevmode\hypertarget{ref-Thompson1994}{}%
+29. \textbf{Thompson JD}, \textbf{Higgins DG}, \textbf{Gibson TJ}. 1994.
+CLUSTAL w: Improving the sensitivity of progressive multiple sequence
+alignment through sequence weighting, position-specific gap penalties
+and weight matrix choice. Nucleic Acids Research \textbf{22}:4673--4680.
+doi:\href{https://doi.org/10.1093/nar/22.22.4673}{10.1093/nar/22.22.4673}.
+
+\leavevmode\hypertarget{ref-Ludwig2004}{}%
+30. \textbf{Ludwig W}. 2004. ARB: A software environment for sequence
+data. Nucleic Acids Research \textbf{32}:1363--1371.
+doi:\href{https://doi.org/10.1093/nar/gkh293}{10.1093/nar/gkh293}.
+
+\leavevmode\hypertarget{ref-Felsenstein1989}{}%
+31. \textbf{Felsenstein J}. 1989. PHYLIP - phylogeny inference package.
+Cladistics \textbf{5}:164--166.
+
+\leavevmode\hypertarget{ref-Stackebrandt1994}{}%
+32. \textbf{Stackebrandt E}, \textbf{Goebel BM}. 1994. Taxonomic note: A
+place for DNA-DNA reassociation and 16S rRNA sequence analysis in the
+present species definition in bacteriology. International Journal of
+Systematic and Evolutionary Microbiology \textbf{44}:846--849.
+doi:\href{https://doi.org/10.1099/00207713-44-4-846}{10.1099/00207713-44-4-846}.
+
+\leavevmode\hypertarget{ref-Seguritan2001}{}%
+33. \textbf{Seguritan V}, \textbf{Rohwer F}. 2001. FastGroup: A program
+to dereplicate libraries of 16S rDNA sequences. BMC Bioinformatics
+\textbf{2}:9.
+doi:\href{https://doi.org/10.1186/1471-2105-2-9}{10.1186/1471-2105-2-9}.
+
+\leavevmode\hypertarget{ref-McCaig1999}{}%
+34. \textbf{McCaig AE}, \textbf{Glover LA}, \textbf{Prosser JI}. 1999.
+Molecular analysis of bacterial community structure and diversity in
+unimproved and improved upland grass pastures. Applied and Environmental
+Microbiology \textbf{65}:1721--1730.
+
+\leavevmode\hypertarget{ref-Pollock2018}{}%
+35. \textbf{Pollock J}, \textbf{Glendinning L}, \textbf{Wisedchanwet T},
+\textbf{Watson M}. 2018. The madness of microbiome: Attempting to find
+consensus ``Best practice'' for 16S microbiome studies. Applied and
+Environmental Microbiology \textbf{84}:e02627--17.
+doi:\href{https://doi.org/10.1128/aem.02627-17}{10.1128/aem.02627-17}.
+
+\leavevmode\hypertarget{ref-Singleton2001}{}%
+36. \textbf{Singleton DR}, \textbf{Furlong MA}, \textbf{Rathbun SL},
+\textbf{Whitman WB}. 2001. Quantitative comparisons of 16S rRNA gene
+sequence libraries from environmental samples. Applied and Environmental
+Microbiology \textbf{67}:4374--4376.
+doi:\href{https://doi.org/10.1128/aem.67.9.4374-4376.2001}{10.1128/aem.67.9.4374-4376.2001}.
+
+\leavevmode\hypertarget{ref-Carey2018}{}%
+37. \textbf{Carey MA}, \textbf{Papin JA}. 2018. Ten simple rules for
+biologists learning to program. PLOS Computational Biology
+\textbf{14}:e1005871.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1005871}{10.1371/journal.pcbi.1005871}.
+
+\leavevmode\hypertarget{ref-Sogin2006}{}%
+38. \textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber JA},
+\textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR}, \textbf{Arrieta
+JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in the deep sea and
+the underexplored "rare biosphere". Proceedings of the National Academy
+of Sciences \textbf{103}:12115--12120.
+doi:\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.
+
+\leavevmode\hypertarget{ref-Cole2013}{}%
+39. \textbf{Cole JR}, \textbf{Wang Q}, \textbf{Fish JA}, \textbf{Chai
+B}, \textbf{McGarrell DM}, \textbf{Sun Y}, \textbf{Brown CT},
+\textbf{Porras-Alfaro A}, \textbf{Kuske CR}, \textbf{Tiedje JM}. 2013.
+Ribosomal database project: Data and tools for high throughput rRNA
+analysis. Nucleic Acids Research \textbf{42}:D633--D642.
+doi:\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+
+\leavevmode\hypertarget{ref-DeSantis2006a}{}%
+40. \textbf{DeSantis TZ}, \textbf{Hugenholtz P}, \textbf{Larsen N},
+\textbf{Rojas M}, \textbf{Brodie EL}, \textbf{Keller K}, \textbf{Huber
+T}, \textbf{Dalevi D}, \textbf{Hu P}, \textbf{Andersen GL}. 2006.
+Greengenes, a chimera-checked 16S rRNA gene database and workbench
+compatible with ARB. Applied and Environmental Microbiology
+\textbf{72}:5069--5072.
+doi:\href{https://doi.org/10.1128/aem.03006-05}{10.1128/aem.03006-05}.
+
+\leavevmode\hypertarget{ref-Yilmaz2013}{}%
+41. \textbf{Yilmaz P}, \textbf{Parfrey LW}, \textbf{Yarza P},
+\textbf{Gerken J}, \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Schweer
+T}, \textbf{Peplies J}, \textbf{Ludwig W}, \textbf{Glöckner FO}. 2013.
+The SILVA and ``All-species living tree project (LTP)'' taxonomic
+frameworks. Nucleic Acids Research \textbf{42}:D643--D648.
+doi:\href{https://doi.org/10.1093/nar/gkt1209}{10.1093/nar/gkt1209}.
+
+\leavevmode\hypertarget{ref-DeSantis2006b}{}%
+42. \textbf{DeSantis TZ}, \textbf{Hugenholtz P}, \textbf{Keller K},
+\textbf{Brodie EL}, \textbf{Larsen N}, \textbf{Piceno YM}, \textbf{Phan
+R}, \textbf{Andersen GL}. 2006. NAST: A multiple sequence alignment
+server for comparative analysis of 16S rRNA genes. Nucleic Acids
+Research \textbf{34}:W394--W399.
+doi:\href{https://doi.org/10.1093/nar/gkl244}{10.1093/nar/gkl244}.
+
+\leavevmode\hypertarget{ref-Schloss2009b}{}%
+43. \textbf{Schloss PD}. 2009. A high-throughput DNA sequence aligner
+for microbial ecology studies. PLoS ONE \textbf{4}:e8230.
+doi:\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.
+
+\leavevmode\hypertarget{ref-Schloss2010a}{}%
+44. \textbf{Schloss PD}. 2010. The effects of alignment quality,
+distance calculation method, sequence filtering, and region on the
+analysis of 16S rRNA gene-based studies. PLoS Computational Biology
+\textbf{6}:e1000844.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1000844}{10.1371/journal.pcbi.1000844}.
+
+\leavevmode\hypertarget{ref-Wang2007}{}%
+45. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
+of rRNA sequences into the new bacterial taxonomy. Applied and
+Environmental Microbiology \textbf{73}:5261--5267.
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
+
+\leavevmode\hypertarget{ref-Caporaso2010a}{}%
+46. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
+\textbf{Bittinger K}, \textbf{Bushman FD}, \textbf{Costello EK},
+\textbf{Fierer N}, \textbf{Peña AG}, \textbf{Goodrich JK},
+\textbf{Gordon JI}, \textbf{Huttley GA}, \textbf{Kelley ST},
+\textbf{Knights D}, \textbf{Koenig JE}, \textbf{Ley RE},
+\textbf{Lozupone CA}, \textbf{McDonald D}, \textbf{Muegge BD},
+\textbf{Pirrung M}, \textbf{Reeder J}, \textbf{Sevinsky JR},
+\textbf{Turnbaugh PJ}, \textbf{Walters WA}, \textbf{Widmann J},
+\textbf{Yatsunenko T}, \textbf{Zaneveld J}, \textbf{Knight R}. 2010.
+QIIME allows analysis of high-throughput community sequencing data.
+Nature Methods \textbf{7}:335--336.
+doi:\href{https://doi.org/10.1038/nmeth.f.303}{10.1038/nmeth.f.303}.
+
+\leavevmode\hypertarget{ref-Quince2009}{}%
+47. \textbf{Quince C}, \textbf{Lanzén A}, \textbf{Curtis TP},
+\textbf{Davenport RJ}, \textbf{Hall N}, \textbf{Head IM}, \textbf{Read
+LF}, \textbf{Sloan WT}. 2009. Accurate determination of microbial
+diversity from 454 pyrosequencing data. Nature Methods
+\textbf{6}:639--641.
+doi:\href{https://doi.org/10.1038/nmeth.1361}{10.1038/nmeth.1361}.
+
+\leavevmode\hypertarget{ref-Masella2012}{}%
+48. \textbf{Masella AP}, \textbf{Bartram AK}, \textbf{Truszkowski JM},
+\textbf{Brown DG}, \textbf{Neufeld JD}. 2012. PANDAseq: Paired-end
+assembler for illumina sequences. BMC Bioinformatics \textbf{13}:31.
+doi:\href{https://doi.org/10.1186/1471-2105-13-31}{10.1186/1471-2105-13-31}.
+
+\leavevmode\hypertarget{ref-Callahan2016}{}%
+49. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
+\textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes SP}. 2016. DADA2:
+High-resolution sample inference from illumina amplicon data. Nature
+Methods \textbf{13}:581--583.
+doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.
+
+\leavevmode\hypertarget{ref-Edgar2011}{}%
+50. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
+and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+
+\leavevmode\hypertarget{ref-Haas2011}{}%
+51. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
+\textbf{Feldgarden M}, \textbf{Ward DV}, \textbf{Giannoukos G},
+\textbf{Ciulla D}, \textbf{Tabbaa D}, \textbf{Highlander SK},
+\textbf{Sodergren E}, \textbf{Methe B}, \textbf{DeSantis TZ},
+\textbf{Petrosino JF}, \textbf{Knight R}, \textbf{and BWB}. 2011.
+Chimeric 16S rRNA sequence formation and detection in Sanger and
+454-pyrosequenced PCR amplicons. Genome Research \textbf{21}:494--504.
+doi:\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
+
+\leavevmode\hypertarget{ref-Quince2011}{}%
+52. \textbf{Quince C}, \textbf{Lanzen A}, \textbf{Davenport RJ},
+\textbf{Turnbaugh PJ}. 2011. Removing noise from pyrosequenced
+amplicons. BMC Bioinformatics \textbf{12}:38.
+doi:\href{https://doi.org/10.1186/1471-2105-12-38}{10.1186/1471-2105-12-38}.
+
+\leavevmode\hypertarget{ref-Edgar2010}{}%
+53. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
+faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
+doi:\href{https://doi.org/10.1093/bioinformatics/btq461}{10.1093/bioinformatics/btq461}.
+
+\leavevmode\hypertarget{ref-Rognes2016}{}%
+54. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
+source tool for metagenomics. PeerJ \textbf{4}:e2584.
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\leavevmode\hypertarget{ref-Mah2015}{}%
+55. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
+\textbf{Vargas C de}, \textbf{Dunthorn M}. 2015. Swarm v2:
+Highly-scalable and high-resolution amplicon clustering. PeerJ
+\textbf{3}:e1420.
+doi:\href{https://doi.org/10.7717/peerj.1420}{10.7717/peerj.1420}.
+
+\leavevmode\hypertarget{ref-Schloss2011a}{}%
+56. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
+improving methods used in operational taxonomic unit-based approaches
+for 16S rRNA gene sequence analysis. Applied and Environmental
+Microbiology \textbf{77}:3219--3226.
+doi:\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.
+
+\leavevmode\hypertarget{ref-Westcott2015}{}%
+57. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
+methods outperform reference-based methods for assigning 16S rRNA gene
+sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
+
+\leavevmode\hypertarget{ref-Schloss2016a}{}%
+58. \textbf{Schloss PD}. 2016. Application of a database-independent
+approach to assess the quality of operational taxonomic unit picking
+methods. mSystems \textbf{1}:e00027--16.
+doi:\href{https://doi.org/10.1128/msystems.00027-16}{10.1128/msystems.00027-16}.
+
+\leavevmode\hypertarget{ref-Westcott2017}{}%
+59. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+improved method for assigning amplicon-based sequence data to
+operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+
+\leavevmode\hypertarget{ref-Kozich2013}{}%
+60. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\leavevmode\hypertarget{ref-Schloss2011b}{}%
+61. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
+Reducing the effects of PCR amplification and sequencing artifacts on
+16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
+doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+
+\leavevmode\hypertarget{ref-Schloss2016b}{}%
+62. \textbf{Schloss PD}, \textbf{Jenior ML}, \textbf{Koumpouras CC},
+\textbf{Westcott SL}, \textbf{Highlander SK}. 2016. Sequencing 16S rRNA
+gene fragments using the PacBio SMRT DNA sequencing system. PeerJ
+\textbf{4}:e1869.
+doi:\href{https://doi.org/10.7717/peerj.1869}{10.7717/peerj.1869}.
+
+\leavevmode\hypertarget{ref-Schloss2012a}{}%
+63. \textbf{Schloss PD}. 2012. Secondary structure improves OTU
+assignments of 16S rRNA gene sequences. The ISME Journal
+\textbf{7}:457--460.
+doi:\href{https://doi.org/10.1038/ismej.2012.102}{10.1038/ismej.2012.102}.
+
+\leavevmode\hypertarget{ref-Schloss2008}{}%
+64. \textbf{Schloss PD}. 2008. Evaluating different approaches that test
+whether microbial communities have the same structure. The ISME Journal
+\textbf{2}:265--275.
+doi:\href{https://doi.org/10.1038/ismej.2008.5}{10.1038/ismej.2008.5}.
+
+\leavevmode\hypertarget{ref-Schloss2018a}{}%
+65. \textbf{Schloss PD}. 2018. The riffomonas reproducible research
+tutorial series. Journal of Open Source Education \textbf{1}:13.
+doi:\href{https://doi.org/10.21105/jose.00013}{10.21105/jose.00013}.
+
+\leavevmode\hypertarget{ref-Eddelbuettel2011}{}%
+66. \textbf{Eddelbuettel D}, \textbf{François R}. 2011. Rcpp: Seamless R
+and C++ integration. Journal of Statistical Software \textbf{40}:1--18.
+doi:\href{https://doi.org/10.18637/jss.v040.i08}{10.18637/jss.v040.i08}.
+
+\leavevmode\hypertarget{ref-VzquezBaeza2013}{}%
+67. \textbf{Vázquez-Baeza Y}, \textbf{Pirrung M}, \textbf{Gonzalez A},
+\textbf{Knight R}. 2013. EMPeror: A tool for visualizing high-throughput
+microbial community data. GigaScience \textbf{2}:16.
+doi:\href{https://doi.org/10.1186/2047-217x-2-16}{10.1186/2047-217x-2-16}.
+
+\leavevmode\hypertarget{ref-Bolyen2019}{}%
+68. \textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon MR},
+\textbf{Bokulich NA}, \textbf{Abnet CC}, \textbf{Al-Ghalith GA},
+\textbf{Alexander H}, \textbf{Alm EJ}, \textbf{Arumugam M},
+\textbf{Asnicar F}, \textbf{Bai Y}, \textbf{Bisanz JE},
+\textbf{Bittinger K}, \textbf{Brejnrod A}, \textbf{Brislawn CJ},
+\textbf{Brown CT}, \textbf{Callahan BJ}, \textbf{Caraballo-Rodrı́guez
+AM}, \textbf{Chase J}, \textbf{Cope EK}, \textbf{Silva RD},
+\textbf{Diener C}, \textbf{Dorrestein PC}, \textbf{Douglas GM},
+\textbf{Durall DM}, \textbf{Duvallet C}, \textbf{Edwardson CF},
+\textbf{Ernst M}, \textbf{Estaki M}, \textbf{Fouquier J},
+\textbf{Gauglitz JM}, \textbf{Gibbons SM}, \textbf{Gibson DL},
+\textbf{Gonzalez A}, \textbf{Gorlick K}, \textbf{Guo J},
+\textbf{Hillmann B}, \textbf{Holmes S}, \textbf{Holste H},
+\textbf{Huttenhower C}, \textbf{Huttley GA}, \textbf{Janssen S},
+\textbf{Jarmusch AK}, \textbf{Jiang L}, \textbf{Kaehler BD},
+\textbf{Kang KB}, \textbf{Keefe CR}, \textbf{Keim P}, \textbf{Kelley
+ST}, \textbf{Knights D}, \textbf{Koester I}, \textbf{Kosciolek T},
+\textbf{Kreps J}, \textbf{Langille MGI}, \textbf{Lee J}, \textbf{Ley R},
+\textbf{Liu Y-X}, \textbf{Loftfield E}, \textbf{Lozupone C},
+\textbf{Maher M}, \textbf{Marotz C}, \textbf{Martin BD},
+\textbf{McDonald D}, \textbf{McIver LJ}, \textbf{Melnik AV},
+\textbf{Metcalf JL}, \textbf{Morgan SC}, \textbf{Morton JT},
+\textbf{Naimey AT}, \textbf{Navas-Molina JA}, \textbf{Nothias LF},
+\textbf{Orchanian SB}, \textbf{Pearson T}, \textbf{Peoples SL},
+\textbf{Petras D}, \textbf{Preuss ML}, \textbf{Pruesse E},
+\textbf{Rasmussen LB}, \textbf{Rivers A}, \textbf{Robeson MS},
+\textbf{Rosenthal P}, \textbf{Segata N}, \textbf{Shaffer M},
+\textbf{Shiffer A}, \textbf{Sinha R}, \textbf{Song SJ}, \textbf{Spear
+JR}, \textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Torres PJ},
+\textbf{Trinh P}, \textbf{Tripathi A}, \textbf{Turnbaugh PJ},
+\textbf{Ul-Hasan S}, \textbf{Hooft JJJ van der}, \textbf{Vargas F},
+\textbf{Vázquez-Baeza Y}, \textbf{Vogtmann E}, \textbf{Hippel M von},
+\textbf{Walters W}, \textbf{Wan Y}, \textbf{Wang M}, \textbf{Warren J},
+\textbf{Weber KC}, \textbf{Williamson CHD}, \textbf{Willis AD},
+\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhang Y}, \textbf{Zhu Q},
+\textbf{Knight R}, \textbf{Caporaso JG}. 2019. Reproducible,
+interactive, scalable and extensible microbiome data science using QIIME
+2. Nature Biotechnology \textbf{37}:852--857.
+doi:\href{https://doi.org/10.1038/s41587-019-0209-9}{10.1038/s41587-019-0209-9}.
+
+\leavevmode\hypertarget{ref-Gonzalez2018}{}%
+69. \textbf{Gonzalez A}, \textbf{Navas-Molina JA}, \textbf{Kosciolek T},
+\textbf{McDonald D}, \textbf{Vázquez-Baeza Y}, \textbf{Ackermann G},
+\textbf{DeReus J}, \textbf{Janssen S}, \textbf{Swafford AD},
+\textbf{Orchanian SB}, \textbf{Sanders JG}, \textbf{Shorenstein J},
+\textbf{Holste H}, \textbf{Petrus S}, \textbf{Robbins-Pianka A},
+\textbf{Brislawn CJ}, \textbf{Wang M}, \textbf{Rideout JR},
+\textbf{Bolyen E}, \textbf{Dillon M}, \textbf{Caporaso JG},
+\textbf{Dorrestein PC}, \textbf{Knight R}. 2018. Qiita: Rapid,
+web-enabled microbiome meta-analysis. Nature Methods
+\textbf{15}:796--798.
+doi:\href{https://doi.org/10.1038/s41592-018-0141-9}{10.1038/s41592-018-0141-9}.
+
+\leavevmode\hypertarget{ref-Schloss2018b}{}%
+70. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
+reproducibility, replicability, robustness, and generalizability in
+microbiome research. mBio \textbf{9}:e00525--18.
+doi:\href{https://doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.
+
+\leavevmode\hypertarget{ref-Paszke2017}{}%
+71. \textbf{Paszke A}, \textbf{Gross S}, \textbf{Chintala S},
+\textbf{Chanan G}, \textbf{Yang E}, \textbf{DeVito Z}, \textbf{Lin Z},
+\textbf{Desmaison A}, \textbf{Antiga L}, \textbf{Lerer A}. 2017.
+Automatic differentiation in PyTorch. \emph{In} NIPS autodiff workshop.
+
+\leavevmode\hypertarget{ref-Kuhn2008}{}%
+72. \textbf{Kuhn M}. 2008. Building predictive models in R using the
+caret package. Journal of Statistical Software, Articles
+\textbf{28}:1--26.
+doi:\href{https://doi.org/10.18637/jss.v028.i05}{10.18637/jss.v028.i05}.
+
+\leavevmode\hypertarget{ref-Baxter2016}{}%
+73. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\leavevmode\hypertarget{ref-Calus2018}{}%
+74. \textbf{Calus ST}, \textbf{Ijaz UZ}, \textbf{Pinto AJ}. 2018.
+NanoAmpli-seq: A workflow for amplicon sequencing for mixed microbial
+communities on the nanopore sequencing platform. GigaScience
+\textbf{7}:12.
+doi:\href{https://doi.org/10.1093/gigascience/giy140}{10.1093/gigascience/giy140}.
+
+\leavevmode\hypertarget{ref-NavasMolina2013}{}%
+75. \textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
+\textbf{González A}, \textbf{McMurdie PJ}, \textbf{Vázquez-Baeza Y},
+\textbf{Xu Z}, \textbf{Ursell LK}, \textbf{Lauber C}, \textbf{Zhou H},
+\textbf{Song SJ}, \textbf{Huntley J}, \textbf{Ackermann GL},
+\textbf{Berg-Lyons D}, \textbf{Holmes S}, \textbf{Caporaso JG},
+\textbf{Knight R}. 2013. Advancing our understanding of the human
+microbiome using QIIME, pp. 371--444. \emph{In} Methods in Enzymology.
+Elsevier.
+
+\leavevmode\hypertarget{ref-Rideout2014}{}%
+76. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
+\textbf{Walters WA}, \textbf{Ursell LK}, \textbf{Gibbons SM},
+\textbf{Chase J}, \textbf{McDonald D}, \textbf{Gonzalez A},
+\textbf{Robbins-Pianka A}, \textbf{Clemente JC}, \textbf{Gilbert JA},
+\textbf{Huse SM}, \textbf{Zhou H-W}, \textbf{Knight R}, \textbf{Caporaso
+JG}. 2014. Subsampled open-reference clustering creates consistent,
+comprehensive OTU definitions and scales to billions of sequences. PeerJ
+\textbf{2}:e545.
+doi:\href{https://doi.org/10.7717/peerj.545}{10.7717/peerj.545}.
+
+\leavevmode\hypertarget{ref-Callahan2017}{}%
+77. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Holmes SP}.
+2017. Exact sequence variants should replace operational taxonomic units
+in marker-gene data analysis. The ISME Journal \textbf{11}:2639--2643.
+doi:\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.
+
+\leavevmode\hypertarget{ref-Bokulich2012}{}%
+78. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
+\textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight R}, \textbf{Mills
+DA}, \textbf{Caporaso JG}. 2012. Quality-filtering vastly improves
+diversity estimates from illumina amplicon sequencing. Nature Methods
+\textbf{10}:57--59.
+doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.
+
+\leavevmode\hypertarget{ref-Salter2014}{}%
+79. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
+\textbf{Calus ST}, \textbf{Cookson WO}, \textbf{Moffatt MF},
+\textbf{Turner P}, \textbf{Parkhill J}, \textbf{Loman NJ},
+\textbf{Walker AW}. 2014. Reagent and laboratory contamination can
+critically impact sequence-based microbiome analyses. BMC Biology
+\textbf{12}:87.
+doi:\href{https://doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
+
+\leavevmode\hypertarget{ref-Davis2018}{}%
+80. \textbf{Davis NM}, \textbf{Proctor DM}, \textbf{Holmes SP},
+\textbf{Relman DA}, \textbf{Callahan BJ}. 2018. Simple statistical
+identification and removal of contaminant sequences in marker-gene and
+metagenomics data. Microbiome \textbf{6}:1.
+doi:\href{https://doi.org/10.1186/s40168-018-0605-2}{10.1186/s40168-018-0605-2}.
+
+\leavevmode\hypertarget{ref-Caporaso2010b}{}%
+81. \textbf{Caporaso JG}, \textbf{Lauber CL}, \textbf{Walters WA},
+\textbf{Berg-Lyons D}, \textbf{Lozupone CA}, \textbf{Turnbaugh PJ},
+\textbf{Fierer N}, \textbf{Knight R}. 2010. Global patterns of 16S rRNA
+diversity at a depth of millions of sequences per sample. Proceedings of
+the National Academy of Sciences \textbf{108}:4516--4522.
+doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  Timeline of mothur/QIIME/etc citations
+\item
+  Timeline of SRA deposits / 16S/microbiome in PubMed
+\end{itemize}
+
+\textbf{Figure 2. Caption caption caption.} Footnotes footnotes
+footnotes.
+
+\begin{itemize}
+\tightlist
+\item
+  Screenshot of mothur welcoming page
+\end{itemize}
+
+\textbf{Figure 3. Caption caption caption.} Footnotes footnotes
+footnotes.
+
+\begin{itemize}
+\tightlist
+\item
+  Screenshot of mothur homepage
+\end{itemize}
+
 \end{document}

### 2019-10-04 submission/manuscript.Rmd
Add abstract

@@
 ## Abstract
-250 words
-
-\newpage
-
-## Importance
-150 words
+More than 10 years ago, we published the manuscript describing the mothur software package in *Applied and Environmental Microbiology*.
+Our goal was to create a comprehensive package that allowed users to analyze amplicon sequence data using the most robust methods available.
+Looking back, it is clear that mothur continues to provide this service to the microbial ecology community as it has helped to lead the community through the ongoing sequencing revolution.
+Beyond its success and impact on the field, mothur's development exposed a series of observations that are generally translatable across science.
+Although it is easy to criticize choices that were made 10 years ago through a modern lens, all science is done in the context of prevailing ideas and available technologies.
+If we were to wait for all of the possible limitations to be solved before proceeding, science would stall.
+Even preceding the development of mothur, it was necessary to address the most important problems and work backwards to other problems that limited access to robust sequence analysis tools.
+At the same time, we strive to expand mothur's capabilities in a data-driven manner to incorporate new ideas and accommodate changes in data and desires of the research community.
+As we develop mothur, we first consider the needs of my research group.
+In hindsight, we appreciate that our needs are representative of the broader community's.
+It has been edifying to see the benefit that a simple set of tools can bring to so many other researchers.
 

### 2019-10-04 submission/manuscript.tex
Add abstract

@@
 
-250 words
-
-\newpage
-
-\hypertarget{importance}{%
-\subsection{Importance}\label{importance}}
-
-150 words
+More than 10 years ago, we published the manuscript describing the
+mothur software package in \emph{Applied and Environmental
+Microbiology}. Our goal was to create a comprehensive package that
+allowed users to analyze amplicon sequence data using the most robust
+methods available. Looking back, it is clear that mothur continues to
+provide this service to the microbial ecology community as it has helped
+to lead the community through the ongoing sequencing revolution. Beyond
+its success and impact on the field, mothur's development exposed a
+series of observations that are generally translatable across science.
+Although it is easy to criticize choices that were made 10 years ago
+through a modern lens, all science is done in the context of prevailing
+ideas and available technologies. If we were to wait for all of the
+possible limitations to be solved before proceeding, science would
+stall. Even preceding the development of mothur, it was necessary to
+address the most important problems and work backwards to other problems
+that limited access to robust sequence analysis tools. At the same time,
+we strive to expand mothur's capabilities in a data-driven manner to
+incorporate new ideas and accommodate changes in data and desires of the
+research community. As we develop mothur, we first consider the needs of
+my research group. In hindsight, we appreciate that our needs are
+representative of the broader community's. It has been edifying to see
+the benefit that a simple set of tools can bring to so many other
+researchers.
 

### 2019-10-04 submission/manuscript.Rmd
Add figures

@@
 
-<!--- \includegraphics{figure_1.png} -->
+\includegraphics{figure_1.png}
 
-**Figure 1. Caption caption caption.** Footnotes footnotes footnotes.
+**Figure 1.
+mothur has consistently been a popular software package over the past ten years with a total of more than 8,800 citations.** Citation data taken from the Web of Science (https://www.webofscience.com) on October 1, 2019.
+Data from 2019 are incomplete.
 
-* Timeline of mothur/QIIME/etc citations
-* Timeline of SRA deposits / 16S/microbiome in PubMed
+\newpage
 
-**Figure 2. Caption caption caption.** Footnotes footnotes footnotes.
+\includegraphics{figure_2.png}
 
-* Screenshot of mothur welcoming page
+**Figure 2.
+The mothur homepage.** From the mothur home page at www.mothur.org, users can download mothur, access a user forum, navigate a wiki with extensive documentation, join the mothur facebook group, and subscribe to the mothur mailing list.
 
-**Figure 3. Caption caption caption.** Footnotes footnotes footnotes.
+\newpage
 
-* Screenshot of mothur homepage
+\includegraphics{figure_3.png}
+
+**Figure 3.
+The start up window when running mothur in Mac OS X in the interactive mode.** mothur can also be run on Windows or Linux.
+In the interactive mode users enter individual commands at the mothur prompt.
+Alternatively, users may run mothur by supplying commands from the command line or using batch scripts.

### 2019-10-04 submission/manuscript.tex
Add figures

@@
 
-\textbf{Figure 1. Caption caption caption.} Footnotes footnotes
-footnotes.
-
-\begin{itemize}
-\tightlist
-\item
-  Timeline of mothur/QIIME/etc citations
-\item
-  Timeline of SRA deposits / 16S/microbiome in PubMed
-\end{itemize}
-
-\textbf{Figure 2. Caption caption caption.} Footnotes footnotes
-footnotes.
-
-\begin{itemize}
-\tightlist
-\item
-  Screenshot of mothur welcoming page
-\end{itemize}
-
-\textbf{Figure 3. Caption caption caption.} Footnotes footnotes
-footnotes.
-
-\begin{itemize}
-\tightlist
-\item
-  Screenshot of mothur homepage
-\end{itemize}
+\includegraphics{figure_1.png}
+
+\textbf{Figure 1. mothur has consistently been a popular software
+package over the past ten years with a total of more than 8,800
+citations.} Citation data taken from the Web of Science
+(\url{https://www.webofscience.com}) on October 1, 2019. Data from 2019
+are incomplete.
+
+\newpage
+
+\includegraphics{figure_2.png}
+
+\textbf{Figure 2. The mothur homepage.} From the mothur home page at
+www.mothur.org, users can download mothur, access a user forum, navigate
+a wiki with extensive documentation, join the mothur facebook group, and
+subscribe to the mothur mailing list.
+
+\newpage
+
+\includegraphics{figure_3.png}
+
+\textbf{Figure 3. The start up window when running mothur in Mac OS X in
+the interactive mode.} mothur can also be run on Windows or Linux. In
+the interactive mode users enter individual commands at the mothur
+prompt. Alternatively, users may run mothur by supplying commands from
+the command line or using batch scripts.
 

### 2019-10-04 code/plot_citations.R

@@ -1,0 +1,12 @@
+library(tidyverse)
+
+read_tsv('data/citations.tsv') %>%
+	gather(-year, key="tool", value="citations") %>%
+	filter(tool == "mothur") %>%
+	ggplot(aes(x=year, y=citations)) +
+		geom_line() +
+		theme_classic() +
+		labs(x="Year", y="Number of Citations") +
+		scale_x_continuous(breaks=seq(2009,2019,2), labels=seq(2009,2019,2))
+
+ggsave("figures/mothur_citations.pdf", width=6, height=4, units="in")

### 2019-10-09 submission/manuscript.Rmd
Incorporate MB / KS comments

@@
 Our goal was to create a comprehensive package that allowed users to analyze amplicon sequence data using the most robust methods available.
-Looking back, it is clear that mothur continues to provide this service to the microbial ecology community as it has helped to lead the community through the ongoing sequencing revolution.
+mothur has helped lead the community through the ongoing sequencing revolution and continues to provide this service to the microbial ecology community.
 Beyond its success and impact on the field, mothur's development exposed a series of observations that are generally translatable across science.
-Although it is easy to criticize choices that were made 10 years ago through a modern lens, all science is done in the context of prevailing ideas and available technologies.
-If we were to wait for all of the possible limitations to be solved before proceeding, science would stall.
+Perhaps the observation that stands out the most is that all science is done in the context of prevailing ideas and available technologies.
+Although it is easy to criticize choices that were made 10 years ago through a modern lens, if we were to wait for all of the possible limitations to be solved before proceeding, science would stall.
 Even preceding the development of mothur, it was necessary to address the most important problems and work backwards to other problems that limited access to robust sequence analysis tools.
 At the same time, we strive to expand mothur's capabilities in a data-driven manner to incorporate new ideas and accommodate changes in data and desires of the research community.
-As we develop mothur, we first consider the needs of my research group.
-In hindsight, we appreciate that our needs are representative of the broader community's.
 It has been edifying to see the benefit that a simple set of tools can bring to so many other researchers.
@@
 Looking back on scientific journeys can be instructive to others who are overwhelmed at the prospect of looking forward at their careers [@Lenski2017; @Smith2018; @Barbour2019, @Casadevall2015; @Colwell2014; @Glckner2017].
-By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on.
+By no means is my scientific journey over, but since 2002 I have been on a journey that I did not realize I was on.
 Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology* [@Schloss2009a], it is worth stepping back and using the continued development of mothur as a story that has parallels to many other research stories.
@@
 This wasn't trivial, but by cobbling together other tools it was possible.
-We would assemble, trim and and correct Sanger sequence reads using Chromas or STADEN [@Bonfield1995], align the sequences using ClustalW [@Thompson1994] or ARB [@Ludwig2004], check for chimeras using partial treeing or Bellerephon [@Hugenholtz2003], and calculate a pairwise distance matrix using DNADIST from the PHYLIP package [@Felsenstein1989].
+We would assemble, trim and correct Sanger sequence reads using Chromas or STADEN [@Bonfield1995], align the sequences using ClustalW [@Thompson1994] or ARB [@Ludwig2004], check for chimeras using partial treeing or Bellerephon [@Hugenholtz2003], and calculate a pairwise distance matrix using DNADIST from the PHYLIP package [@Felsenstein1989].
 At the time, we knew that we only had a loose concept of a species based on these distances [@Stackebrandt1994].
@@
 At the time, I felt that the biggest problems that I could solve were how to cluster the sequences into OTUs and how to use those clusterings to test our hypotheses.
-The only tool available at the time that automated the clustering step was FastGroup, which implemented a greedy version of the single linkage algorithm [@Seguritan2001].
+The only tool available at the time that automated the clustering step was FastGroup, which implemented an approximation of the single linkage algorithm [@Seguritan2001].
 The high cost of sequencing was also an impediment to experimentation and analysis in microbial ecology.
 It was rare for a study design to have experimental replicates so that one could perform a statistical test to compare treatment groups.
-For example, in our testing we frequently used a dataset comparing Scottish soils from from Alison McCaig and colleagues [@McCaig1999].
+For example, in our testing we frequently used a dataset comparing Scottish soils from Alison McCaig and colleagues [@McCaig1999].
 This dataset consisted of two experimental groups, each replicated three times with 45 sequences per replicate.
@@
 My undergraduate and graduate training as a biological engineer prepared me to think about research questions from a systems perspective, to think quantitatively, and to understand the value of using computer programs to help solve problems.
-As an undergraduate student, I learned the Pascal programming language and promptly forgot much of it as a graduate student and learned MatLab in its place.
-As a postdoc, I learned the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [@Singleton2001].
-After writing my own version of LIBSHUFF, $\int$-LIBSHUFF and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
+As an undergraduate student, I learned the Pascal programming language and promptly forgot much of it.
+Although it was a good language for teaching programming concepts, it did not catch on outside of the classroom.
+Later, I learned MATLAB.
+Because it was an expensive commercial programming environment and never caught on with biologists, I also forgot much of it.
+Even if I forgot the programming syntax of these languages, what these programming experiences taught me was the logic and structure of programming.
+As a postdoc, I would use these experiences to learn the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [@Singleton2001].
+After writing my own version of LIBSHUFF, $\int$-LIBSHUFF, and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
 At the time, the conversion from Perl to C++ seemed like an academic exercise to learn a new language.
@@
 Was that really such a big difference?
-In hindsight, as we now process datasets with millions of sequences, the decision to learn to C++ was critical.
-The ability to pick up computer languages to solve problems was enabled by my prior training.
-It was also a skill that was virtually unheard of in microbiology.
+In hindsight, as we now process datasets with millions of sequences, the decision to learn C++ was critical.
+The ability to pick up computer languages to solve problems, enabled by my prior training in engineering, was a skill that was virtually unheard of in microbiology.
 Today, researchers without the ability to program are at a significant disadvantage [@Carey2018].
@@
 
-Milling about at a poster session at an ASM General Meeting in New Orleans, I ran into Mitch who asked what my plans were for my new lab.
+Milling about at a poster session at the 2007 ASM General Meeting in Toronto, I ran into Mitch who asked what my plans were for my new lab.
 I told him that I wanted to make a tool like ARB, a powerful database tool and phylogenetics package [@Ludwig2004], but for microbial ecology analysis.
@@
 There are two primary downsides to these tools.
-First, researchers running the online tool must pay the computational expenses leading to slow process times when hardware becomes outdated and when numerous users simultaneously attempt their analysis.
+First, researchers running the online tool must pay the computational expenses.
+When their hardware becomes outdated because it is expensive to replace or maintain, processing times slow down.
 Eventually this limitation would result in the termination of the greengenes website.
@@
 Regardless, both programs are very popular.
-From my perspective, most of the differences are cosmetic.
-QIIME is effectively a bundle of scripts to run other developers' software.
+From my perspective, most of the differences are cosmetic (http://blog.mothur.org/2016/01/12/mothur-and-qiime/).
+To me the most meaningful difference between mothur and QIIME is the choice of algorithms used to cluster sequences into OTUs.
+QIIME's advocacy for open and closed-reference clustering and USEARCH or VSEARCH-based *de novo* clustering results in lower quality OTU assignments relative to the *de novo* clustering algorithms available within mothur [@Westcott2015; @Westcott2017].
+QIIME is set of wrapper scripts that help users to transition data between independent packages.
 For example, with QIIME (through version 1.9.1), it was even possible to run mothur through QIIME.
@@
 
-Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone software tools for sequence curation (e.g.
+Beyond the large packages like mothur and QIIME, there has been significant growth in standalone software tools for sequence curation (e.g.
 PyroNoise [@Quince2009], PANDAseq [@Masella2012], DADA2 [@Callahan2016]), chimera checking (e.g.
@@
 USEARCH [@Edgar2010], VSEARCH [@Rognes2016], Swarm [@Mah2015]).
-Where possible and when warrented, we have implemented many of these algorithms directly into mothur.
+Where possible and when warranted, we have implemented many of these algorithms directly into mothur.
 We have also used this diversity of methods to perform head-to-head comparisons.
@@
 We applied an objective metric, the Matthew's Correlation Coefficient (MCC), to evaluate numerous algorithms for clustering sequences into OTUs.
-By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing algorithms.
+By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing ones.
 Beyond evaluating clustering algorithms, we have also evaluated methods of denoising sequence data [@Kozich2013; @Schloss2011b;  @Schloss2016b], assessed reference alignments [@Schloss2009b;  @Schloss2010a], considered the importance of incorporating secondary structure information in alignments [@Schloss2012a], quantified the variation along the 16S rRNA gene [@Schloss2010a], and compared the statistical hypotheses tested by commonly used tools [@Schloss2008].
-We have embraced the competition and diversity of all methods being used to analyze amplicon data as an opportunity to identify the strengths and weaknesses of the methods in an attempt to make recommendations to other researchers.
+We have embraced the competition and diversity of all methods being used to analyze amplicon data.
+This competition forces us to identify the strengths and weaknesses of various methods so that we can make recommendations to other researchers.
 
@@
 We were content for mothur to be an improved combination of DOTUR and SONS and leverage existing tools for other steps.
-Yet, when we learned that the code for NAST, the algorithm behind the greengenes's aligner [@DeSantis2006b],  was not open source or publicly available.
-Similarly, although the SINA aligner was available through ARB and the SILVA website performed well it was closed source.
-Because the ARB implementation did not scale to large datasets, researchers were left to pay a processing fee to the SILVA website to align sequences.
-Thus, we realized that such an important tool needed to be opened to the community [@Schloss2009b].
-More recently, the rejection of closed source, commercial tools such as USEARCH can be seen by the broader adoption of its open source, free competitor, VSEARCH, within the microbial ecology community [@Edgar2010; @Rognes2016].
+Yet, we quickly appreciated the need for providing other steps in a sequence analysis pipeline to make other tools more accessible.
+This decision was motivated by learning that the code for greengenes's [@DeSantis2006b] and ARB/SILVA's aligners were not open source or publicly available.
+Thus, we realized that such an important functionality needed to be opened to the community [@Schloss2009b].
+More recently, the rejection of closed source, commercial tools can be seen by the broader adoption of open source alternatives.
+This has been the case with the rising popularity of VSEARCH over USEARCH within the microbial ecology community [@Edgar2010; @Rognes2016].
 Related to insuring that mothur's code is open source, our second core principle is that we maintain transparency to our users.
 Perhaps a user does not need to interrogate every line of code, but they need to understand what is happening.
-Many programs including online workflows encapsulate large elements of a pipeline in a single command.
+Many programs, including online workflows, encapsulate large elements of a pipeline in a single command.
 In contrast, mothur forces the user to specify each step of the pipeline.
-Although the former approach makes an analysis easier for a beginner, it potentially stifles users that need greater control or understanding of the assumptions at each step.
+Although the former approach makes an analysis easier for a beginner, it stifles users that need greater control or understanding of the assumptions at each step.
 This control over the pipeline has made it easier for researchers to customize databases or adapt the pipeline to analyze non-16S rRNA gene sequence data.
+Furthermore, we have provided ample instructional materials to teach users how to implement robust pipelines and the theory behind each step.
 Third, as I mentioned above, there has been a plethora of methods proposed for generating amplicon sequence data, and curating, aligning, checking for chimeras, classifying, and clustering the data.
 I am proud of the data-driven approach we have taken to comparing these methods.
-A description of a new method is of limited value if it is not benchmarked against other methods or control datasets.
+A description of a new method has limited value if it is not benchmarked against other methods or control datasets.
 Through this core principle and mothur's large reach into the community, we have helped to develop standards in the analysis of 16S rRNA gene sequence data.
@@
 From the beginning, mothur's logfiles have represented a transcript of the user's command and outputs.
-When researchers were reluctant to submit sequence data to the Sequence Read Archive (SRA), we worked with the SRA developers to create a mothur command (make.sra) to create the package to submit sequence data through a special mothur portal.
+When it became clear that researchers were not submitting their sequence data to the Sequence Read Archive (SRA), we worked with the SRA developers to create a mothur command (make.sra) that creates a package for submitting sequence data through a special mothur portal.
 A more ambitious project had its seed on April 1, 2013 when we announced a new "function" in mothur: write.paper.
@@
 Perhaps the most important core principle is that my research group uses mothur to analyze the data we generate.
-We "eat our own dogfood".
 This has proven critical as it again represents transparency and hopefully provides confidence to mothur's users that we are not making recommendations that we do not follow ourselves.
@@
 One challenge we face is that if we looked to third parties to contribute code to mothur, they would need to know C++.
-Given the paucity of microbiologists with skills programing in a compiled language like C++, expecting that community to provide contributors that can write code in a syntax that prizes execution efficiency over developer efficiency was not likely.
+Given the paucity of microbiologists that can program in a compiled language like C++, expecting that community to provide contributors who can write code in a syntax that prizes execution efficiency over developer efficiency was not realistic.
 In contrast, the QIIME development team could be more distributed because their code base was primarily written in Python, which prizes developer efficiency over execution efficiency.
-QIIME is a series of wrappers that allow users to execute other developers' code making the use of a scripting language like Python attractive.
+QIIME is a series of wrappers that allow users to execute other developers' code, making the use of a scripting language like Python attractive.
 Their choices resulted in many tradeoffs that have impacted ease of installation, usability, execution speed, and flexibility.
@@
 For example, the modern version of the Rcpp package was first released in 2009 and its popularity was not immediate [@Eddelbuettel2011].
-Again, the development of mothur has been a product of the environment that it was created in.
+The development of mothur has been a product of the environment that it was created in.
 Although these decisions have largely had positive outcomes, there have been tradeoffs that caused us to sacrifice other goals.
@@
 Beyond contributing to the mothur code base, we sought out other ways to include the community as developers.
-The paper describing mothur included 15 co-authors, all but three (Schloss, Westcott, and Ryabin) responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
-Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for other seeking to adapt mothur for their uses (https://www.mothur.org/wiki).
+The paper describing mothur included 15 co-authors, most of whom responded to a call to provide a wiki page that described how they used an early version of mothur to analyze a data set.
+Our vision was that authors might use the mothur wiki to document reproducible workflows for papers using mothur but to also provide instructional materials for others seeking to adapt mothur for their uses (https://www.mothur.org/wiki).
 Unfortunately, once the incentive of co-authorship was removed, researchers stopped contributing their workflows to the wiki.
@@
 One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and Venn diagrams depicting the overlap between microbial communities.
-Such visuals were helpful for exploring or data; however, I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote.
+Such visuals were helpful for exploring data; however, I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote.
 Such visuals require far too much customization to be publication-quality.
-Although QIIME has incorporated visualization tools through the Emperor package [@VzquezBaeza2013], the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature.
+Although QIIME has incorporated visualization tools through the Emperor package [@VzquezBaeza2013], the challenge of users taking default values has downsides, especially when those defaults do not follow good data visualization principles.
+For example, ordinations with black backgrounds and 3-D ordinations in a 2-D medium now litter the literature.
 Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials that I have developed (http://www.riffomonas.org/minimalR/).
@@
 Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks.
-Documentation of commands and parameter values is explicit in mothur as users can provide the software a file with a list of commands and all commands and output are recorded in a logfile.
-Given the heightened focus on reproducibility in recent years [@Schloss2018b], we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project [@Schloss2018a].
+Documentation of commands and parameter values is explicit in users can provide a file with a list of commands and the software returns a logfile with all commands and output recorded.
+Given the heightened focus on reproducibility in recent years, we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript [@Schloss2018a; @Schloss2018b].
 A final example of a failed experiment was a collaboration with programmers through Google Summer of Code to develop commands in mothur that ran the random forest and SVM machine learning algorithms.
@@
 Thankfully, there are numerous R and Python packages that do a better job of developing these models [@Paszke2017; @Kuhn2008].
-Again, we have put our efforts into developing instructional materials that mothur users can use to fit such models to their data.
 In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.
@@
 It is important to note that the two methods answer fundamentally different questions.
-16S rRNA gene sequence data describes the taxonomic composition and metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community.
-Both tools provide important information but they cannot easily replace each other.
+16S rRNA gene sequence data describes the taxonomic composition, while metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community.
+Both tools provide important information, but they cannot easily replace each other.
 Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is at least an order of magnitude higher than that of amplicon data.
@@
 Furthermore, analyzing such a large dataset with an approach that captures the full genetic diversity of the community would be financially and technically prohibitive.
-Going forward, there is still a place for 16S rRNA gene sequencing.
-Although sequencing technologies will evolve to capture longer and more high quality data, there will likely always be a need for characterizing the taxonomic diversity of microbial communities.
+Going forward, sequencing technologies will continue to evolve to capture longer and more high quality data and there will always be a need for characterizing the taxonomic diversity of microbial communities.
 With this in mind, there will always be a place for tools like mothur that can analyze amplicon sequence data.
@@
 Second, with these improvements, we will need to continue to improve our algorithms.
-We have already seen that attempts to use low quality MiSeq and HiSeq data causes computational problems leading to the creation of open and closed reference clustering methods [@NavasMolina2013; @Rideout2014].
-Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods [@Westcott2015].
+We have already seen that attempts to use low quality MiSeq and HiSeq data caused computational problems leading to the creation of open and closed reference clustering methods, which attempted to circumvent those problems [@NavasMolina2013; @Rideout2014].
+Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods [@Westcott2015; @Westcott2017].
 More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality.
@@
 These include the validity and utility of amplicon sequence variants [@Callahan2017], the wisdom of removing low frequency sequences [@Bokulich2012], and methods of identifying and removing contaminant 16S rRNA gene sequences [@Salter2014; @Davis2018].
-With each of these areas of development, the broader community can count on our same data-driven approach to meeting evaluating these questions.
+With each of these areas of development, the broader community can count on our same data-driven approach to answer these questions.
 It is common for researchers to comment that they pick a specific method or deviate from a suggestion because they "like how the data look".
 When pressed for an objective definition of how they know the data look "right", they go quiet.
-Through the use of mock communities and simulations where we actually know what looks right and objective metrics of quality like the MCC or sequencing error rates, we will continue to base recommendations on data rather than a gut feeling.
+Through the use of data where we actually know what looks right and objective metrics of quality, we will continue to base recommendations on data rather than a gut feeling.
 
@@
 Second, as we have developed mothur, we have attempted to do so in a data-driven approach where we compare multiple methods.
-It has not merely been enough to propose a new method, we must show that it meaningfully advances the field.
+It has not merely been enough to propose a new method: we must show that it meaningfully advances the field.
 Third, through our failures and successes we have learned to focus on what mothur is good at and create products separate from mothur when distinct needs arise.
-For example, we have learned that mothur should not have a graphical interface or data visualization tools.
-Instead, we will provide instructional materials to teach users how to use the command line interface and other programs like R for data visualization.
+For example, we have learned that mothur should not have a graphical interface or data visualization tool.
+Instead, we provide instructional materials to teach users how to use the command line interface and other computational skills like programming in R for data visualization.
 Finally, mothur was born out of a need for automating the analysis of large 16S rRNA gene sequence datasets.
-I realized I had a set of skills to fill that need.
 It has been refreshing to see the computational skills of the microbial ecology field grow over the past two decades.
@@
 The development of mothur would not be possible without the contributions of its many supporters, developers, and users.
-Although not a complete list, I would be remiss if I did not express my gratitude to Sarah Westcott and the other members of my research group, Jo Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra Mass, and Marcy Balunas for their unique contributions to the continued development of mothur.
+Although not a complete list, I would be remiss if I did not express my gratitude to Sarah Westcott and the other members of my research group, Jo Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra Maas, and Marcy Balunas for their unique contributions to the continued development of mothur.
 
@@
 **Figure 1.
-mothur has consistently been a popular software package over the past ten years with a total of more than 8,800 citations.** Citation data taken from the Web of Science (https://www.webofscience.com) on October 1, 2019.
-Data from 2019 are incomplete.
+mothur has consistently been a popular software package over the past ten years with more than 8,800 citations.** Citation data taken from the Web of Science (https://www.webofscience.com) on October 1, 2019.
+The gray line segment depicts the projected number of citations for 2019 based on the current number of citations for the year and historical trends.
 

### 2019-10-09 submission/manuscript.tex
Incorporate MB / KS comments

@@
 allowed users to analyze amplicon sequence data using the most robust
-methods available. Looking back, it is clear that mothur continues to
-provide this service to the microbial ecology community as it has helped
-to lead the community through the ongoing sequencing revolution. Beyond
-its success and impact on the field, mothur's development exposed a
-series of observations that are generally translatable across science.
-Although it is easy to criticize choices that were made 10 years ago
-through a modern lens, all science is done in the context of prevailing
-ideas and available technologies. If we were to wait for all of the
-possible limitations to be solved before proceeding, science would
-stall. Even preceding the development of mothur, it was necessary to
-address the most important problems and work backwards to other problems
-that limited access to robust sequence analysis tools. At the same time,
-we strive to expand mothur's capabilities in a data-driven manner to
-incorporate new ideas and accommodate changes in data and desires of the
-research community. As we develop mothur, we first consider the needs of
-my research group. In hindsight, we appreciate that our needs are
-representative of the broader community's. It has been edifying to see
-the benefit that a simple set of tools can bring to so many other
-researchers.
+methods available. mothur has helped lead the community through the
+ongoing sequencing revolution and continues to provide this service to
+the microbial ecology community. Beyond its success and impact on the
+field, mothur's development exposed a series of observations that are
+generally translatable across science. Perhaps the observation that
+stands out the most is that all science is done in the context of
+prevailing ideas and available technologies. Although it is easy to
+criticize choices that were made 10 years ago through a modern lens, if
+we were to wait for all of the possible limitations to be solved before
+proceeding, science would stall. Even preceding the development of
+mothur, it was necessary to address the most important problems and work
+backwards to other problems that limited access to robust sequence
+analysis tools. At the same time, we strive to expand mothur's
+capabilities in a data-driven manner to incorporate new ideas and
+accommodate changes in data and desires of the research community. It
+has been edifying to see the benefit that a simple set of tools can
+bring to so many other researchers.
 
@@
 overwhelmed at the prospect of looking forward at their careers (1--5).
-By no means is my personal journey over, but since 2002 I have been on a
-journey that I did not realize I was on. Now that the paper introducing
-the mothur software package is ten years old and has become the most
-cited paper published by \emph{Applied and Environmental Microbiology}
-(7), it is worth stepping back and using the continued development of
-mothur as a story that has parallels to many other research stories.
+By no means is my scientific journey over, but since 2002 I have been on
+a journey that I did not realize I was on. Now that the paper
+introducing the mothur software package is ten years old and has become
+the most cited paper published by \emph{Applied and Environmental
+Microbiology} (7), it is worth stepping back and using the continued
+development of mothur as a story that has parallels to many other
+research stories.
 
@@
 trivial, but by cobbling together other tools it was possible. We would
-assemble, trim and and correct Sanger sequence reads using Chromas or
-STADEN (28), align the sequences using ClustalW (29) or ARB (30), check
-for chimeras using partial treeing or Bellerephon (27), and calculate a
+assemble, trim and correct Sanger sequence reads using Chromas or STADEN
+(28), align the sequences using ClustalW (29) or ARB (30), check for
+chimeras using partial treeing or Bellerephon (27), and calculate a
 pairwise distance matrix using DNADIST from the PHYLIP package (31). At
@@
 clusterings to test our hypotheses. The only tool available at the time
-that automated the clustering step was FastGroup, which implemented a
-greedy version of the single linkage algorithm (33). The high cost of
+that automated the clustering step was FastGroup, which implemented an
+approximation of the single linkage algorithm (33). The high cost of
 sequencing was also an impediment to experimentation and analysis in
@@
 treatment groups. For example, in our testing we frequently used a
-dataset comparing Scottish soils from from Alison McCaig and colleagues
-(34). This dataset consisted of two experimental groups, each replicated
-three times with 45 sequences per replicate. Although great focus has
-been placed on the depth of sampling afforded by 454 and Illumina
-sequencing, the true benefit of the modern sequencing platforms is the
-ability to affordably sequence a large number of technical and
-biological replicates. In spite of the many technical challenges, we had
-excuses and heuristics to solve problems that served our needs. It is
-telling that a recent review of ``best practices'' in generating and
-analyzing 16S rRNA gene sequences shows that we still have not solved
-many of these issues and that we have even identified additional
-problems (35).
+dataset comparing Scottish soils from Alison McCaig and colleagues (34).
+This dataset consisted of two experimental groups, each replicated three
+times with 45 sequences per replicate. Although great focus has been
+placed on the depth of sampling afforded by 454 and Illumina sequencing,
+the true benefit of the modern sequencing platforms is the ability to
+affordably sequence a large number of technical and biological
+replicates. In spite of the many technical challenges, we had excuses
+and heuristics to solve problems that served our needs. It is telling
+that a recent review of ``best practices'' in generating and analyzing
+16S rRNA gene sequences shows that we still have not solved many of
+these issues and that we have even identified additional problems (35).
 
@@
 to help solve problems. As an undergraduate student, I learned the
-Pascal programming language and promptly forgot much of it as a graduate
-student and learned MatLab in its place. As a postdoc, I learned the
-Perl programming language to better understand how LIBSHUFF, a tool for
-comparing the structure of two communities, worked since it was written
-in Perl (36). After writing my own version of LIBSHUFF,
-\(\int\)-LIBSHUFF and seeing the speed of the version written in C++ by
-my collaborator, Bret Larget, I converted my Perl version of DOTUR into
-C++. At the time, the conversion from Perl to C++ seemed like an
-academic exercise to learn a new language. My Perl version only took a
-minute or so to process the final collection of 1,000 sequences and the
-C++ version took seconds. Was that really such a big difference? In
-hindsight, as we now process datasets with millions of sequences, the
-decision to learn to C++ was critical. The ability to pick up computer
-languages to solve problems was enabled by my prior training. It was
-also a skill that was virtually unheard of in microbiology. Today,
-researchers without the ability to program are at a significant
-disadvantage (37).
+Pascal programming language and promptly forgot much of it. Although it
+was a good language for teaching programming concepts, it did not catch
+on outside of the classroom. Later, I learned MATLAB. Because it was an
+expensive commercial programming environment and never caught on with
+biologists, I also forgot much of it. Even if I forgot the programming
+syntax of these languages, what these programming experiences taught me
+was the logic and structure of programming. As a postdoc, I would use
+these experiences to learn the Perl programming language to better
+understand how LIBSHUFF, a tool for comparing the structure of two
+communities, worked since it was written in Perl (36). After writing my
+own version of LIBSHUFF, \(\int\)-LIBSHUFF, and seeing the speed of the
+version written in C++ by my collaborator, Bret Larget, I converted my
+Perl version of DOTUR into C++. At the time, the conversion from Perl to
+C++ seemed like an academic exercise to learn a new language. My Perl
+version only took a minute or so to process the final collection of
+1,000 sequences and the C++ version took seconds. Was that really such a
+big difference? In hindsight, as we now process datasets with millions
+of sequences, the decision to learn C++ was critical. The ability to
+pick up computer languages to solve problems, enabled by my prior
+training in engineering, was a skill that was virtually unheard of in
+microbiology. Today, researchers without the ability to program are at a
+significant disadvantage (37).
 
@@
 
-Milling about at a poster session at an ASM General Meeting in New
-Orleans, I ran into Mitch who asked what my plans were for my new lab. I
+Milling about at a poster session at the 2007 ASM General Meeting in
+Toronto, I ran into Mitch who asked what my plans were for my new lab. I
 told him that I wanted to make a tool like ARB, a powerful database tool
@@
 tools. First, researchers running the online tool must pay the
-computational expenses leading to slow process times when hardware
-becomes outdated and when numerous users simultaneously attempt their
-analysis. Eventually this limitation would result in the termination of
-the greengenes website. Second, these platforms provide a
-one-size-fits-all analysis. These tools only allow a user to analyze 16S
-and in some cases 18S rRNA gene sequences. If a user sequences a
-different gene, then the tool will not serve them. These observations
-resulted in two design goals we have had with mothur: bringing the
-analysis to a user's computer and separating a tool from a specific
-database. For example, we commonly use a sequence alignment method that
-was originally developed for greengenes (42), but use a SILVA-based
-reference alignment because its superior quality (43, 44). In addition,
-we offer the \text{na\"ive} Bayesian classifier developed by the RDP
-(45) and allow users to train it to any database they want, including
-customized databases. In both examples, users can align or classify
-non-rRNA gene sequence data. As the bioinformatics tools have matured,
-both the RDP and SILVA offer integrated pipelines for analyzing large
-datasets, albeit in one-size-fits-all black box implementations.
+computational expenses. When their hardware becomes outdated because it
+is expensive to replace or maintain, processing times slow down.
+Eventually this limitation would result in the termination of the
+greengenes website. Second, these platforms provide a one-size-fits-all
+analysis. These tools only allow a user to analyze 16S and in some cases
+18S rRNA gene sequences. If a user sequences a different gene, then the
+tool will not serve them. These observations resulted in two design
+goals we have had with mothur: bringing the analysis to a user's
+computer and separating a tool from a specific database. For example, we
+commonly use a sequence alignment method that was originally developed
+for greengenes (42), but use a SILVA-based reference alignment because
+its superior quality (43, 44). In addition, we offer the \text{na\"ive}
+Bayesian classifier developed by the RDP (45) and allow users to train
+it to any database they want, including customized databases. In both
+examples, users can align or classify non-rRNA gene sequence data. As
+the bioinformatics tools have matured, both the RDP and SILVA offer
+integrated pipelines for analyzing large datasets, albeit in
+one-size-fits-all black box implementations.
 
@@
 complement or an insult. Regardless, both programs are very popular.
-From my perspective, most of the differences are cosmetic. QIIME is
-effectively a bundle of scripts to run other developers' software. For
-example, with QIIME (through version 1.9.1), it was even possible to run
-mothur through QIIME. One can also run the \text{na\"ive} Bayesian
-classifier through QIIME using the original code developed by the RDP.
-This caused great frustration for many users because there were numerous
-software dependencies that had to be installed. Although the QIIME
-developers would go on to create virtual machines and use packaging
-tools to simplify installation, these fixes required sophistication by
-users who we knew struggled with the basics of navigating a command
-line. In contrast, when a user runs mothur, they are running mothur. The
+From my perspective, most of the differences are cosmetic
+(\url{http://blog.mothur.org/2016/01/12/mothur-and-qiime/}). To me the
+most meaningful difference between mothur and QIIME is the choice of
+algorithms used to cluster sequences into OTUs. QIIME's advocacy for
+open and closed-reference clustering and USEARCH or VSEARCH-based
+\emph{de novo} clustering results in lower quality OTU assignments
+relative to the \emph{de novo} clustering algorithms available within
+mothur (47, 48). QIIME is set of wrapper scripts that help users to
+transition data between independent packages. For example, with QIIME
+(through version 1.9.1), it was even possible to run mothur through
+QIIME. One can also run the \text{na\"ive} Bayesian classifier through
+QIIME using the original code developed by the RDP. This caused great
+frustration for many users because there were numerous software
+dependencies that had to be installed. Although the QIIME developers
+would go on to create virtual machines and use packaging tools to
+simplify installation, these fixes required sophistication by users who
+we knew struggled with the basics of navigating a command line. In
+contrast, when a user runs mothur, they are running mothur. The
 \text{na\"ive} Bayesian classifier code that is in mothur is a rewritten
@@
 rRNA gene sequence data. For example, while 454 data was popular,
-PyroNoise was an effective tool for denoising flowgram data (47).
+PyroNoise was an effective tool for denoising flowgram data (49).
 Running the original code required a large Linux computer cluster and
@@
 Beyond the large packages like mothur and QIIME, there has been
-significant growth in stand alone software tools for sequence curation
-(e.g.~PyroNoise (47), PANDAseq (48), DADA2 (49)), chimera checking
-(e.g.~UCHIME (50), ChimeraSlayer (51), Perseus (52)), and clustering
-(e.g.~USEARCH (53), VSEARCH (54), Swarm (55)). Where possible and when
-warrented, we have implemented many of these algorithms directly into
+significant growth in standalone software tools for sequence curation
+(e.g.~PyroNoise (49), PANDAseq (50), DADA2 (51)), chimera checking
+(e.g.~UCHIME (52), ChimeraSlayer (53), Perseus (54)), and clustering
+(e.g.~USEARCH (55), VSEARCH (56), Swarm (57)). Where possible and when
+warranted, we have implemented many of these algorithms directly into
 mothur. We have also used this diversity of methods to perform
@@
 algorithms where there have been a large number of algorithms developed
-without an obvious method to objectively compare them (56--59). We
-applied an objective metric, the Matthew's Correlation Coefficient
+without an obvious method to objectively compare them (47, 48, 58, 59).
+We applied an objective metric, the Matthew's Correlation Coefficient
 (MCC), to evaluate numerous algorithms for clustering sequences into
@@
 compare the algorithms, make recommendations to the field, and develop
-new algorithms that outperformed the existing algorithms. Beyond
-evaluating clustering algorithms, we have also evaluated methods of
-denoising sequence data (60--62), assessed reference alignments (43,
-44), considered the importance of incorporating secondary structure
+new algorithms that outperformed the existing ones. Beyond evaluating
+clustering algorithms, we have also evaluated methods of denoising
+sequence data (60--62), assessed reference alignments (43, 44),
+considered the importance of incorporating secondary structure
 information in alignments (63), quantified the variation along the 16S
@@
 commonly used tools (64). We have embraced the competition and diversity
-of all methods being used to analyze amplicon data as an opportunity to
-identify the strengths and weaknesses of the methods in an attempt to
-make recommendations to other researchers.
+of all methods being used to analyze amplicon data. This competition
+forces us to identify the strengths and weaknesses of various methods so
+that we can make recommendations to other researchers.
 
@@
 content for mothur to be an improved combination of DOTUR and SONS and
-leverage existing tools for other steps. Yet, when we learned that the
-code for NAST, the algorithm behind the greengenes's aligner (42), was
-not open source or publicly available. Similarly, although the SINA
-aligner was available through ARB and the SILVA website performed well
-it was closed source. Because the ARB implementation did not scale to
-large datasets, researchers were left to pay a processing fee to the
-SILVA website to align sequences. Thus, we realized that such an
-important tool needed to be opened to the community (43). More recently,
-the rejection of closed source, commercial tools such as USEARCH can be
-seen by the broader adoption of its open source, free competitor,
-VSEARCH, within the microbial ecology community (53, 54). Related to
-insuring that mothur's code is open source, our second core principle is
-that we maintain transparency to our users. Perhaps a user does not need
-to interrogate every line of code, but they need to understand what is
-happening. Many programs including online workflows encapsulate large
-elements of a pipeline in a single command. In contrast, mothur forces
-the user to specify each step of the pipeline. Although the former
-approach makes an analysis easier for a beginner, it potentially stifles
-users that need greater control or understanding of the assumptions at
-each step. This control over the pipeline has made it easier for
-researchers to customize databases or adapt the pipeline to analyze
-non-16S rRNA gene sequence data. Third, as I mentioned above, there has
-been a plethora of methods proposed for generating amplicon sequence
-data, and curating, aligning, checking for chimeras, classifying, and
-clustering the data. I am proud of the data-driven approach we have
-taken to comparing these methods. A description of a new method is of
-limited value if it is not benchmarked against other methods or control
-datasets. Through this core principle and mothur's large reach into the
-community, we have helped to develop standards in the analysis of 16S
-rRNA gene sequence data. Fourth, a focus on enabling reproducibility has
-always been central to the functionality of mothur. From the beginning,
-mothur's logfiles have represented a transcript of the user's command
-and outputs. When researchers were reluctant to submit sequence data to
+leverage existing tools for other steps. Yet, we quickly appreciated the
+need for providing other steps in a sequence analysis pipeline to make
+other tools more accessible. This decision was motivated by learning
+that the code for greengenes's (42) and ARB/SILVA's aligners were not
+open source or publicly available. Thus, we realized that such an
+important functionality needed to be opened to the community (43). More
+recently, the rejection of closed source, commercial tools can be seen
+by the broader adoption of open source alternatives. This has been the
+case with the rising popularity of VSEARCH over USEARCH within the
+microbial ecology community (55, 56). Related to insuring that mothur's
+code is open source, our second core principle is that we maintain
+transparency to our users. Perhaps a user does not need to interrogate
+every line of code, but they need to understand what is happening. Many
+programs, including online workflows, encapsulate large elements of a
+pipeline in a single command. In contrast, mothur forces the user to
+specify each step of the pipeline. Although the former approach makes an
+analysis easier for a beginner, it stifles users that need greater
+control or understanding of the assumptions at each step. This control
+over the pipeline has made it easier for researchers to customize
+databases or adapt the pipeline to analyze non-16S rRNA gene sequence
+data. Furthermore, we have provided ample instructional materials to
+teach users how to implement robust pipelines and the theory behind each
+step. Third, as I mentioned above, there has been a plethora of methods
+proposed for generating amplicon sequence data, and curating, aligning,
+checking for chimeras, classifying, and clustering the data. I am proud
+of the data-driven approach we have taken to comparing these methods. A
+description of a new method has limited value if it is not benchmarked
+against other methods or control datasets. Through this core principle
+and mothur's large reach into the community, we have helped to develop
+standards in the analysis of 16S rRNA gene sequence data. Fourth, a
+focus on enabling reproducibility has always been central to the
+functionality of mothur. From the beginning, mothur's logfiles have
+represented a transcript of the user's command and outputs. When it
+became clear that researchers were not submitting their sequence data to
 the Sequence Read Archive (SRA), we worked with the SRA developers to
-create a mothur command (make.sra) to create the package to submit
+create a mothur command (make.sra) that creates a package for submitting
 sequence data through a special mothur portal. A more ambitious project
@@
 principle is that my research group uses mothur to analyze the data we
-generate. We ``eat our own dogfood''. This has proven critical as it
-again represents transparency and hopefully provides confidence to
-mothur's users that we are not making recommendations that we do not
-follow ourselves.
+generate. This has proven critical as it again represents transparency
+and hopefully provides confidence to mothur's users that we are not
+making recommendations that we do not follow ourselves.
 
@@
 to contribute code to mothur, they would need to know C++. Given the
-paucity of microbiologists with skills programing in a compiled language
-like C++, expecting that community to provide contributors that can
-write code in a syntax that prizes execution efficiency over developer
-efficiency was not likely. In contrast, the QIIME development team could
-be more distributed because their code base was primarily written in
-Python, which prizes developer efficiency over execution efficiency.
-QIIME is a series of wrappers that allow users to execute other
-developers' code making the use of a scripting language like Python
-attractive. Their choices resulted in many tradeoffs that have impacted
-ease of installation, usability, execution speed, and flexibility. If we
-were offered funding to rewrite mothur, we would likely rewrite it as an
-R package that leaned heavily on the R language's C++ interface
-packages. Of course, such choices are always best in hindsight. Yet,
-when we started developing mothur, the ability to interface between
-scripting languages like R and Python and C++ code was not as well
-developed as it is today. For example, the modern version of the Rcpp
-package was first released in 2009 and its popularity was not immediate
-(66). Again, the development of mothur has been a product of the
-environment that it was created in. Although these decisions have
-largely had positive outcomes, there have been tradeoffs that caused us
-to sacrifice other goals.
+paucity of microbiologists that can program in a compiled language like
+C++, expecting that community to provide contributors who can write code
+in a syntax that prizes execution efficiency over developer efficiency
+was not realistic. In contrast, the QIIME development team could be more
+distributed because their code base was primarily written in Python,
+which prizes developer efficiency over execution efficiency. QIIME is a
+series of wrappers that allow users to execute other developers' code,
+making the use of a scripting language like Python attractive. Their
+choices resulted in many tradeoffs that have impacted ease of
+installation, usability, execution speed, and flexibility. If we were
+offered funding to rewrite mothur, we would likely rewrite it as an R
+package that leaned heavily on the R language's C++ interface packages.
+Of course, such choices are always best in hindsight. Yet, when we
+started developing mothur, the ability to interface between scripting
+languages like R and Python and C++ code was not as well developed as it
+is today. For example, the modern version of the Rcpp package was first
+released in 2009 and its popularity was not immediate (66). The
+development of mothur has been a product of the environment that it was
+created in. Although these decisions have largely had positive outcomes,
+there have been tradeoffs that caused us to sacrifice other goals.
 
@@
 include the community as developers. The paper describing mothur
-included 15 co-authors, all but three (Schloss, Westcott, and Ryabin)
-responded to a call to provide a wiki page that described how they used
-an early version of mothur to analyze a data set. Our vision was that
-authors might use the mothur wiki to document reproducible workflows for
-papers using mothur but to also provide instructional materials for
-other seeking to adapt mothur for their uses
-(\url{https://www.mothur.org/wiki}). Unfortunately, once the incentive
-of co-authorship was removed, researchers stopped contributing their
-workflows to the wiki. Again, this vision and the lack of the
-community's adoption of wikis as a mechanism for reporting workflows was
-a product of the environment. Although wikis were popular in the late
-2000's, they lacked the ability to directly execute the commands that
-researchers reported. Such technology would not be possible until the
-creation of IPython notebooks (2011) and R markdown (2012). Another
-problem with the wiki approach was that potential contributors did not
-see the wiki as a community resource. I frequently received emails from
-scientists telling me that there was a typo on a specific page when the
-intention was that they could correct the typos without my input. We
-have been more successful in soliciting input and contributions from the
-user community through the mothur discussion forum and GitHub-based
-issue tracker. As mothur has matured, we have been dependent on the user
-community to use these resources to tell us what features they would
-like to see included in mothur and where the documentation is confusing
-or incomplete (\url{https://forum.mothur.org}). Often we can count on
-people not directly affiliated with mothur to provide instruction and
-their own experience to other users on the forum. We are constantly
-trying to recruit our ``army'' and are happy to take any contributions
-we can. Whether the contributions are to the code base, discussion
-forum, or suggestions for new tools, these contributions have been
-invaluable to the growth and popularity of mothur.
+included 15 co-authors, most of whom responded to a call to provide a
+wiki page that described how they used an early version of mothur to
+analyze a data set. Our vision was that authors might use the mothur
+wiki to document reproducible workflows for papers using mothur but to
+also provide instructional materials for others seeking to adapt mothur
+for their uses (\url{https://www.mothur.org/wiki}). Unfortunately, once
+the incentive of co-authorship was removed, researchers stopped
+contributing their workflows to the wiki. Again, this vision and the
+lack of the community's adoption of wikis as a mechanism for reporting
+workflows was a product of the environment. Although wikis were popular
+in the late 2000's, they lacked the ability to directly execute the
+commands that researchers reported. Such technology would not be
+possible until the creation of IPython notebooks (2011) and R markdown
+(2012). Another problem with the wiki approach was that potential
+contributors did not see the wiki as a community resource. I frequently
+received emails from scientists telling me that there was a typo on a
+specific page when the intention was that they could correct the typos
+without my input. We have been more successful in soliciting input and
+contributions from the user community through the mothur discussion
+forum and GitHub-based issue tracker. As mothur has matured, we have
+been dependent on the user community to use these resources to tell us
+what features they would like to see included in mothur and where the
+documentation is confusing or incomplete
+(\url{https://forum.mothur.org}). Often we can count on people not
+directly affiliated with mothur to provide instruction and their own
+experience to other users on the forum. We are constantly trying to
+recruit our ``army'' and are happy to take any contributions we can.
+Whether the contributions are to the code base, discussion forum, or
+suggestions for new tools, these contributions have been invaluable to
+the growth and popularity of mothur.
 
@@
 overlap between microbial communities. Such visuals were helpful for
-exploring or data; however, I quickly realized that I would never put a
+exploring data; however, I quickly realized that I would never put a
 mothur-generated figure into a manuscript I wrote. Such visuals require
@@
 incorporated visualization tools through the Emperor package (67), the
-challenge of users taking default values has downsides as ordinations
-with black background or publishing 3-D ordinations in a 2-D medium
-litter the literature. Instead, we have encouraged users to use R
-packages to visualize mothur-generated results using the minimalR
+challenge of users taking default values has downsides, especially when
+those defaults do not follow good data visualization principles. For
+example, ordinations with black backgrounds and 3-D ordinations in a 2-D
+medium now litter the literature. Instead, we have encouraged users to
+use R packages to visualize mothur-generated results using the minimalR
 instructional materials that I have developed
@@
 users do not have a mechanism to document their mouse clicks.
-Documentation of commands and parameter values is explicit in mothur as
-users can provide the software a file with a list of commands and all
-commands and output are recorded in a logfile. Given the heightened
-focus on reproducibility in recent years (70), we have extended
-significant effort in developing instructional materials teaching users
-how to organize, document, and execute reproducible pipelines that allow
-a user to go from raw sequence data to a compiled manuscript with
-figures through the Riffomonas project (65). A final example of a failed
-experiment was a collaboration with programmers through Google Summer of
-Code to develop commands in mothur that ran the random forest and SVM
-machine learning algorithms. Similar to the challenges of developing
-attractive visuals, fitting the algorithms' hyperparameters, testing,
-and deploying the resulting models require a significant amount of
-customization. Furthermore, machine learning is an active area of
-research where methods are still being developed and improved.
-Thankfully, there are numerous R and Python packages that do a better
-job of developing these models (71, 72). Again, we have put our efforts
-into developing instructional materials that mothur users can use to fit
-such models to their data. In each of our ``failed'' experiments, the
-real problems were straying from what mothur does well and failing to
-grasp what we really wanted the innovation to do.
+Documentation of commands and parameter values is explicit in users can
+provide a file with a list of commands and the software returns a
+logfile with all commands and output recorded. Given the heightened
+focus on reproducibility in recent years, we have extended significant
+effort in developing instructional materials teaching users how to
+organize, document, and execute reproducible pipelines that allow a user
+to go from raw sequence data to a compiled manuscript (65, 70). A final
+example of a failed experiment was a collaboration with programmers
+through Google Summer of Code to develop commands in mothur that ran the
+random forest and SVM machine learning algorithms. Similar to the
+challenges of developing attractive visuals, fitting the algorithms'
+hyperparameters, testing, and deploying the resulting models require a
+significant amount of customization. Furthermore, machine learning is an
+active area of research where methods are still being developed and
+improved. Thankfully, there are numerous R and Python packages that do a
+better job of developing these models (71, 72). In each of our
+``failed'' experiments, the real problems were straying from what mothur
+does well and failing to grasp what we really wanted the innovation to
+do.
 
@@
 the two methods answer fundamentally different questions. 16S rRNA gene
-sequence data describes the taxonomic composition and metagenomic
+sequence data describes the taxonomic composition, while metagenomic
 sequence data tells a researcher about the functional potential and
 genetic diversity of a community. Both tools provide important
-information but they cannot easily replace each other. Although
+information, but they cannot easily replace each other. Although
 metagenomic data does provide highly resolved taxonomic information, the
@@
 approach that captures the full genetic diversity of the community would
-be financially and technically prohibitive. Going forward, there is
-still a place for 16S rRNA gene sequencing. Although sequencing
-technologies will evolve to capture longer and more high quality data,
-there will likely always be a need for characterizing the taxonomic
-diversity of microbial communities. With this in mind, there will always
-be a place for tools like mothur that can analyze amplicon sequence
-data.
+be financially and technically prohibitive. Going forward, sequencing
+technologies will continue to evolve to capture longer and more high
+quality data and there will always be a need for characterizing the
+taxonomic diversity of microbial communities. With this in mind, there
+will always be a place for tools like mothur that can analyze amplicon
+sequence data.
 
@@
 algorithms. We have already seen that attempts to use low quality MiSeq
-and HiSeq data causes computational problems leading to the creation of
-open and closed reference clustering methods (75, 76). Unfortunately,
-comparative analyses showed that these methods fail relative to \emph{de
-novo} clustering methods (57). More work is needed to improve
-reference-based clustering methods so that larger datasets can be
-analyzed without sacrificing quality. Finally, there are ongoing
-controversies that need further exploration. These include the validity
-and utility of amplicon sequence variants (77), the wisdom of removing
-low frequency sequences (78), and methods of identifying and removing
-contaminant 16S rRNA gene sequences (79, 80). With each of these areas
-of development, the broader community can count on our same data-driven
-approach to meeting evaluating these questions. It is common for
-researchers to comment that they pick a specific method or deviate from
-a suggestion because they ``like how the data look''. When pressed for
-an objective definition of how they know the data look ``right'', they
-go quiet. Through the use of mock communities and simulations where we
-actually know what looks right and objective metrics of quality like the
-MCC or sequencing error rates, we will continue to base recommendations
-on data rather than a gut feeling.
+and HiSeq data caused computational problems leading to the creation of
+open and closed reference clustering methods, which attempted to
+circumvent those problems (75, 76). Unfortunately, comparative analyses
+showed that these methods fail relative to \emph{de novo} clustering
+methods (47, 48). More work is needed to improve reference-based
+clustering methods so that larger datasets can be analyzed without
+sacrificing quality. Finally, there are ongoing controversies that need
+further exploration. These include the validity and utility of amplicon
+sequence variants (77), the wisdom of removing low frequency sequences
+(78), and methods of identifying and removing contaminant 16S rRNA gene
+sequences (79, 80). With each of these areas of development, the broader
+community can count on our same data-driven approach to answer these
+questions. It is common for researchers to comment that they pick a
+specific method or deviate from a suggestion because they ``like how the
+data look''. When pressed for an objective definition of how they know
+the data look ``right'', they go quiet. Through the use of data where we
+actually know what looks right and objective metrics of quality, we will
+continue to base recommendations on data rather than a gut feeling.
 
@@
 data-driven approach where we compare multiple methods. It has not
-merely been enough to propose a new method, we must show that it
+merely been enough to propose a new method: we must show that it
 meaningfully advances the field. Third, through our failures and
@@
 have learned that mothur should not have a graphical interface or data
-visualization tools. Instead, we will provide instructional materials to
-teach users how to use the command line interface and other programs
-like R for data visualization. Finally, mothur was born out of a need
-for automating the analysis of large 16S rRNA gene sequence datasets. I
-realized I had a set of skills to fill that need. It has been refreshing
-to see the computational skills of the microbial ecology field grow over
-the past two decades. Looking ahead, we must all take stock of the
-challenges we face in microbial ecology and how our individual skills
-and interests can address these challenges to turn them into
-opportunities.
+visualization tool. Instead, we provide instructional materials to teach
+users how to use the command line interface and other computational
+skills like programming in R for data visualization. Finally, mothur was
+born out of a need for automating the analysis of large 16S rRNA gene
+sequence datasets. It has been refreshing to see the computational
+skills of the microbial ecology field grow over the past two decades.
+Looking ahead, we must all take stock of the challenges we face in
+microbial ecology and how our individual skills and interests can
+address these challenges to turn them into opportunities.
 
@@
 Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra
-Mass, and Marcy Balunas for their unique contributions to the continued
+Maas, and Marcy Balunas for their unique contributions to the continued
 development of mothur.
@@
 
+\leavevmode\hypertarget{ref-Westcott2015}{}%
+47. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
+methods outperform reference-based methods for assigning 16S rRNA gene
+sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
+
+\leavevmode\hypertarget{ref-Westcott2017}{}%
+48. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+improved method for assigning amplicon-based sequence data to
+operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+
 \leavevmode\hypertarget{ref-Quince2009}{}%
-47. \textbf{Quince C}, \textbf{Lanzén A}, \textbf{Curtis TP},
+49. \textbf{Quince C}, \textbf{Lanzén A}, \textbf{Curtis TP},
 \textbf{Davenport RJ}, \textbf{Hall N}, \textbf{Head IM}, \textbf{Read
@@
 \leavevmode\hypertarget{ref-Masella2012}{}%
-48. \textbf{Masella AP}, \textbf{Bartram AK}, \textbf{Truszkowski JM},
+50. \textbf{Masella AP}, \textbf{Bartram AK}, \textbf{Truszkowski JM},
 \textbf{Brown DG}, \textbf{Neufeld JD}. 2012. PANDAseq: Paired-end
@@
 \leavevmode\hypertarget{ref-Callahan2016}{}%
-49. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
+51. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
 \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes SP}. 2016. DADA2:
@@
 \leavevmode\hypertarget{ref-Edgar2011}{}%
-50. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+52. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
 \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
@@
 \leavevmode\hypertarget{ref-Haas2011}{}%
-51. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
+53. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
 \textbf{Feldgarden M}, \textbf{Ward DV}, \textbf{Giannoukos G},
@@
 \leavevmode\hypertarget{ref-Quince2011}{}%
-52. \textbf{Quince C}, \textbf{Lanzen A}, \textbf{Davenport RJ},
+54. \textbf{Quince C}, \textbf{Lanzen A}, \textbf{Davenport RJ},
 \textbf{Turnbaugh PJ}. 2011. Removing noise from pyrosequenced
@@
 \leavevmode\hypertarget{ref-Edgar2010}{}%
-53. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
+55. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
 faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
@@
 \leavevmode\hypertarget{ref-Rognes2016}{}%
-54. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+56. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \leavevmode\hypertarget{ref-Mah2015}{}%
-55. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
+57. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
 \textbf{Vargas C de}, \textbf{Dunthorn M}. 2015. Swarm v2:
@@
 \leavevmode\hypertarget{ref-Schloss2011a}{}%
-56. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
+58. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
 improving methods used in operational taxonomic unit-based approaches
@@
 
-\leavevmode\hypertarget{ref-Westcott2015}{}%
-57. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
-methods outperform reference-based methods for assigning 16S rRNA gene
-sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
-
 \leavevmode\hypertarget{ref-Schloss2016a}{}%
-58. \textbf{Schloss PD}. 2016. Application of a database-independent
+59. \textbf{Schloss PD}. 2016. Application of a database-independent
 approach to assess the quality of operational taxonomic unit picking
@@
 doi:\href{https://doi.org/10.1128/msystems.00027-16}{10.1128/msystems.00027-16}.
-
-\leavevmode\hypertarget{ref-Westcott2017}{}%
-59. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-improved method for assigning amplicon-based sequence data to
-operational taxonomic units. mSphere \textbf{2}:e00073--17.
-doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
 
@@
 \textbf{Figure 1. mothur has consistently been a popular software
-package over the past ten years with a total of more than 8,800
-citations.} Citation data taken from the Web of Science
-(\url{https://www.webofscience.com}) on October 1, 2019. Data from 2019
-are incomplete.
+package over the past ten years with more than 8,800 citations.}
+Citation data taken from the Web of Science
+(\url{https://www.webofscience.com}) on October 1, 2019. The gray line
+segment depicts the projected number of citations for 2019 based on the
+current number of citations for the year and historical trends.
 

### 2019-10-09 code/plot_citations.R

@@ -1,10 +1,19 @@
 library(tidyverse)
 
-read_tsv('data/citations.tsv') %>%
+actual <- read_tsv('data/citations.tsv') %>%
 	gather(-year, key="tool", value="citations") %>%
 	filter(tool == "mothur") %>%
-	ggplot(aes(x=year, y=citations)) +
-		geom_line() +
+	select(-tool) %>%
+	mutate(numbers = "actual")
+
+projected <- actual %>%
+	mutate(numbers = "projected") %>%
+	mutate(citations = if_else(year == 2019, citations * 3/2, citations))
+
+rbind(projected, actual)	%>%
+	ggplot(aes(x=year, y=citations, color=reorder(numbers, desc(numbers)))) +
+		geom_line(show.legend=FALSE) +
+		scale_color_manual(values=c("gray", "black")) +
 		theme_classic() +
 		labs(x="Year", y="Number of Citations") +
 		scale_x_continuous(breaks=seq(2009,2019,2), labels=seq(2009,2019,2))

### 2019-10-10 submission/manuscript.Rmd
Final edits

@@
 By no means is my scientific journey over, but since 2002 I have been on a journey that I did not realize I was on.
-Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology* [@Schloss2009a], it is worth stepping back and using the continued development of mothur as a story that has parallels to many other research stories.
+Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology* [@Schloss2009a] (Figure 1), it is worth stepping back and using the continued development of mothur as a story that has parallels to many other research stories.
 
@@
 This was greater sequencing depth than anyone else had achieved for a single sample.
-As I was preparing the poster, I walked into the office of Jo Handelsman, my postdoctoral research advisor, and laid out the outline for the poster.
+As I was preparing the poster, I walked into the office of Jo Handelsman, my postdoctoral research advisor at the University of Wisconsin, and laid out the outline for the poster.
 She asked if I could add one of those "curvy things", a rarefaction curve, to show where I was in sampling the community.
@@
 Their seminal paper introduced the field to operational taxonomic units (OTUs), rarefaction curves, and richness estimates.
-I do not recall whether my poster had a rarefaction curve on it, but Jo's question primed my career.
+I do not recall whether my poster had a rarefaction curve on it, but Jo's question and that review article primed my career.
 
@@
 Hughes and her colleagues did it manually and with fewer than 300 sequences.
-Although I could possibly do that for my 600 sequences, my goal was to generate 1,000 sequences from the sample and to repeat that sampling effort for other samples.
+Although I could possibly do that for my 600 sequences, my goal was to generate 1,000 sequences from the sample and to repeat that sampling effort with other samples.
 I needed something that could be automated.
-Furthermore, the software that Hughes used, EstimateS [@Colwell2014], required a series of tedious data formatting steps to perform the analyses we were interested in performing.
+Furthermore, the software that Hughes used to build rarefaction curves, EstimateS [@Colwell2014], required a series of tedious data formatting steps to perform the analyses we were interested in performing.
 I had found my first problem.
@@
 With these tools, the field of microbial ecology had a quantitative toolbox for describing and comparing microbial communities.
-Along the way Jo and I would demonstrate the utility of such tools to answer questions like how many OTUs were there in that sample of Alaskan soil and how many sequences were needed to sample each of those OTUs [@Schloss2006c]?
+Along the way Jo and I would demonstrate the utility of such tools for answering questions like how many OTUs were there in that sample of Alaskan soil and how many sequences were needed to sample each of those OTUs [@Schloss2006c]?
 Where were we in the global bacterial census [@Schloss2004b]?
 How does the word usage of *Goodnight, Moon* compare to that of *Portrait of a Lady* and more importantly how is this relevant to microbial ecology [@Schloss2007]?
-Most edifying were the more than 2,400 papers that used DOTUR, SONS, TreeClimber, or $\int$-LIBSHUFF to facilitate their own research questions (Web of Science, 10/1/2019).
+Most edifying were the more than 2,400 papers that used DOTUR, SONS, TreeClimber, or $\int$-LIBSHUFF to facilitate their own research questions (Web of Science, October 1, 2019).
 Had we waited to solve all of the problems that plagued 16S rRNA gene sequencing, we would still be waiting.
@@
 Generating a distance matrix was a prerequisite to using my tools.
-This wasn't trivial, but by cobbling together other tools it was possible.
+This was not trivial, but by cobbling together other tools it was possible.
 We would assemble, trim and correct Sanger sequence reads using Chromas or STADEN [@Bonfield1995], align the sequences using ClustalW [@Thompson1994] or ARB [@Ludwig2004], check for chimeras using partial treeing or Bellerephon [@Hugenholtz2003], and calculate a pairwise distance matrix using DNADIST from the PHYLIP package [@Felsenstein1989].
@@
 Although great focus has been placed on the depth of sampling afforded by 454 and Illumina sequencing, the true benefit of the modern sequencing platforms is the ability to affordably sequence a large number of technical and biological replicates.
+In my opinion, this expansion in the number of replicates more than makes up for the potential limitations incurred by their shorter read lenght.
 In spite of the many technical challenges, we had excuses and heuristics to solve problems that served our needs.
-It is telling that a recent review of "best practices" in generating and analyzing 16S rRNA gene sequences shows that we still have not solved many of these issues and that we have even identified additional problems [@Pollock2018].
+It is telling that a recent review of "best practices" in generating and analyzing 16S rRNA gene sequences shows that we still have not solved many of these issues and that we have, of course, identified additional problems [@Pollock2018].
 
@@
 Because it was an expensive commercial programming environment and never caught on with biologists, I also forgot much of it.
-Even if I forgot the programming syntax of these languages, what these programming experiences taught me was the logic and structure of programming.
-As a postdoc, I would use these experiences to learn the Perl programming language to better understand how LIBSHUFF, a tool for comparing the structure of two communities, worked since it was written in Perl [@Singleton2001].
+Even if I forgot the programming syntax of these languages, what learning these languages taught me was the logic and structure of programming.
+As a postdoc, I would use this background to learn the Perl programming language to better understand how LIBSHUFF (i.e.
+LIBrary SHUFFle), a tool for comparing the structure of two communities, worked since it was written in Perl [@Singleton2001].
 After writing my own version of LIBSHUFF, $\int$-LIBSHUFF, and seeing the speed of the version written in C++ by my collaborator, Bret Larget, I converted my Perl version of DOTUR into C++.
 At the time, the conversion from Perl to C++ seemed like an academic exercise to learn a new language.
-My Perl version only took a minute or so to process the final collection of 1,000 sequences and the C++ version took seconds.
+My Perl version of DOTUR took a minute or so to process the final collection of 1,000 sequences and the C++ version took seconds.
 Was that really such a big difference?
-In hindsight, as we now process datasets with millions of sequences, the decision to learn C++ was critical.
+In hindsight, as we now process datasets with tens of millions of sequences, the decision to learn C++ was critical.
 The ability to pick up computer languages to solve problems, enabled by my prior training in engineering, was a skill that was virtually unheard of in microbiology.
@@
 Without answering his question, I asked where he found a million sequences.
-Little did I know that his email would represent another pivot in the development of these tools.
+Little did I know that his email would represent another pivot in the development of these tools and my career.
 His group would be the first to use 454 sequencing technology to generate 16S rRNA gene sequences [@Sogin2006].
-Although DOTUR could assign those sequences to OTUs, at the scale of millions of sequences, it was slow and required a significant amount of RAM.
+Although DOTUR could assign millions of sequences to OTUs, it was slow and required a significant amount of RAM.
 As I left my postdoc to start my independent career across the state from Sogin's lab at the University of Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS, $\int$-LIBSHUFF, and TreeClimber for the new world of massively parallelized sequencing.
@@
 My "army" ended up being Sarah Westcott who has worked on the mothur project from its inception.
-Today, mothur is over 200,000 lines of code and Sarah has touched or written nearly every line of it.
+Today, mothur is nearly 200,000 lines of code and Sarah has touched or written nearly every line of it.
 Beyond writing and testing mothur's code base, she has become a conduit for many who are trying to learn the tools of microbial ecology.
@@
 This control over the pipeline has made it easier for researchers to customize databases or adapt the pipeline to analyze non-16S rRNA gene sequence data.
-Furthermore, we have provided ample instructional materials to teach users how to implement robust pipelines and the theory behind each step.
+Furthermore, we have provided ample instructional materials to teach users how to implement robust pipelines and the theory behind each step through the project's website (https://www.mothur.org).
 Third, as I mentioned above, there has been a plethora of methods proposed for generating amplicon sequence data, and curating, aligning, checking for chimeras, classifying, and clustering the data.
@@
 Frankly, we have struggled to expand the number of people that make contributions to the mothur code base.
-One challenge we face is that if we looked to third parties to contribute code to mothur, they would need to know C++.
+One challenge we face is that if we looked to others to contribute code to mothur, they would need to know C++.
 Given the paucity of microbiologists that can program in a compiled language like C++, expecting that community to provide contributors who can write code in a syntax that prizes execution efficiency over developer efficiency was not realistic.
@@
 Of course, such choices are always best in hindsight.
-Yet, when we started developing mothur, the ability to interface between scripting languages like R and Python and C++ code was not as well developed as it is today.
+When we started developing mothur, the ability to interface between scripting languages like R and Python and C++ code was not as well developed as it is today.
 For example, the modern version of the Rcpp package was first released in 2009 and its popularity was not immediate [@Eddelbuettel2011].
@@
 A second experiment was the creation of a graphical user interface (GUI) for running mothur.
-Forcing users to interact with mothur through the command line has been a significant hurdle for many.
+Forcing users to interact with mothur through the command line has been a significant hurdle for many (Figure 3).
 Unfortunately, the development effort required to create and maintain a GUI is significant and there is limited funding for such efforts.
@@
 Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks.
-Documentation of commands and parameter values is explicit in users can provide a file with a list of commands and the software returns a logfile with all commands and output recorded.
+A significant downside for web interfaces is the frequent inability to document or return to old versions of software and databases.
+As was experienced with greengenes, if the website is terminated, reproducing old analyses becomes impossible.
+In mothur, documentation of commands and parameter values is explicit in users can provide a file with a list of commands and the software returns a logfile with all commands and output recorded.
 Given the heightened focus on reproducibility in recent years, we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript [@Schloss2018a; @Schloss2018b].
@@
 Both tools provide important information, but they cannot easily replace each other.
-Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is at least an order of magnitude higher than that of amplicon data.
+Although metagenomic data does provide highly resolved taxonomic information, its practical limit of detection is at least an order of magnitude higher than that of amplicon data.
 For example, we analyzed 10,000 16S rRNA sequences from each of about 500 subjects [@Baxter2016].
@@
 We see three key areas that we will continue to help the field to move forward.
-First, just as we adapted through the transitions from Sanger to 454 to MiSeq and PacBio sequencing platforms [@Kozich2013; @Schloss2011b; @Schloss2016b], we must learn whether data from Oxford Nanopore can be an alternative sequencing approach that generates sequence data that is the same quality as existing approaches; thus far, the approach has significant shortcomings for sequencing 16S rRNA gene sequences [@Calus2018].
+First, just as we adapted through the transitions from Sanger to 454 to MiSeq and PacBio sequencing platforms [@Kozich2013; @Schloss2011b; @Schloss2016b], we must learn whether data from Oxford Nanopore and other developing sequencing technologies can be an alternative sequencing approach that generates sequence data that is the same quality as existing approaches; thus far, the approach has significant shortcomings for sequencing 16S rRNA gene sequences [@Calus2018].
 As with the earlier platforms, we must better understand its error profile so that sequencing errors can be corrected.
@@
 Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods [@Westcott2015; @Westcott2017].
-More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality.
+More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing the quality of OTU assignments.
 Finally, there are ongoing controversies that need further exploration.
@@
 The development of mothur would not be possible without the contributions of its many supporters, developers, and users.
-Although not a complete list, I would be remiss if I did not express my gratitude to Sarah Westcott and the other members of my research group, Jo Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra Maas, and Marcy Balunas for their unique contributions to the continued development of mothur.
+Although not a complete list, I would be remiss if I did not express my gratitude to Sarah Westcott and the other members of my research group, Jo Handelsman, Mitch Sogin, Susan Huse, Sarah Highlander, Vincent Young, Lita Proctor, Kendra Maas, and Marcy Balunas for their unique contributions to the continued development of mothur.
 
@@
 **Figure 2.
-The mothur homepage.** From the mothur home page at www.mothur.org, users can download mothur, access a user forum, navigate a wiki with extensive documentation, join the mothur facebook group, and subscribe to the mothur mailing list.
+The mothur homepage.** From the mothur home page at www.mothur.org, users can download mothur, access a user forum, navigate a wiki with extensive documentation, find blog posts that provide additional examples of how to use mothur, join the mothur facebook group, and subscribe to the mothur mailing list.
 

### 2019-10-10 submission/manuscript.tex
Final edits

@@
 the most cited paper published by \emph{Applied and Environmental
-Microbiology} (7), it is worth stepping back and using the continued
-development of mothur as a story that has parallels to many other
-research stories.
+Microbiology} (7) (Figure 1), it is worth stepping back and using the
+continued development of mothur as a story that has parallels to many
+other research stories.
 
@@
 sample. As I was preparing the poster, I walked into the office of Jo
-Handelsman, my postdoctoral research advisor, and laid out the outline
-for the poster. She asked if I could add one of those ``curvy things'',
-a rarefaction curve, to show where I was in sampling the community.
-Rarefaction curves and attempts to estimate the taxonomic richness of
-soil had become popular because of the impactful review by Jennifer
-Hughes and her colleagues (8). Their seminal paper introduced the field
-to operational taxonomic units (OTUs), rarefaction curves, and richness
-estimates. I do not recall whether my poster had a rarefaction curve on
-it, but Jo's question primed my career.
+Handelsman, my postdoctoral research advisor at the University of
+Wisconsin, and laid out the outline for the poster. She asked if I could
+add one of those ``curvy things'', a rarefaction curve, to show where I
+was in sampling the community. Rarefaction curves and attempts to
+estimate the taxonomic richness of soil had become popular because of
+the impactful review by Jennifer Hughes and her colleagues (8). Their
+seminal paper introduced the field to operational taxonomic units
+(OTUs), rarefaction curves, and richness estimates. I do not recall
+whether my poster had a rarefaction curve on it, but Jo's question and
+that review article primed my career.
 
@@
 could possibly do that for my 600 sequences, my goal was to generate
-1,000 sequences from the sample and to repeat that sampling effort for
+1,000 sequences from the sample and to repeat that sampling effort with
 other samples. I needed something that could be automated. Furthermore,
-the software that Hughes used, EstimateS (4), required a series of
-tedious data formatting steps to perform the analyses we were interested
-in performing. I had found my first problem. How would I assign
-sequences to OTUs and use that data to estimate the richness and
-diversity of a sample? The second problem would involve comparing the
-abundance of OTUs found in one sample to another sample. The solution to
-the first problem, DOTUR (Distance-based OTUs and Richness), took us two
-years to develop (9). DOTUR did two things: given a matrix quantifying
-the genetic distance between pairs of sequences, it would cluster those
-sequences into OTUs for any distance threshold to define the OTUs and
-then it would use the frequency of each OTU to calculate a variety of
-alpha diversity metrics. The solutions to the second problem would come
-from our work to develop software including \(\int\)-LIBSHUFF (10), SONS
-(Shared OTUs and Similarity) (11), and TreeClimber (12). Around the same
-time, Catherine Lozupone and Rob Knight were developing their UniFrac
-tools to compare communities with a phylogenetic rather than OTU-based
-approach (13, 14). With these tools, the field of microbial ecology had
-a quantitative toolbox for describing and comparing microbial
-communities. Along the way Jo and I would demonstrate the utility of
-such tools to answer questions like how many OTUs were there in that
-sample of Alaskan soil and how many sequences were needed to sample each
-of those OTUs (15)? Where were we in the global bacterial census (16)?
-How does the word usage of \emph{Goodnight, Moon} compare to that of
-\emph{Portrait of a Lady} and more importantly how is this relevant to
-microbial ecology (17)? Most edifying were the more than 2,400 papers
-that used DOTUR, SONS, TreeClimber, or \(\int\)-LIBSHUFF to facilitate
-their own research questions (Web of Science, 10/1/2019). Had we waited
-to solve all of the problems that plagued 16S rRNA gene sequencing, we
-would still be waiting.
+the software that Hughes used to build rarefaction curves, EstimateS
+(4), required a series of tedious data formatting steps to perform the
+analyses we were interested in performing. I had found my first problem.
+How would I assign sequences to OTUs and use that data to estimate the
+richness and diversity of a sample? The second problem would involve
+comparing the abundance of OTUs found in one sample to another sample.
+The solution to the first problem, DOTUR (Distance-based OTUs and
+Richness), took us two years to develop (9). DOTUR did two things: given
+a matrix quantifying the genetic distance between pairs of sequences, it
+would cluster those sequences into OTUs for any distance threshold to
+define the OTUs and then it would use the frequency of each OTU to
+calculate a variety of alpha diversity metrics. The solutions to the
+second problem would come from our work to develop software including
+\(\int\)-LIBSHUFF (10), SONS (Shared OTUs and Similarity) (11), and
+TreeClimber (12). Around the same time, Catherine Lozupone and Rob
+Knight were developing their UniFrac tools to compare communities with a
+phylogenetic rather than OTU-based approach (13, 14). With these tools,
+the field of microbial ecology had a quantitative toolbox for describing
+and comparing microbial communities. Along the way Jo and I would
+demonstrate the utility of such tools for answering questions like how
+many OTUs were there in that sample of Alaskan soil and how many
+sequences were needed to sample each of those OTUs (15)? Where were we
+in the global bacterial census (16)? How does the word usage of
+\emph{Goodnight, Moon} compare to that of \emph{Portrait of a Lady} and
+more importantly how is this relevant to microbial ecology (17)? Most
+edifying were the more than 2,400 papers that used DOTUR, SONS,
+TreeClimber, or \(\int\)-LIBSHUFF to facilitate their own research
+questions (Web of Science, October 1, 2019). Had we waited to solve all
+of the problems that plagued 16S rRNA gene sequencing, we would still be
+waiting.
 
@@
 that bacteria varied in their \emph{rrn} copy number. Generating a
-distance matrix was a prerequisite to using my tools. This wasn't
+distance matrix was a prerequisite to using my tools. This was not
 trivial, but by cobbling together other tools it was possible. We would
@@
 affordably sequence a large number of technical and biological
-replicates. In spite of the many technical challenges, we had excuses
-and heuristics to solve problems that served our needs. It is telling
-that a recent review of ``best practices'' in generating and analyzing
-16S rRNA gene sequences shows that we still have not solved many of
-these issues and that we have even identified additional problems (35).
+replicates. In my opinion, this expansion in the number of replicates
+more than makes up for the potential limitations incurred by their
+shorter read lenght. In spite of the many technical challenges, we had
+excuses and heuristics to solve problems that served our needs. It is
+telling that a recent review of ``best practices'' in generating and
+analyzing 16S rRNA gene sequences shows that we still have not solved
+many of these issues and that we have, of course, identified additional
+problems (35).
 
@@
 biologists, I also forgot much of it. Even if I forgot the programming
-syntax of these languages, what these programming experiences taught me
-was the logic and structure of programming. As a postdoc, I would use
-these experiences to learn the Perl programming language to better
-understand how LIBSHUFF, a tool for comparing the structure of two
-communities, worked since it was written in Perl (36). After writing my
-own version of LIBSHUFF, \(\int\)-LIBSHUFF, and seeing the speed of the
-version written in C++ by my collaborator, Bret Larget, I converted my
-Perl version of DOTUR into C++. At the time, the conversion from Perl to
-C++ seemed like an academic exercise to learn a new language. My Perl
-version only took a minute or so to process the final collection of
-1,000 sequences and the C++ version took seconds. Was that really such a
-big difference? In hindsight, as we now process datasets with millions
-of sequences, the decision to learn C++ was critical. The ability to
-pick up computer languages to solve problems, enabled by my prior
-training in engineering, was a skill that was virtually unheard of in
-microbiology. Today, researchers without the ability to program are at a
-significant disadvantage (37).
+syntax of these languages, what learning these languages taught me was
+the logic and structure of programming. As a postdoc, I would use this
+background to learn the Perl programming language to better understand
+how LIBSHUFF (i.e.~LIBrary SHUFFle), a tool for comparing the structure
+of two communities, worked since it was written in Perl (36). After
+writing my own version of LIBSHUFF, \(\int\)-LIBSHUFF, and seeing the
+speed of the version written in C++ by my collaborator, Bret Larget, I
+converted my Perl version of DOTUR into C++. At the time, the conversion
+from Perl to C++ seemed like an academic exercise to learn a new
+language. My Perl version of DOTUR took a minute or so to process the
+final collection of 1,000 sequences and the C++ version took seconds.
+Was that really such a big difference? In hindsight, as we now process
+datasets with tens of millions of sequences, the decision to learn C++
+was critical. The ability to pick up computer languages to solve
+problems, enabled by my prior training in engineering, was a skill that
+was virtually unheard of in microbiology. Today, researchers without the
+ability to program are at a significant disadvantage (37).
 
@@
 he found a million sequences. Little did I know that his email would
-represent another pivot in the development of these tools. His group
-would be the first to use 454 sequencing technology to generate 16S rRNA
-gene sequences (38). Although DOTUR could assign those sequences to
-OTUs, at the scale of millions of sequences, it was slow and required a
-significant amount of RAM. As I left my postdoc to start my independent
-career across the state from Sogin's lab at the University of
-Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS,
-\(\int\)-LIBSHUFF, and TreeClimber for the new world of massively
-parallelized sequencing. The new tool would become mothur.
+represent another pivot in the development of these tools and my career.
+His group would be the first to use 454 sequencing technology to
+generate 16S rRNA gene sequences (38). Although DOTUR could assign
+millions of sequences to OTUs, it was slow and required a significant
+amount of RAM. As I left my postdoc to start my independent career
+across the state from Sogin's lab at the University of Massachusetts in
+Amherst, my plan was to rewrite DOTUR, SONS, \(\int\)-LIBSHUFF, and
+TreeClimber for the new world of massively parallelized sequencing. The
+new tool would become mothur.
 
@@
 My ``army'' ended up being Sarah Westcott who has worked on the mothur
-project from its inception. Today, mothur is over 200,000 lines of code
-and Sarah has touched or written nearly every line of it. Beyond writing
-and testing mothur's code base, she has become a conduit for many who
-are trying to learn the tools of microbial ecology. She patiently
-answers questions via email and on the package's discussion forum
-(\url{https://forum.mothur.org}). The community and I are lucky that
-Sarah has stayed with the project for more than a decade. To be honest,
-such dependency on a single person makes the project brittle. In
+project from its inception. Today, mothur is nearly 200,000 lines of
+code and Sarah has touched or written nearly every line of it. Beyond
+writing and testing mothur's code base, she has become a conduit for
+many who are trying to learn the tools of microbial ecology. She
+patiently answers questions via email and on the package's discussion
+forum (\url{https://forum.mothur.org}). The community and I are lucky
+that Sarah has stayed with the project for more than a decade. To be
+honest, such dependency on a single person makes the project brittle. In
 hindsight, it would have been better to have developed mothur with more
@@
 teach users how to implement robust pipelines and the theory behind each
-step. Third, as I mentioned above, there has been a plethora of methods
+step through the project's website (\url{https://www.mothur.org}).
+Third, as I mentioned above, there has been a plethora of methods
 proposed for generating amplicon sequence data, and curating, aligning,
@@
 to expand the number of people that make contributions to the mothur
-code base. One challenge we face is that if we looked to third parties
-to contribute code to mothur, they would need to know C++. Given the
+code base. One challenge we face is that if we looked to others to
+contribute code to mothur, they would need to know C++. Given the
 paucity of microbiologists that can program in a compiled language like
@@
 package that leaned heavily on the R language's C++ interface packages.
-Of course, such choices are always best in hindsight. Yet, when we
-started developing mothur, the ability to interface between scripting
-languages like R and Python and C++ code was not as well developed as it
-is today. For example, the modern version of the Rcpp package was first
-released in 2009 and its popularity was not immediate (66). The
-development of mothur has been a product of the environment that it was
-created in. Although these decisions have largely had positive outcomes,
-there have been tradeoffs that caused us to sacrifice other goals.
+Of course, such choices are always best in hindsight. When we started
+developing mothur, the ability to interface between scripting languages
+like R and Python and C++ code was not as well developed as it is today.
+For example, the modern version of the Rcpp package was first released
+in 2009 and its popularity was not immediate (66). The development of
+mothur has been a product of the environment that it was created in.
+Although these decisions have largely had positive outcomes, there have
+been tradeoffs that caused us to sacrifice other goals.
 
@@
 users to interact with mothur through the command line has been a
-significant hurdle for many. Unfortunately, the development effort
-required to create and maintain a GUI is significant and there is
+significant hurdle for many (Figure 3). Unfortunately, the development
+effort required to create and maintain a GUI is significant and there is
 limited funding for such efforts. The newest version of QIIME (starting
@@
 using a GUI is that there is a risk that reproducibility will suffer if
-users do not have a mechanism to document their mouse clicks.
-Documentation of commands and parameter values is explicit in users can
-provide a file with a list of commands and the software returns a
-logfile with all commands and output recorded. Given the heightened
-focus on reproducibility in recent years, we have extended significant
-effort in developing instructional materials teaching users how to
-organize, document, and execute reproducible pipelines that allow a user
-to go from raw sequence data to a compiled manuscript (65, 70). A final
-example of a failed experiment was a collaboration with programmers
-through Google Summer of Code to develop commands in mothur that ran the
-random forest and SVM machine learning algorithms. Similar to the
-challenges of developing attractive visuals, fitting the algorithms'
-hyperparameters, testing, and deploying the resulting models require a
-significant amount of customization. Furthermore, machine learning is an
-active area of research where methods are still being developed and
-improved. Thankfully, there are numerous R and Python packages that do a
-better job of developing these models (71, 72). In each of our
-``failed'' experiments, the real problems were straying from what mothur
-does well and failing to grasp what we really wanted the innovation to
-do.
+users do not have a mechanism to document their mouse clicks. A
+significant downside for web interfaces is the frequent inability to
+document or return to old versions of software and databases. As was
+experienced with greengenes, if the website is terminated, reproducing
+old analyses becomes impossible. In mothur, documentation of commands
+and parameter values is explicit in users can provide a file with a list
+of commands and the software returns a logfile with all commands and
+output recorded. Given the heightened focus on reproducibility in recent
+years, we have extended significant effort in developing instructional
+materials teaching users how to organize, document, and execute
+reproducible pipelines that allow a user to go from raw sequence data to
+a compiled manuscript (65, 70). A final example of a failed experiment
+was a collaboration with programmers through Google Summer of Code to
+develop commands in mothur that ran the random forest and SVM machine
+learning algorithms. Similar to the challenges of developing attractive
+visuals, fitting the algorithms' hyperparameters, testing, and deploying
+the resulting models require a significant amount of customization.
+Furthermore, machine learning is an active area of research where
+methods are still being developed and improved. Thankfully, there are
+numerous R and Python packages that do a better job of developing these
+models (71, 72). In each of our ``failed'' experiments, the real
+problems were straying from what mothur does well and failing to grasp
+what we really wanted the innovation to do.
 
@@
 information, but they cannot easily replace each other. Although
-metagenomic data does provide highly resolved taxonomic information, the
-limit of detection is at least an order of magnitude higher than that of
-amplicon data. For example, we analyzed 10,000 16S rRNA sequences from
-each of about 500 subjects (73). We can think of this as representing
-about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500
-subjects / 5 16S rRNA gene sequences/genome). Assuming a genome is 4
-Mbp, this would represent a sequencing depth of 4 Tbp. Although such a
-sequencing effort is technically possible, the cost of such an endeavor
-would be considerable and unlikely to be pursued by most researchers. We
-estimate that generating and sequencing the libraries at the University
-of Michigan sequencing core would cost approximately \$150 per library.
-The parallel 16S rRNA gene sequences data would cost approximately \$8
-per library. Furthermore, analyzing such a large dataset with an
-approach that captures the full genetic diversity of the community would
-be financially and technically prohibitive. Going forward, sequencing
-technologies will continue to evolve to capture longer and more high
-quality data and there will always be a need for characterizing the
-taxonomic diversity of microbial communities. With this in mind, there
-will always be a place for tools like mothur that can analyze amplicon
-sequence data.
+metagenomic data does provide highly resolved taxonomic information, its
+practical limit of detection is at least an order of magnitude higher
+than that of amplicon data. For example, we analyzed 10,000 16S rRNA
+sequences from each of about 500 subjects (73). We can think of this as
+representing about 1,000,000 genome equivalents (10,000 16S rRNA
+genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome).
+Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4
+Tbp. Although such a sequencing effort is technically possible, the cost
+of such an endeavor would be considerable and unlikely to be pursued by
+most researchers. We estimate that generating and sequencing the
+libraries at the University of Michigan sequencing core would cost
+approximately \$150 per library. The parallel 16S rRNA gene sequences
+data would cost approximately \$8 per library. Furthermore, analyzing
+such a large dataset with an approach that captures the full genetic
+diversity of the community would be financially and technically
+prohibitive. Going forward, sequencing technologies will continue to
+evolve to capture longer and more high quality data and there will
+always be a need for characterizing the taxonomic diversity of microbial
+communities. With this in mind, there will always be a place for tools
+like mothur that can analyze amplicon sequence data.
 
@@
 MiSeq and PacBio sequencing platforms (60--62), we must learn whether
-data from Oxford Nanopore can be an alternative sequencing approach that
-generates sequence data that is the same quality as existing approaches;
-thus far, the approach has significant shortcomings for sequencing 16S
-rRNA gene sequences (74). As with the earlier platforms, we must better
-understand its error profile so that sequencing errors can be corrected.
-We have learned that moving forward requires that we maintain or improve
-sequence quality. No doubt, datasets and read lengths will improve, but
-these advances should not be made at the cost of data quality. Second,
-with these improvements, we will need to continue to improve our
-algorithms. We have already seen that attempts to use low quality MiSeq
-and HiSeq data caused computational problems leading to the creation of
-open and closed reference clustering methods, which attempted to
-circumvent those problems (75, 76). Unfortunately, comparative analyses
-showed that these methods fail relative to \emph{de novo} clustering
-methods (47, 48). More work is needed to improve reference-based
-clustering methods so that larger datasets can be analyzed without
-sacrificing quality. Finally, there are ongoing controversies that need
+data from Oxford Nanopore and other developing sequencing technologies
+can be an alternative sequencing approach that generates sequence data
+that is the same quality as existing approaches; thus far, the approach
+has significant shortcomings for sequencing 16S rRNA gene sequences
+(74). As with the earlier platforms, we must better understand its error
+profile so that sequencing errors can be corrected. We have learned that
+moving forward requires that we maintain or improve sequence quality. No
+doubt, datasets and read lengths will improve, but these advances should
+not be made at the cost of data quality. Second, with these
+improvements, we will need to continue to improve our algorithms. We
+have already seen that attempts to use low quality MiSeq and HiSeq data
+caused computational problems leading to the creation of open and closed
+reference clustering methods, which attempted to circumvent those
+problems (75, 76). Unfortunately, comparative analyses showed that these
+methods fail relative to \emph{de novo} clustering methods (47, 48).
+More work is needed to improve reference-based clustering methods so
+that larger datasets can be analyzed without sacrificing the quality of
+OTU assignments. Finally, there are ongoing controversies that need
 further exploration. These include the validity and utility of amplicon
@@
 to Sarah Westcott and the other members of my research group, Jo
-Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra
-Maas, and Marcy Balunas for their unique contributions to the continued
-development of mothur.
+Handelsman, Mitch Sogin, Susan Huse, Sarah Highlander, Vincent Young,
+Lita Proctor, Kendra Maas, and Marcy Balunas for their unique
+contributions to the continued development of mothur.
 
@@
 www.mothur.org, users can download mothur, access a user forum, navigate
-a wiki with extensive documentation, join the mothur facebook group, and
-subscribe to the mothur mailing list.
+a wiki with extensive documentation, find blog posts that provide
+additional examples of how to use mothur, join the mothur facebook
+group, and subscribe to the mothur mailing list.
 

### 2019-11-04 submission/manuscript.Rmd
Response to reviewers

@@
 \newpage
-
-Few scientists set out on a nearly two decade-long journey with a specific goal in mind.
-Often we fail to start a scientific journey because it looks too hard.
-Perhaps we get discouraged by all of the things that could go wrong.
-Maybe we stray from the path because we find something that is more interesting.
-Every scientist picks their own path and takes their own forks in the road.
-From the outside, it may appear to be a random walk.
-Nevertheless, these meanderings are common in science.
 
@@
 One can also run the \text{na\"ive} Bayesian classifier through QIIME using the original code developed by the RDP.
-This caused great frustration for many users because there were numerous software dependencies that had to be installed.
+Structuring QIIME as a set of wrappers caused great frustration for many users because there were numerous software dependencies that had to be installed.
+The benefits included the ability for users to access to a wider set of tools and for developers to tie their tool into the popular software package.
 Although the QIIME developers would go on to create virtual machines and use packaging tools to simplify installation, these fixes required sophistication by users who we knew struggled with the basics of navigating a command line.

### 2019-11-04 submission/manuscript.tex
Response to reviewers

@@
 
-Few scientists set out on a nearly two decade-long journey with a
-specific goal in mind. Often we fail to start a scientific journey
-because it looks too hard. Perhaps we get discouraged by all of the
-things that could go wrong. Maybe we stray from the path because we find
-something that is more interesting. Every scientist picks their own path
-and takes their own forks in the road. From the outside, it may appear
-to be a random walk. Nevertheless, these meanderings are common in
-science.
-
 Looking back on scientific journeys can be instructive to others who are
@@
 QIIME. One can also run the \text{na\"ive} Bayesian classifier through
-QIIME using the original code developed by the RDP. This caused great
-frustration for many users because there were numerous software
-dependencies that had to be installed. Although the QIIME developers
-would go on to create virtual machines and use packaging tools to
-simplify installation, these fixes required sophistication by users who
-we knew struggled with the basics of navigating a command line. In
-contrast, when a user runs mothur, they are running mothur. The
-\text{na\"ive} Bayesian classifier code that is in mothur is a rewritten
-version of the original code. When we rewrite someone's software we do
-it with an eye to improving performance, access, and utility for non-16S
-rRNA gene sequence data. For example, while 454 data was popular,
-PyroNoise was an effective tool for denoising flowgram data (49).
-Running the original code required a large Linux computer cluster and
-knowledge of bash and Perl scripting. When we rewrote the code for
-mothur, we made it accessible to people using any operating system with
-a simple command interface (i.e.~trim.flows and shhh.flows). Our
-approach requires significant developer effort, but saves considerable
-user effort. As this benefit is multiplied across thousands of projects,
-the savings to users has been considerable.
+QIIME using the original code developed by the RDP. Structuring QIIME as
+a set of wrappers caused great frustration for many users because there
+were numerous software dependencies that had to be installed. The
+benefits included the ability for users to access to a wider set of
+tools and for developers to tie their tool into the popular software
+package. Although the QIIME developers would go on to create virtual
+machines and use packaging tools to simplify installation, these fixes
+required sophistication by users who we knew struggled with the basics
+of navigating a command line. In contrast, when a user runs mothur, they
+are running mothur. The \text{na\"ive} Bayesian classifier code that is
+in mothur is a rewritten version of the original code. When we rewrite
+someone's software we do it with an eye to improving performance,
+access, and utility for non-16S rRNA gene sequence data. For example,
+while 454 data was popular, PyroNoise was an effective tool for
+denoising flowgram data (49). Running the original code required a large
+Linux computer cluster and knowledge of bash and Perl scripting. When we
+rewrote the code for mothur, we made it accessible to people using any
+operating system with a simple command interface (i.e.~trim.flows and
+shhh.flows). Our approach requires significant developer effort, but
+saves considerable user effort. As this benefit is multiplied across
+thousands of projects, the savings to users has been considerable.
 
